### 2015-08-05 fossilDating.tex
fossil dating

\documentclass[11pt]{article}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[style=authoryear-comp]{biblatex}
\bibliography{fossilDating.bib}

\begin{document}

\title{Phylogenetic estimation of the age of a fossil}
\author{Alexei Drummond, Tanja Stadler}
\date{\today{}}
\maketitle

\section*{Introduction}

% Papers
% http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4033271/ Beyond fossil calibrations: realities of molecular clock practices in evolutionary biology
% http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066245 A Simple Method for Estimating Informative Node Age Priors for the Fossil Calibration of Molecular Divergence Time Analyses
% http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3108556/  Bayesian Phylogenetic Method to Estimate Unknown Sequence Ages

Although there has been much controversy surrounding apparent discrepancies between palaeontological data and molecular phylogenetic inferences it is clear that fossil and molecular data both produce broadly concordant views of evolutionary history\footnote{For example see \url{http://www.actionbioscience.org/evolution/benton.html}}. Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. In particular, although models of molecular evolution have continually been refined over many decades there has been little change in the modelling of fossil data until very recently. One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record actually resulted in concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.

Traditionally the practice has been to use fossils from radiometrically-dated strata as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny. This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock. 
Here we instead focus on what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics.

\subsection*{Phylogenetic tip-dating}

The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and viral taxa. For example, this technique has been successfully employed to estimate the age of human subfossil remains based on ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}. The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite{gray2013evolutionary}.

We extend this work into the realm of morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters. This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. It is distinct from alternative divergence time dating approaches in that is provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. This leads to an estimate of the age of the most recent common ancestor of a group of fossils. We exploit this framework to attempt the phylogenetic estimation of the age of individual fossils based solely on morphological data. The method is applied to a rich and well-characterized fossil data set of penguins and their ancestors.

\section*{Methods}

\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal statistical partners. We extend their work further by investigating the internal consistency of fossil age estimates. The model of time-tree phylogeny employed is the so-called fossilized birth-death process, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a phylogeny. Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. This allows for their ages to be estimated solely based on their morphological characters and the phylogenetic affinities of those characters to other fossils with known ages in the phylogeny. We refer to this estimate of a fossil's age as its {\em phylogenetic age}. In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about a fossil's age is available from phylogenetic data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range, and (iii) how does the amount of morphological data available for a new fossil and the number of existing fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age. These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).

\subsection*{Bayes factors}

The Bayes factor ($BF$) computes the evidence for one model ($M_1$) over another ($M_2$) as the ratio of the marginal probability of the data under each of the two models:

\begin{equation}
BF = \frac{\Pr\{D|M_1\}}{\Pr\{D|M_2\}} = \frac{p(M_1|D)}{p(M_2|D)}\frac{p(M_2)}{p(M_1)}
\end{equation}

We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil. In this case $M_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $M_2$ is the hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya. So for each fossil the prior probability of $M_1$, $p(M_1)$, is easily computed as the width of the geological range divided by 160 and the prior probability of the alternative is simply $p(M_2) = 1 - p(M_1)$. We assume a prior probability of zero that the fossil age is older than 160Mya. Likewise $p(M_1 | D)$ is the posterior probability  that the age falls in the geological range and $p(M_2 | D) = 1 - p(M_1 | D)$. Thus computing the Bayes factor in this instance is easy to do for each fossil once a posterior distribution of the phylogenetic age of the fossil is obtained under a Uniform prior from $(0,160)$ Mya.

\section*{Results}

The phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological ages. Figure \ref{fig:phyloAgeVsGeoAge} plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.9$. The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr.

 Only three of the 36 fossils showed significant inconsistency between the phylogenetic ages and their presumed palaeontological ages. These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.017, 0.0035 and 0.0027 respectively. All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range. It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M. mirandus}: 6.6Myr vs 10Myr (phylogenetic versus palaeontological), {\em P. antarcticus}: 29.8 vs 22, {\em S. muizoni}: 5.0 vs 9.1. The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, suggesting very precise geological knowledge of the ages of these three fossils.

\begin{figure}
\includegraphics[width=5in]{run8_3/8_phyloAgeVsGeoAge.pdf}
\caption{\label{fig:phyloAgeVsGeoAge}
The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates. The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. Blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. The three labelled fossils show significant inconsistency.}
\end{figure}

\begin{figure}
\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{run8_3/fossilDatingDensities_younger.pdf}
\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

\begin{figure}
\includegraphics[width=6in]{run8_3/fossilDatingDensities_older.pdf}
\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr. Red boxes are the superimposed age ranges derived from geological data.}
\end{figure}

% latex table generated in R 3.0.3 by xtable 1.7-4 package
% Tue Jul 28 17:38:04 2015
\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrr}
  \hline
 & Name & post & Error & Rel Err & Age & Lower & Upper \\ 
  \hline
1 & Anthropornis grandis & 0.932 & 5.05 & 0.12 & 38.22 & 33.17 & 43.42 \\ 
  2 & Anthropornis nordenskjoeldi & 0.903 & 5.01 & 0.12 & 38.30 & 32.04 & 44.33 \\ 
  3 & Archaeospheniscus lopdelli & 0.427 & 1.95 & 0.07 & 26.03 & 21.74 & 31.42 \\ 
  4 & Archaeospheniscus lowei & 0.527 & 0.44 & 0.02 & 27.67 & 21.76 & 34.22 \\ 
  5 & Burnside Palaeudyptes & 0.476 & 1.21 & 0.03 & 36.03 & 33.16 & 39.30 \\ 
  6 & Delphinornis arctowskii & 0.477 & 2.01 & 0.05 & 35.87 & 21.34 & 49.43 \\ 
  7 & Delphinornis gracilis & 0.213 & 6.60 & 0.18 & 44.14 & 35.84 & 51.98 \\ 
  8 & Delphinornis larseni & 0.955 & 1.67 & 0.04 & 40.72 & 33.48 & 51.07 \\ 
  9 & Delphinornis wimani & 0.193 & 15.68 & 0.36 & 26.43 & 17.51 & 41.30 \\ 
  10 & Duntroonornis parvus & 0.771 & 0.12 & 0.00 & 26.00 & 16.07 & 32.26 \\ 
  11 & Eretiscus tonnii & 0.476 & 2.78 & 0.15 & 15.89 & 10.36 & 20.30 \\ 
  12 & Icadyptes salasi & 0.198 & 1.77 & 0.05 & 34.70 & 31.48 & 37.38 \\ 
  13 & Inkayacu paracasensis & 0.387 & 0.79 & 0.02 & 35.84 & 32.49 & 38.48 \\ 
  14 & Kairuku grebneffi & 0.614 & 1.39 & 0.05 & 29.36 & 26.11 & 32.74 \\ 
  15 & Kairuku waitaki & 0.609 & 1.49 & 0.05 & 29.21 & 25.11 & 33.98 \\ 
  16 & Madrynornis mirandus & 0.017 & 3.65 & 0.36 & 6.58 & 2.00 & 9.58 \\ 
  17 & Marambiornis exilis & 0.643 & 1.87 & 0.05 & 39.48 & 33.03 & 46.07 \\ 
  18 & Marplesornis novaezealandiae & 0.377 & 5.55 & 0.52 & 16.54 & 12.41 & 19.85 \\ 
  19 & Mesetaornis polaris & 0.716 & 1.12 & 0.03 & 38.71 & 30.61 & 44.81 \\ 
  20 & Pachydyptes ponderosus & 0.364 & 0.46 & 0.01 & 35.88 & 32.35 & 38.79 \\ 
  21 & Palaeeudyptes antarcticus & 0.148 & 4.49 & 0.14 & 37.36 & 30.67 & 42.56 \\ 
  22 & Palaeeudyptes gunnari & 0.873 & 7.10 & 0.16 & 37.20 & 32.44 & 41.79 \\ 
  23 & Palaeeudyptes klekowskii & 0.857 & 6.22 & 0.14 & 37.42 & 32.46 & 41.78 \\ 
  24 & Palaeospheniscus bergi & 0.932 & 2.47 & 0.16 & 17.77 & 14.11 & 21.95 \\ 
  25 & Palaeospheniscus biloculata & 0.784 & 0.95 & 0.05 & 17.58 & 12.20 & 21.61 \\ 
  26 & Palaeospheniscus patagonicus & 0.931 & 0.18 & 0.01 & 18.53 & 15.61 & 20.66 \\ 
  27 & Paraptenodytes antarcticus & 0.003 & 7.63 & 0.35 & 29.72 & 26.34 & 33.67 \\ 
  28 & Perudyptes devriesi & 0.126 & 5.69 & 0.14 & 45.63 & 39.72 & 53.69 \\ 
  29 & Platydyptes marplesi & 0.768 & 2.17 & 0.08 & 24.43 & 20.72 & 27.79 \\ 
  30 & Platydyptes novaezealandiae & 0.515 & 0.83 & 0.03 & 23.92 & 19.11 & 28.55 \\ 
  31 & Pygoscelis grandis & 0.756 & 0.43 & 0.08 & 5.61 & 1.38 & 10.72 \\ 
  32 & Spheniscus megaramphus & 0.535 & 1.60 & 0.20 & 6.54 & 3.28 & 9.11 \\ 
  33 & Spheniscus muizoni & 0.003 & 4.09 & 0.45 & 4.97 & 2.48 & 7.81 \\ 
  34 & Spheniscus urbinai & 0.882 & 0.01 & 0.00 & 7.82 & 4.99 & 9.95 \\ 
  35 & Waimanu manneringi & 0.068 & 3.31 & 0.05 & 57.91 & 53.32 & 62.91 \\ 
  36 & Waimanu tuatahi & 0.408 & 2.27 & 0.04 & 60.67 & 55.09 & 66.38 \\ 
   \hline
\end{tabular}
\caption{\label{fossilTable}Table of 36 fossil penguins. $post$ is the posterior probability that the phylogenetic age is within the paleaontological age range. Error is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\em Age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\% HPD credible interval.}
\end{table}

\printbibliography

\end{document}

### 2015-08-05 fossilDating.tex
methods

@@
 \usepackage{hyperref}
-\usepackage[style=authoryear-comp]{biblatex}
+\usepackage[style=authoryear-comp,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
 \bibliography{fossilDating.bib}
@@
 In particular, although models of molecular evolution have continually been refined over many decades there has been little change in the modelling of fossil data until very recently.
-One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record actually resulted in concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.
+One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.
 
 Traditionally the practice has been to use fossils from radiometrically-dated strata as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny.
-This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock.
+This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
 Here we instead focus on what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics.
@@
 
-The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and viral taxa.
-For example, this technique has been successfully employed to estimate the age of human subfossil remains based on ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}.
+The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa.
+For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}.
 The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite{gray2013evolutionary}.
@@
 This leads to an estimate of the age of the most recent common ancestor of a group of fossils.
-We exploit this framework to attempt the phylogenetic estimation of the age of individual fossils based solely on morphological data.
-The method is applied to a rich and well-characterized fossil data set of penguins and their ancestors.
+We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to other fossils of known age.
+The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors.
 
@@
 
-\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal statistical partners.
-We extend their work further by investigating the internal consistency of fossil age estimates.
-The model of time-tree phylogeny employed is the so-called fossilized birth-death process, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.
+\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal partners under the rule of probability.
+We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age ranges based on radiometrically-dated geological strata.
+The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.
 
@@
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}.
-In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about a fossil's age is available from phylogenetic data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range, and (iii) how does the amount of morphological data available for a new fossil and the number of existing fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
+In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) how does the amount of morphological data available for a new fossil and the number of existing fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
 These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
+
+\subsection*{Estimating the phylogenetic age of penguin fossils}
+
+For each of the 36 penguin fossils in our data set in turn we produced a separate Bayesian phylogenetic analysis in which the fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
+We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya.
+For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two alternative modelling combinations.
+The first was Model 1 in which a strict clock and the simplest model of mophological substitution was assumed \autocite{Lewis2001}.
+We also employed Model 8 from \textcite{gavryushkina2015bayesian} which involved a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
 
@@
 
-The Bayes factor ($BF$) computes the evidence for one model ($M_1$) over another ($M_2$) as the ratio of the marginal probability of the data under each of the two models:
+The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses:
 
 \begin{equation}
-BF = \frac{\Pr\{D|M_1\}}{\Pr\{D|M_2\}} = \frac{p(M_1|D)}{p(M_2|D)}\frac{p(M_2)}{p(M_1)}
+BF = \frac{\Pr\{D|H_1\}}{\Pr\{D|H_2\}} = \frac{p(H_1|D)}{p(H_2|D)}\frac{p(H_2)}{p(H_1)}
 \end{equation}
@@
 We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
-In this case $M_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $M_2$ is the hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya.
-So for each fossil the prior probability of $M_1$, $p(M_1)$, is easily computed as the width of the geological range divided by 160 and the prior probability of the alternative is simply $p(M_2) = 1 - p(M_1)$.
+In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya.
+So for each fossil the prior probability of $H_1$, $p(H_1)$, is easily computed as the width of the geological range divided by 160 and the prior probability of the alternative is simply $p(H_2) = 1 - p(H_1)$.
 We assume a prior probability of zero that the fossil age is older than 160Mya.
-Likewise $p(M_1 | D)$ is the posterior probability  that the age falls in the geological range and $p(M_2 | D) = 1 - p(M_1 | D)$.
+Likewise $p(H_1 | D)$ is the posterior probability that the age falls in the geological range and $p(H_2 | D) = 1 - p(H_1 | D)$.
 Thus computing the Bayes factor in this instance is easy to do for each fossil once a posterior distribution of the phylogenetic age of the fossil is obtained under a Uniform prior from $(0,160)$ Mya.
@@
 
-The phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological ages.
+\subsection*{Strict clock model, no rate variation across morphological characters}
+Although Model 1 is very simple, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological ages.
 Figure \ref{fig:phyloAgeVsGeoAge} plots the geological age and range against the phylogenetic age estimates.
@@
 
- Only three of the 36 fossils showed significant inconsistency between the phylogenetic ages and their presumed palaeontological ages.
+As judged by Bayes factors, none of the fossils showed evidence that the phylogenetic age was inconsistent with the geological age range if we took into account the prior probability of the geological range in the context of a uniform prior from $(0,160)$ Mya.
+However if we were to assign equal prior probability to the competing hypotheses (i.e.
+$p(H_1) = p(H_2) = 0.5$) then only three of the 36 fossils had a Bayes factor $> XX$, suggesting substantial inconsistency between the phylogenetic ages and their presumed palaeontological ages.
 These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.017, 0.0035 and 0.0027 respectively.
@@
 \begin{figure}
-\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
+%\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
 \end{figure}

### 2015-08-06 fossilDating.tex
methods discussion

@@
 \usepackage{hyperref}
-\usepackage[style=authoryear-comp,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
+\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
 \bibliography{fossilDating.bib}
+
+\newcommand{\Mstrict}{{M1}}
+\newcommand{\Mrelaxed}{{M8}}
 
@@
 
-\title{Phylogenetic estimation of the age of a fossil}
+\title{Estimating the phylogenetic age of a fossil}
 \author{Alexei Drummond, Tanja Stadler}
@@
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}.
-In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) how does the amount of morphological data available for a new fossil and the number of existing fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
+In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) how does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
 These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
@@
 
-For each of the 36 penguin fossils in our data set in turn we produced a separate Bayesian phylogenetic analysis in which the fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
-We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya.
+For each of the 36 penguin fossils in our data set in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
+We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya and it also corresponds the upper limit on the origin of the whole phylogeny.
 For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two alternative modelling combinations.
-The first was Model 1 in which a strict clock and the simplest model of mophological substitution was assumed \autocite{Lewis2001}.
-We also employed Model 8 from \textcite{gavryushkina2015bayesian} which involved a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
+The first was Model 1 in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
+We also employed Model 8 from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
 
@@
 
-\subsection*{Strict clock model, no rate variation across morphological characters}
-Although Model 1 is very simple, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological ages.
+\subsection*{\Mstrict{}: Strict clock model, no rate variation among sites}
+Although \Mstrict{} is very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges.
 Figure \ref{fig:phyloAgeVsGeoAge} plots the geological age and range against the phylogenetic age estimates.
@@
 As judged by Bayes factors, none of the fossils showed evidence that the phylogenetic age was inconsistent with the geological age range if we took into account the prior probability of the geological range in the context of a uniform prior from $(0,160)$ Mya.
-However if we were to assign equal prior probability to the competing hypotheses (i.e.
-$p(H_1) = p(H_2) = 0.5$) then only three of the 36 fossils had a Bayes factor $> XX$, suggesting substantial inconsistency between the phylogenetic ages and their presumed palaeontological ages.
+However if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
 These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.017, 0.0035 and 0.0027 respectively.
@@
 It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.6Myr vs 10Myr (phylogenetic versus palaeontological), {\em P.
+mirandus}: 6.6Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
 antarcticus}: 29.8 vs 22, {\em S.
@@
 The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, suggesting very precise geological knowledge of the ages of these three fossils.
+
+\subsection*{\Mrelaxed{}: Relaxed clock, site partitions, rate variation among sites}
+
+\Mrelaxed{} was the best-fitting model according to the analysis of \textcite{gavryushkina2015bayesian}.
+As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils.
+The median error was 2.1 Myr across all 36 fossils.
+All Bayes factors supported the geological age range above the hypothesis of outside the geological age range.
 
@@
 %\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
+\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of its phylogenetic age (i.e.
+the precisions is 1/variance in the marginal posterior distribution of the age).}
 \end{figure}
@@
 
+\section*{Discussion}
+
+In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model.
+The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of morphological change.
+In both cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
+
 \printbibliography

### 2015-08-11 fossilDating.tex

@@
 Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions.
-In particular, although models of molecular evolution have continually been refined over many decades there has been little change in the modelling of fossil data until very recently.
+In particular, although models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu} there has been little change in the modelling of fossil data until very recently.
 One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.
@@
 For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}.
-The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite{gray2013evolutionary}.
+The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite[e.g.][]{gray2013evolutionary}.
 
@@
 \textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal partners under the rule of probability.
-We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age ranges based on radiometrically-dated geological strata.
+We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age range based on radiometrically-dated geological strata.
 The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.
 
-A previous study described how a set of fossils with discrete morphological characters could be used to estimate a phylogeny.
+A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree or chrono-phylogeny.
 Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age.
-This allows for their ages to be estimated solely based on their morphological characters and the phylogenetic affinities of those characters to other fossils with known ages in the phylogeny.
+This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree.
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}.
-In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) how does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
+In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
 These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
@@
 \begin{figure}
-\includegraphics[width=5in]{run8_3/8_phyloAgeVsGeoAge.pdf}
+\includegraphics[width=5in]{run8_5/8_phyloAgeVsGeoAge.pdf}
 \caption{\label{fig:phyloAgeVsGeoAge}
@@
 \begin{figure}
-%\includegraphics[width=5in]{run8_3/8_precisionVsKnownCharacters.pdf}
+\includegraphics[width=5in]{run8_5/8_precisionVsKnownCharacters.pdf}
 \caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of its phylogenetic age (i.e.
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_3/fossilDatingDensities_younger.pdf}
+\includegraphics[width=6in]{run8_5/fossilDatingDensities_younger.pdf}
 \caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr.
@@
 
-% latex table generated in R 3.0.3 by xtable 1.7-4 package
-% Tue Jul 28 17:38:04 2015
-\begin{table}[ht]
-\centering
-\begin{tabular}{rlrrrrrr}
-  \hline
- & Name & post & Error & Rel Err & Age & Lower & Upper \\ 
-  \hline
-1 & Anthropornis grandis & 0.932 & 5.05 & 0.12 & 38.22 & 33.17 & 43.42 \\ 
-  2 & Anthropornis nordenskjoeldi & 0.903 & 5.01 & 0.12 & 38.30 & 32.04 & 44.33 \\ 
-  3 & Archaeospheniscus lopdelli & 0.427 & 1.95 & 0.07 & 26.03 & 21.74 & 31.42 \\ 
-  4 & Archaeospheniscus lowei & 0.527 & 0.44 & 0.02 & 27.67 & 21.76 & 34.22 \\ 
-  5 & Burnside Palaeudyptes & 0.476 & 1.21 & 0.03 & 36.03 & 33.16 & 39.30 \\ 
-  6 & Delphinornis arctowskii & 0.477 & 2.01 & 0.05 & 35.87 & 21.34 & 49.43 \\ 
-  7 & Delphinornis gracilis & 0.213 & 6.60 & 0.18 & 44.14 & 35.84 & 51.98 \\ 
-  8 & Delphinornis larseni & 0.955 & 1.67 & 0.04 & 40.72 & 33.48 & 51.07 \\ 
-  9 & Delphinornis wimani & 0.193 & 15.68 & 0.36 & 26.43 & 17.51 & 41.30 \\ 
-  10 & Duntroonornis parvus & 0.771 & 0.12 & 0.00 & 26.00 & 16.07 & 32.26 \\ 
-  11 & Eretiscus tonnii & 0.476 & 2.78 & 0.15 & 15.89 & 10.36 & 20.30 \\ 
-  12 & Icadyptes salasi & 0.198 & 1.77 & 0.05 & 34.70 & 31.48 & 37.38 \\ 
-  13 & Inkayacu paracasensis & 0.387 & 0.79 & 0.02 & 35.84 & 32.49 & 38.48 \\ 
-  14 & Kairuku grebneffi & 0.614 & 1.39 & 0.05 & 29.36 & 26.11 & 32.74 \\ 
-  15 & Kairuku waitaki & 0.609 & 1.49 & 0.05 & 29.21 & 25.11 & 33.98 \\ 
-  16 & Madrynornis mirandus & 0.017 & 3.65 & 0.36 & 6.58 & 2.00 & 9.58 \\ 
-  17 & Marambiornis exilis & 0.643 & 1.87 & 0.05 & 39.48 & 33.03 & 46.07 \\ 
-  18 & Marplesornis novaezealandiae & 0.377 & 5.55 & 0.52 & 16.54 & 12.41 & 19.85 \\ 
-  19 & Mesetaornis polaris & 0.716 & 1.12 & 0.03 & 38.71 & 30.61 & 44.81 \\ 
-  20 & Pachydyptes ponderosus & 0.364 & 0.46 & 0.01 & 35.88 & 32.35 & 38.79 \\ 
-  21 & Palaeeudyptes antarcticus & 0.148 & 4.49 & 0.14 & 37.36 & 30.67 & 42.56 \\ 
-  22 & Palaeeudyptes gunnari & 0.873 & 7.10 & 0.16 & 37.20 & 32.44 & 41.79 \\ 
-  23 & Palaeeudyptes klekowskii & 0.857 & 6.22 & 0.14 & 37.42 & 32.46 & 41.78 \\ 
-  24 & Palaeospheniscus bergi & 0.932 & 2.47 & 0.16 & 17.77 & 14.11 & 21.95 \\ 
-  25 & Palaeospheniscus biloculata & 0.784 & 0.95 & 0.05 & 17.58 & 12.20 & 21.61 \\ 
-  26 & Palaeospheniscus patagonicus & 0.931 & 0.18 & 0.01 & 18.53 & 15.61 & 20.66 \\ 
-  27 & Paraptenodytes antarcticus & 0.003 & 7.63 & 0.35 & 29.72 & 26.34 & 33.67 \\ 
-  28 & Perudyptes devriesi & 0.126 & 5.69 & 0.14 & 45.63 & 39.72 & 53.69 \\ 
-  29 & Platydyptes marplesi & 0.768 & 2.17 & 0.08 & 24.43 & 20.72 & 27.79 \\ 
-  30 & Platydyptes novaezealandiae & 0.515 & 0.83 & 0.03 & 23.92 & 19.11 & 28.55 \\ 
-  31 & Pygoscelis grandis & 0.756 & 0.43 & 0.08 & 5.61 & 1.38 & 10.72 \\ 
-  32 & Spheniscus megaramphus & 0.535 & 1.60 & 0.20 & 6.54 & 3.28 & 9.11 \\ 
-  33 & Spheniscus muizoni & 0.003 & 4.09 & 0.45 & 4.97 & 2.48 & 7.81 \\ 
-  34 & Spheniscus urbinai & 0.882 & 0.01 & 0.00 & 7.82 & 4.99 & 9.95 \\ 
-  35 & Waimanu manneringi & 0.068 & 3.31 & 0.05 & 57.91 & 53.32 & 62.91 \\ 
-  36 & Waimanu tuatahi & 0.408 & 2.27 & 0.04 & 60.67 & 55.09 & 66.38 \\ 
-   \hline
-\end{tabular}
-\caption{\label{fossilTable}Table of 36 fossil penguins.
-$post$ is the posterior probability that the phylogenetic age is within the paleaontological age range.
-Error is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range.
-{\em Age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\% HPD credible interval.}
-\end{table}
+\input{run8_5/8_summaryTable.tex}
 

### 2015-08-11 fossilDatingFunctions.r

@@ -1,0 +1,231 @@
+
+###############################################################################
+# read the names and age estimates for each of the fossil analyses
+###############################################################################
+loadFossilFiles <- function(dir, regxp, fossilTable, minLogRows) {
+
+  require(coda)
+
+  height <- list()
+  names <- character(0)
+  for (i in 1:nrow(fossilTable)) {
+
+    folder <- dir
+    filename <- list.files(path=folder,pattern=paste(fossilTable$File.name[i],regexp,sep=""))
+
+    if (length(filename) == 1) {
+   
+      name <- as.character(fossilTable$Species.name[i])
+
+      logfile <- read.table(paste(folder,filename[1],sep="/"), sep="\t", header=T)
+
+      if (nrow(logfile) >= minLogRows) {
+
+        logfile$X <- NULL
+        lastState <- logfile$Sample[nrow(logfile)]    
+        thinState <- lastState - logfile$Sample[nrow(logfile)-1]
+   
+        logfile <- mcmc(logfile[,2:ncol(logfile)],start=0, end=lastState, thin=thinState)
+
+        varName <- paste("height.",name,".",sep="")
+        varName <- gsub(" ","_",varName)
+
+        height[[i]] <- logfile[,(varnames(logfile) == varName ), drop=FALSE]
+        names[i] <- name
+        varnames(height[[i]]) <- name        
+      } else {
+        print(paste("Didn't process",filename[1]," because expected at least ", minLogRows, " rows but found ",nrow(logfile), "rows."))
+      }
+    } else {
+      print(paste("Didn't process",fossilTable$File.name[i]," because found ",length(filename), "file matching name."))
+    }
+  }
+  list(heights=height, names=names)
+}
+
+###############################################################################
+# Construct and return main summary data frame of statistics
+###############################################################################
+createSummaryTable <- function(names, height, fossilTable, sequences) {
+
+  require(coda)
+
+  est <- sapply(height,median)
[…]

### 2015-08-11 run8_5/8_fossilDating.R

@@ -1,0 +1,26 @@
+require(coda)
+source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Subversion/compevol/research/fossilDating/run8_5")
+fossilTable8 <- read.table("~/Subversion/compevol/research/fossilDating/run8_2/fossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Subversion/compevol/research/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)
+height8 <- result8$heights
+df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
+
+caption <- "Table of 36 fossil penguins. $post$ is the posterior probability that the phylogenetic age is within the paleaontological age range. Error is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em Age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval."
+label <- "fossilTable"
+
+writeSummaryTable("8_summaryTable.tex",caption, label, df8)
+
+ind <- 1:nrow(df8)
+ind <- ind[df8$post < 0.05]
+orient <- rep(4,length(ind))
+
+createPhyloAgeVsGeoAgePDF("8_phyloAgeVsGeoAge.pdf", df8, fossilTable8, ind, orient)
+createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, ind, orient);
+
+createDensityPlotFigures("fossilDatingDensities", result8$names, height8, fossilTable8, combinedPlots=TRUE)

### 2015-08-16 fossilDating.tex

@@
 In particular, although models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu} there has been little change in the modelling of fossil data until very recently.
-One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data \autocite{tavare2002using}.
+One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
 
@@
 This leads to an estimate of the age of the most recent common ancestor of a group of fossils.
-We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to other fossils of known age.
+We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age.
 The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors.
@@
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}.
-In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age effect the accuracy of the phylogenetic estimate of a fossil's age.
-These three questions are investigated in the context of a morphological data set of of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
+In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
+These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
 
@@
 
-For each of the 36 penguin fossils in our data set in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
-We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya and it also corresponds the upper limit on the origin of the whole phylogeny.
-For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two alternative modelling combinations.
+For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
+We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny.
+For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations.
 The first was Model 1 in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
@@
 \subsection*{\Mstrict{}: Strict clock model, no rate variation among sites}
-Although \Mstrict{} is very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges.
-Figure \ref{fig:phyloAgeVsGeoAge} plots the geological age and range against the phylogenetic age estimates.
+Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges.
+Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates.
 The points in this plot have $R^2 = 0.9$.
@@
 
-\Mrelaxed{} was the best-fitting model according to the analysis of \textcite{gavryushkina2015bayesian}.
-As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils.
+\Mrelaxed{} was the best-fitting model for this data set according to the analysis of \textcite{gavryushkina2015bayesian}.
+As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b).
 The median error was 2.1 Myr across all 36 fossils.
@@
 \begin{figure}
-\includegraphics[width=5in]{run8_5/8_phyloAgeVsGeoAge.pdf}
+\includegraphics{Figure1.pdf}
 \caption{\label{fig:phyloAgeVsGeoAge}
-The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates.
+The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates, under two alternative site and molecular clock models.
 The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits.

### 2015-08-16 fossilDatingFunctions.r

@@ -86,7 +86,11 @@ createSummaryTable <- function(names, height, fossilTable, sequences) {
 
 writeSummaryTable <- function(filename, caption, label, df) {
   require(xtable)
-  xtab <- xtable(df8[,c("post", "bf", "err", "est", "hpd_lower", "hpd_upper","ess")], 
+  
+  dfshort <- df[,c("post", "bf", "est", "hpd_lower", "hpd_upper","err", "ess")]
+  colnames(dfshort) <- c("post", "BF", "phylo age", "lower", "upper", "error", "ESS")
+  
+  xtab <- xtable(dfshort, 
     caption=caption, label=label, digits=c(2,2,1,1,1,1,1,0))
   print(file=filename, xtab)
 }

### 2015-08-16 run8_5/8_fossilDating.R

@@ -11,14 +11,16 @@ result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)
 height8 <- result8$heights
 df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
 
-caption <- "Table of 36 fossil penguins. $post$ is the posterior probability that the phylogenetic age is within the paleaontological age range. Error is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em Age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval."
+caption <- "Table of 36 fossil penguins. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimates effective sample size for the MCMC analysis that produced the phylogenetic age estimate."
 label <- "fossilTable"
+headers <- 
 
 writeSummaryTable("8_summaryTable.tex",caption, label, df8)
 
 ind <- 1:nrow(df8)
 ind <- ind[df8$post < 0.05]
 orient <- rep(4,length(ind))
+orient <- c(4,2,4,4,2)
 
 createPhyloAgeVsGeoAgePDF("8_phyloAgeVsGeoAge.pdf", df8, fossilTable8, ind, orient)
 createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, ind, orient);

### 2015-08-16 fossilDating.tex
added comparison between model 1 and model 8

@@
 
-\title{Estimating the phylogenetic age of a fossil}
+\title{Bayesian phylogenetic estimation of fossil ages reveals remarkable consistency of rocks and clocks}
 \author{Alexei Drummond, Tanja Stadler}
@@
 \maketitle
+
+\abstract{
+Recent advances have allowed for both morphological fossil evidence and molecular sequence data to be integrated into a single combined divergence date estimation analysis under the rule of Bayesian probability.
+In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic location of fossil and extant taxa.
+We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within a rich and well-characterized data set of fossil and extant penguins.
+We find that we can accurately estimate the age of individual penguin fossils based only on phylogenetic evidence.
+In fact in the data set we analyze the {\em phylogenetic age} of a fossil is on average only 2 My from the midpoint of the radiometrically-derived age of the geological strata from which the fossil was excavated.
+The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the palaeontological and molecular data, and provides striking evidence from a real data set that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+}
 
@@
 
+\subsection*{Comparison of M1 and M8 results}
+
+Overall the results from the M1 and M8 models were strikingly concordant.
+Figure \ref{fig:compareM1M8} shows three regressions between the two models: (a) Regression of estimated phylogenetic age of Model 1 against Model 8, (b) Regression of posterior probability of palaeontological range of Model 1 against Model 8, (c) Regression of Bayes factor (BF) for palaeontological range of Model 1 against Model 8.
+
 \begin{figure}
@@
 
+\begin{figure}
+\includegraphics{Figure_comparison.pdf}
+\caption{\label{fig:compareM1M8}
+Comparison of Model 1 against Model 8.
+(a) Regression of estimated phylogenetic age of Model 1 against Model 8, (b) Regression of error in estimated phylogenetic age of Model 1 against Model 8, (c) Regression of posterior probability of palaeontological range of Model 1 against Model 8, (c) Regression of Bayes factor (BF) for palaeontological range of Model 1 against Model 8.}
+\end{figure}
+
 \input{run8_5/8_summaryTable.tex}

### 2015-08-16 compareM1M8.r

@@ -1,0 +1,38 @@
+setwd("~/Subversion/compevol/research/fossilDating")
+createPhyloAgeVsPhyloAgePDF("compareAgeM1M8.pdf",df8,df1,"Model 8", "Model 1",c(),c())
+
+pdf(file="comparePostM1M8.pdf", width=5, height=5, pointsize=10)
+par(mar = c(4,4,1,1) + 0.1)
+plot(df8$post, df1$post, xlab="Posterior probability in Model 8", ylab="Posterior probability in Model 1", pch=16, col="blue", xlim=c(0,1), ylim=c(0,1))
+lm <- lm(df1$post ~ df8$post)
+abline(lm, col="red")
+r2 <- format(summary(lm)$r.squared,digits=3)
+eq <- bquote(bold(R^2 == .(r2)))
+    
+text(0.05,0.925, labels=eq,pos=4)
+dev.off()
+
+logbf1 <- log(df1$bf)
+logbf8 <- log(df8$bf)
+
+pdf(file="compareBFM1M8.pdf", width=5, height=5, pointsize=10)
+par(mar = c(4,4,1,1) + 0.1)
+plot(logbf8, logbf1, xlab="log(BF) in Model 8", ylab="log(BF) in Model 1", pch=16, col="blue")
+lm <- lm(logbf1 ~ logbf8)
+abline(lm, col="red")
+r2 <- format(summary(lm)$r.squared,digits=3)
+eq <- bquote(bold(R^2 == .(r2)))
+    
+text(0.5,5.5, labels=eq,pos=4)
+dev.off()
+
+pdf(file="compareErrorM1M8.pdf", width=5, height=5, pointsize=10)
+par(mar = c(4,4,1,1) + 0.1)
+plot(df8$err, df1$err, xlab="Error in Model 8 age (My)", ylab="Error in Model 1 age (My)", pch=16, col="blue")
+lm <- lm(df1$err ~ df8$err)
+abline(lm, col="red")
+r2 <- format(summary(lm)$r.squared,digits=3)
+eq <- bquote(bold(R^2 == .(r2)))
+    
+text(1,10, labels=eq,pos=4)
+dev.off()

### 2015-08-16 fossilDatingFunctions.r

@@ -132,7 +132,9 @@ createPhyloAgeVsPhyloAgePDF <- function(filename, df1, df2, name1, name2, labell
 
   require(plotrix)
 
-  pdf(file=filename, width=5, height=5, pointsize=10)
+  if (!is.null(filename)) {
+    pdf(file=filename, width=5, height=5, pointsize=10)
+  }
 
   par(mar = c(4,4,1,1) + 0.1)
   plot(c(0,60),c(0,60), type="n", xlab=paste(name1, "age estimate (Myr)"), ylab=paste(name2,"age estimate (Myr)"), ylim=c(0,65), xlim=c(0,65))
@@ -142,7 +144,7 @@ createPhyloAgeVsPhyloAgePDF <- function(filename, df1, df2, name1, name2, labell
   draw.ellipse(df1$est, df2$est, df1$est-df1$hpd_lower, df2$est-df2$hpd_lower, angle = 0, segment = c(180,270), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
   draw.ellipse(df1$est, df2$est, df1$hpd_upper-df1$est, df2$est-df2$hpd_lower, angle = 0, segment = c(270,360), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
-  for (i in 1: nrow(df)) {
+  for (i in 1: nrow(df1)) {
     lines(c(df1$est[i],df1$est[i]), c(df2$hpd_lower[i],df2$hpd_upper[i]), col="black")
     lines(c(df1$hpd_lower[i],df1$hpd_upper[i]), c(df2$est[i],df2$est[i]), col="black")
   }
@@ -151,10 +153,20 @@ createPhyloAgeVsPhyloAgePDF <- function(filename, df1, df2, name1, name2, labell
 
   points(df1$est, df2$est,pch=21, col="white", bg="black")
 
-  text(df1$est[labelled], df2$est[labelled], labels=df1$names[labelled],pos=labelPos, offset=0.4, cex=0.7)
+  if (length(labelled) > 0) {
+    text(df1$est[labelled], df2$est[labelled], labels=df1$names[labelled],pos=labelPos, offset=0.4, cex=0.7)
+  }
 
-  lm <- lm(df1$est~df2$est)
-  dev.off()
[…]

### 2015-08-16 run8_5/8_fossilDating.R

@@ -11,9 +11,8 @@ result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)
 height8 <- result8$heights
 df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
 
-caption <- "Table of 36 fossil penguins. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimates effective sample size for the MCMC analysis that produced the phylogenetic age estimate."
+caption <- "Summary of results for 36 fossil penguins under Model 8. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
 label <- "fossilTable"
-headers <- 
 
 writeSummaryTable("8_summaryTable.tex",caption, label, df8)
 

### 2015-08-16 fossilDating.tex
updated comparison and all ESS greater than 100

@@
 \usepackage{hyperref}
+\usepackage{amsmath}
 \usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
@@
 In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
-These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian} (see Table \ref{fossilTable}).
+These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian}.
 
@@
 For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations.
-The first was Model 1 in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
-We also employed Model 8 from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
+The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
+We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
 
@@
 The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr.
+A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
 
-As judged by Bayes factors, none of the fossils showed evidence that the phylogenetic age was inconsistent with the geological age range if we took into account the prior probability of the geological range in the context of a uniform prior from $(0,160)$ Mya.
+As judged by Bayes factors, none of the fossils exhibited strong evidence (i.e.
+$|\text{log BF}| > 3.0$) that the phylogenetic age was inconsistent with the geological age range assuming a uniform prior from $(0,160)$ Mya.
 However if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
-These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.017, 0.0035 and 0.0027 respectively.
+These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.01, 0.0011 and 0.0007 respectively.
 All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range.
 It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.6Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
-antarcticus}: 29.8 vs 22, {\em S.
-muizoni}: 5.0 vs 9.1.
+mirandus}: 6.3Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
+antarcticus}: 29.9 vs 22, {\em S.
+muizoni}: 5.1 vs 9.1.
 The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, suggesting very precise geological knowledge of the ages of these three fossils.
@@
 The median error was 2.1 Myr across all 36 fossils.
-All Bayes factors supported the geological age range above the hypothesis of outside the geological age range.
+Again, as with \Mstrict{}, none of the fossils exhibited strong evidence (i.e.
+$|\text{log BF}| > 3.0$) that the phylogenetic age was inconsistent with the geological age range assuming a uniform prior from $(0,160)$ Mya.
+However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
+These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.024, 0.016, 0.048, 0.007, 0.037 respectively.
+All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range.
+Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M.
+mirandus}: 6.5Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
+antarcticus}: 28.1 vs 22, {\em P.
+devriesi}: 49.0 vs 40, {\em S.
+muizoni}: 5.1 vs 9.1 and {\em W.
+manneringi}: 56.8 vs 61.05.
+A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}.
+The individual marginal posterior distributions of phylogenetic age under \Mrelaxed{} and corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
 
@@
 Overall the results from the M1 and M8 models were strikingly concordant.
-Figure \ref{fig:compareM1M8} shows three regressions between the two models: (a) Regression of estimated phylogenetic age of Model 1 against Model 8, (b) Regression of posterior probability of palaeontological range of Model 1 against Model 8, (c) Regression of Bayes factor (BF) for palaeontological range of Model 1 against Model 8.
+Figure \ref{fig:compareM1M8} shows three regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
 
@@
 \includegraphics[width=5in]{run8_5/8_precisionVsKnownCharacters.pdf}
-\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of its phylogenetic age (i.e.
+\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age under \Mrelaxed{} (i.e.
 the precisions is 1/variance in the marginal posterior distribution of the age).}
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_5/fossilDatingDensities_younger.pdf}
-\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr.
+\includegraphics[width=6in]{run8_5/fossilDatingHist_younger.pdf}
+\caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
 Red boxes are the superimposed age ranges derived from geological data.}
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_3/fossilDatingDensities_older.pdf}
-\caption{Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr.
+\includegraphics[width=6in]{run8_5/fossilDatingHist_older.pdf}
+\caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
 Red boxes are the superimposed age ranges derived from geological data.}
@@
 \caption{\label{fig:compareM1M8}
-Comparison of Model 1 against Model 8.
-(a) Regression of estimated phylogenetic age of Model 1 against Model 8, (b) Regression of error in estimated phylogenetic age of Model 1 against Model 8, (c) Regression of posterior probability of palaeontological range of Model 1 against Model 8, (c) Regression of Bayes factor (BF) for palaeontological range of Model 1 against Model 8.}
+Comparison of \Mstrict{} against \Mrelaxed{}.
+(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue) to guide eye (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
 \end{figure}
+
+\input{run1_2/1_summaryTable.tex}
 

### 2015-08-16 compareM1M8.r

@@ -34,5 +34,5 @@ abline(lm, col="red")
 r2 <- format(summary(lm)$r.squared,digits=3)
 eq <- bquote(bold(R^2 == .(r2)))
     
-text(1,10, labels=eq,pos=4)
+text(0.5,15.5, labels=eq,pos=4)
 dev.off()

### 2015-08-16 fossilDatingFunctions.r

@@ -232,6 +232,70 @@ createDensityPlotFigures <- function (filePrefix, names, height, fossilTable, co
 }
 
 ####################################################################
+# produce the hist plots and write to pdf files
+####################################################################
+createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combinedPlots) {
+  xmax <- 66
+
+  if (combinedPlots) {
+  
+    pdf(file=paste(filePrefix, "_younger.pdf", sep=""), width=8, height=11)
+    par(mfrow = c(6,3),
+        oma = c(4,4,0,0) + 0.1,
+        mar = c(1,1,2,1) + 0.1,
+        mgp=c(2.5,0.65, 0),lwd=0.5)
+        plot1 <- dev.cur()
+    pdf(file=paste(filePrefix, "_older.pdf", sep=""), width=8, height=11)
+    par(mfrow = c(6,3),
+        oma = c(4,4,0,0) + 0.1,
+        mar = c(1,1,2,1) + 0.1,
+        mgp=c(2.5,0.65, 0),lwd=0.5)
+        plot2 <- dev.cur()
+  }
+
+  for (i in 1:length(height)) {
+
+    if (!combinedPlots) {
+    pdf(file=paste(fossilTable$File.name[i],".pdf",sep=""))
+    }
+ 
+    lower <- fossilTable$Lower[i] 
+    upper <- fossilTable$Upper[i]
+    density <- 1.0/(upper-lower)
+
+    x <- c(lower, lower, upper, upper)
+    y <- c(0.0, density, density, 0.0)
+
+    if (combinedPlots) {
+    if (upper < 32) {
+      dev.set(plot1)
+      xmin <- 0
+      xmax <- 45
+    } else {
+      dev.set(plot2)
+      xmin <- 10
+      xmax <- 70
+    }
+    }
+    hist(height[[i]], xlim=c(xmin,xmax),breaks=c(0:66,160),freq=FALSE, xlab="",main="",col="dark gray", border="white")
+    title(main = list(names[i], cex = 1.0))
+    #rect(lower, 0.0, upper, density, col=rgb(1.0,0.0,0.0,0.5))
+    lines(x,y, col="red",lwd=1)
+
+    if (!combinedPlots) {
+    dev.off()
+    }
+  }            
+   
+  if (combinedPlots) {
+    dev.set(plot1)
+    dev.off()
+    dev.set(plot2)
+    dev.off()
+  }
+}
+
+####################################################################
 # produce the precision versus number of known characters plot
[…]

### 2015-08-16 run1_2/1_fossilDating.r

@@ -1,0 +1,27 @@
+require(coda)
+source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Subversion/compevol/research/fossilDating/run1_2")
+fossilTable1 <- read.table("1_fossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Subversion/compevol/research/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+result1 <- loadFossilFiles(".", regexp, fossilTable1, 10001)
+height1 <- result1$heights
+df1 <- createSummaryTable(result1$names, height1, fossilTable1, seq)
+
+caption <- "Summary of results for 36 fossil penguins under Model 1. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
+label <- "fossilTable1"
+
+writeSummaryTable("1_summaryTable.tex",caption, label, df1)
+
+ind <- 1:nrow(df8)
+ind <- ind[df8$post < 0.05]
+orient <- rep(4,length(ind))
+orient <- c(4,2,4,4,2)
+
+createPhyloAgeVsGeoAgePDF("1_phyloAgeVsGeoAge.pdf", df1, fossilTable1, ind, orient)
+createPrecisionVsKnownCharactersFigure("1_precisionVsKnownCharacters.pdf", df1, ind, orient);
+
+createHistPlotFigures("1_fossilDatingHist", result1$names, height1, fossilTable1, combinedPlots=TRUE)

### 2015-08-16 run8_5/8_fossilDating.R

@@ -12,7 +12,7 @@ height8 <- result8$heights
 df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
 
 caption <- "Summary of results for 36 fossil penguins under Model 8. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
-label <- "fossilTable"
+label <- "fossilTable8"
 
 writeSummaryTable("8_summaryTable.tex",caption, label, df8)
 

### 2015-08-17 run8_5/8_fossilDating.R

@@ -24,4 +24,4 @@ orient <- c(4,2,4,4,2)
 createPhyloAgeVsGeoAgePDF("8_phyloAgeVsGeoAge.pdf", df8, fossilTable8, ind, orient)
 createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, ind, orient);
 
-createDensityPlotFigures("fossilDatingDensities", result8$names, height8, fossilTable8, combinedPlots=TRUE)
+createHistPlotFigures("fossilDatingHist", result8$names, height8, fossilTable8, combinedPlots=TRUE)

### 2015-08-17 fossilDating.tex
hominins

@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model.
-The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of morphological change.
+The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of discrete morphological change.
 In both cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.

### 2015-08-17 fossilDatingFunctions.r

@@ -46,7 +46,7 @@ loadFossilFiles <- function(dir, regxp, fossilTable, minLogRows) {
 ###############################################################################
 # Construct and return main summary data frame of statistics
 ###############################################################################
-createSummaryTable <- function(names, height, fossilTable, sequences) {
+createSummaryTable <- function(names, height, fossilTable, sequences, priorRange=160, hpdProb=0.95) {
 
   require(coda)
 
@@ -55,7 +55,7 @@ createSummaryTable <- function(names, height, fossilTable, sequences) {
   err <- abs(est - geo)
   rel_err <- err/geo
 
-  hpds <- sapply(height, HPDinterval)
+  hpds <- sapply(height, function(x) {HPDinterval(x,prob=hpdProb)})
   hpd_lower <- hpds[1,]
   hpd_upper <- hpds[2,]
   ess <- sapply(height, effectiveSize)
@@ -67,18 +67,31 @@ createSummaryTable <- function(names, height, fossilTable, sequences) {
   
   prior <- numeric(0)
   for (i in 1:length(height)) {
-    prior[i] <- (fossilTable$Upper[i] - fossilTable$Lower[i]) / 160.0
+    prior[i] <- (fossilTable$Upper[i] - fossilTable$Lower[i]) / priorRange
+  }
+  
+  probInHPD <- numeric(0)
+  for (i in 1:length(height)) {
+    if (fossilTable$Upper[i] == fossilTable$Lower[i]) {
+    	if (geo[i] > hpd_lower[i] & geo[i] <= hpd_upper[i]) {
+    		probInHPD[i] = 1.0;
+    	} else {
+    		probInHPD[i] = 0.0;
+    	}
+    } else {
+      probInHPD[i] <- (min(hpd_upper[i], fossilTable$Upper[i]) - max(hpd_lower[i], fossilTable$Lower[i])) / (fossilTable$Upper[i] - fossilTable$Lower[i])
+    }
   }
   
   bf <- (post / prior) / ((1.0-post) / (1.0 - prior))
 
-  df <- data.frame(names, post, prior, bf, err, rel_err, est, hpd_lower, hpd_upper, ess, geo)
+  df <- data.frame(names, post, probInHPD, prior, bf, err, rel_err, est, hpd_lower, hpd_upper, ess, geo)
   df$variance <- sapply(height,var)
[…]

### 2015-08-17 hominin1/hominin.r

@@ -1,0 +1,26 @@
+require(coda)
+source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Subversion/compevol/research/fossilDating/hominin1")
+fossilTable_h <- read.table("homininFossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Subversion/compevol/research/fossilDating/hominin1/hominin_sorted_sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+result_h <- loadFossilFiles(".", regexp, fossilTable_h, 5001)
+height_h <- result_h$heights
+df_h <- createSummaryTable(result_h$names, height_h, fossilTable_h, seq, hpdProb=0.99)
+
+caption <- "Summary of results for 19 fossil hominins under Model 1. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
+label <- "fossilTable_h"
+
+writeSummaryTable("1h_summaryTable.tex",caption, label, df_h)
+
+ind <- 1:nrow(df_h)
+ind <- ind[df_h$probInHPD == 00]
+orient <- rep(4,length(ind))
+
+createPhyloAgeVsGeoAgePDF("1h_phyloAgeVsGeoAge.pdf", df_h, fossilTable_h, ind, orient, min=0, max=8)
+createPrecisionVsKnownCharactersFigure("1h_precisionVsKnownCharacters.pdf", df_h, ind, orient);
+
+createHistPlotFigures("1h_fossilDatingHist", result_h$names, height_h, fossilTable_h, combinedPlots=TRUE, xmax=8, binsPerMY=2)

### 2015-08-17 run1_2/1_fossilDating.r

@@ -16,10 +16,10 @@ label <- "fossilTable1"
 
 writeSummaryTable("1_summaryTable.tex",caption, label, df1)
 
-ind <- 1:nrow(df8)
-ind <- ind[df8$post < 0.05]
+ind <- 1:nrow(df1)
+ind <- ind[df1$post < 0.05]
 orient <- rep(4,length(ind))
-orient <- c(4,2,4,4,2)
+orient <- c(4,2,4)
 
 createPhyloAgeVsGeoAgePDF("1_phyloAgeVsGeoAge.pdf", df1, fossilTable1, ind, orient)
 createPrecisionVsKnownCharactersFigure("1_precisionVsKnownCharacters.pdf", df1, ind, orient);

### 2015-08-17 run1_2/1_fossilDating.r
script tidying

@@ -24,4 +24,8 @@ orient <- c(4,2,4)
 createPhyloAgeVsGeoAgePDF("1_phyloAgeVsGeoAge.pdf", df1, fossilTable1, ind, orient)
 createPrecisionVsKnownCharactersFigure("1_precisionVsKnownCharacters.pdf", df1, ind, orient);
 
-createHistPlotFigures("1_fossilDatingHist", result1$names, height1, fossilTable1, combinedPlots=TRUE)
+# Need to split younger and older ones
+
+createHistPlotFigures("1_fossilDatingHist_younger", result1$names[df1$geo < 32], height1[df1$geo < 32], fossilTable1[df1$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+
+createHistPlotFigures("1_fossilDatingHist_older", result1$names[df1$geo > 32], height1[df1$geo > 32], fossilTable1[df1$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)

### 2015-08-17 run8_5/8_fossilDating.R

@@ -24,4 +24,8 @@ orient <- c(4,2,4,4,2)
 createPhyloAgeVsGeoAgePDF("8_phyloAgeVsGeoAge.pdf", df8, fossilTable8, ind, orient)
 createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, ind, orient);
 
-createHistPlotFigures("fossilDatingHist", result8$names, height8, fossilTable8, combinedPlots=TRUE)
+# Need to split younger and older ones
+
+createHistPlotFigures("8_fossilDatingHist_younger", result8$names[df8$geo < 32], height8[df8$geo < 32], fossilTable8[df8$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+
+createHistPlotFigures("8_fossilDatingHist_older", result8$names[df8$geo > 32], height8[df8$geo > 32], fossilTable8[df8$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)

### 2015-08-18 fossilDating.tex
script tidying
ksepka references

@@
 We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age.
-The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors.
+The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors \autocite{ksepka2010,ksepka2012}.
 
@@
 In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
-These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{gavryushkina2015bayesian}.
+These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}.
 
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_5/fossilDatingHist_younger.pdf}
+\includegraphics[width=6in]{run8_5/8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_5/fossilDatingHist_older.pdf}
+\includegraphics[width=6in]{run8_5/8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.

### 2015-08-25 fossilDating.tex
text changes

@@
+% !BIB TS-program = biber
 \documentclass[11pt]{article}
@@
 
+% 250 words
 \abstract{
-Recent advances have allowed for both morphological fossil evidence and molecular sequence data to be integrated into a single combined divergence date estimation analysis under the rule of Bayesian probability.
-In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic location of fossil and extant taxa.
+Recent advances have allowed for both morphological fossil evidence and molecular sequence data to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability.
+In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic relationships between fossil and extant taxa.
 We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within a rich and well-characterized data set of fossil and extant penguins.
 We find that we can accurately estimate the age of individual penguin fossils based only on phylogenetic evidence. 
-In fact in the data set we analyze the {\em phylogenetic age} of a fossil is on average only 2 My from the midpoint of the radiometrically-derived age of the geological strata from which the fossil was excavated.
+In fact in the data set we analyze the {\em phylogenetic age} of a fossil species is on average only 2 My from the midpoint age of the geological strata from which the fossil species was excavated. 
 The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the palaeontological and molecular data, and provides striking evidence from a real data set that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on morphological data.
 }
@@
 Although there has been much controversy surrounding apparent discrepancies between palaeontological data and molecular phylogenetic inferences it is clear that fossil and molecular data both produce broadly concordant views of evolutionary history\footnote{For example see \url{http://www.actionbioscience.org/evolution/benton.html}}.
-Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions.
-In particular, although models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu} there has been little change in the modelling of fossil data until very recently.
-One of the first examples of endeavouring to apply an evolutionary model to perform inference on the fossil record in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
+Indeed one of the first examples of endeavouring to apply a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
+% AJD: The following two sentences are inflammatory to palaeontologists
+%Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. 
+%Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
+The continual improvement of models and methods for statistical phylogenetic inference of molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang:2006yu}. 
+However, it may be less widely appreciated by molecular evolutionary biologists that phylogenetic reasoning has also been heavily applied to palaeontological questions including to questions of biogeography, character evolution, and the completeness of the fossil record.
+In particular probabilistic model based approaches to biogeography, ancestral state reconstruction, and rates of morphological character change have been common in recent years.
+But the application of phylogenetic approaches to palaeontological questions extends back for decades.
+Nevertheless there has been a call by some palaeontologists for
+%The earlier literature is awash with research papers that include fossils in phylogenies, utilize a phylogenetic bracket to constrain inferences of soft tissue or physiology, or use trees to reconstruct biogeography.
 
-Traditionally the practice has been to use fossils from radiometrically-dated strata as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny.
-This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
+Traditionally the practice has been to use a fossil as a ``node calibration'' by associated its geologically-derived age to a particular divergence in a molecular phylogeny.
+The age can be derived by radiometric aging of strata above and/or below the strata, or by biostratigraphy wish employs This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
 Here we instead focus on what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics.
-
 \subsection*{Phylogenetic tip-dating}
@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model.
+There are a diverse array of potential applications for this methodology. 
+The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found or a complete lack of provenance data (e.g.
+four-legged snake).
+It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 
 The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of discrete morphological change.

### 2015-08-25 fossilDatingFunctions.r

@@ -79,7 +79,7 @@ createSummaryTable <- function(names, height, fossilTable, sequences, priorRange
     		probInHPD[i] = 0.0;
     	}
     } else {
-      probInHPD[i] <- (min(hpd_upper[i], fossilTable$Upper[i]) - max(hpd_lower[i], fossilTable$Lower[i])) / (fossilTable$Upper[i] - fossilTable$Lower[i])
+      probInHPD[i] <- max(0,(min(hpd_upper[i], fossilTable$Upper[i]) - max(hpd_lower[i], fossilTable$Lower[i])) / (fossilTable$Upper[i] - fossilTable$Lower[i]))
     }
   }
   
@@ -271,7 +271,7 @@ createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combi
       pdf(file=paste(fossilTable$File.name[i],".pdf",sep=""))
     }
     
-    hist <- hist(height[[i]], xlim=c(xmin,xmax),breaks=c((0:(xmax*binsPerMY))/binsPerMY,2*xmax),freq=FALSE, xlab="",main="",col="dark gray", border="white")
+    hist <- hist(height[[i]], xlim=c(xmin,xmax),breaks=c((0:(xmax*binsPerMY)/binsPerMY),10*xmax),freq=FALSE, xlab="",main="",col="dark gray", border="white")
     title(main = list(names[i], cex = 1.0))
     #rect(lower, 0.0, upper, density, col=rgb(1.0,0.0,0.0,0.5))
     

### 2015-08-25 fossilDating.tex
four-legged snake footnote
Tanja priors

@@
 
+\subsection*{Bayes factor - update Tanja}
+
+The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses and a model $M$:
+
+\begin{equation}
+BF = \frac{\Pr\{D|H_1,M\}}{\Pr\{D|H_2,M\}} = \frac{p(H_1|D,M)}{p(H_2|D,M)}\frac{p(H_2|M)}{p(H_1|M)}
+\end{equation}
+
+We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
+In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya.
+The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,160)$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1, and the turnover $r$ and sampling probabiltiy $s$ have uniform priors on $[0,1]$)
+
+The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
+It remains to calculate the probabilities $p(H_1|M)$ and $p(H_2|M)$.
+In order to do so, we derive the probability density of sampling a fossil at time $t$ in the past, given the model $M$. 
+For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
+$$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
+with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
+We simplify,
+\begin{eqnarray*}
+p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
+&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
+\end{eqnarray*}
+Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,160)$ on a grid.
+Figuring out now how things change in case we assume uniform prior on $N$, and then evaluate $T=\log(N)/d$.
+
 \section*{Results}
@@
 The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found or a complete lack of provenance data (e.g.
-four-legged snake).
+a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}).
 It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 

### 2015-08-25 canids/canids.r
canids

@@ -1,0 +1,30 @@
+require(coda)
+source("~/Git/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Git/fossilDating/canids")
+fossilTable_c <- read.table("canidFossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Git/fossilDating/canids/canids-sorted-sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+result_c <- loadFossilFiles(".", regexp, fossilTable_c, 5001)
+height_c <- result_c$heights
+df_c <- createSummaryTable(result_c$names, height_c, fossilTable_c, seq, hpdProb=0.99)
+
+caption <- "Summary of results for 123 fossil canids under Model 1. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
+label <- "fossilTable_c"
+
+writeSummaryTable("1c_summaryTable.tex",caption, label, df_c)
+
+ind <- 1:nrow(df_c)
+ind <- ind[df_c$probInHPD == 0]
+orient <- rep(4,length(ind))
+
+createPhyloAgeVsGeoAgePDF("1c_phyloAgeVsGeoAge.pdf", df_c, fossilTable_c, ind, orient, min=0, max=38)
+createPrecisionVsKnownCharactersFigure("1c_precisionVsKnownCharacters.pdf", df_c, ind, orient);
+
+createHistPlotFigures("1c_fossilDatingHist_0to3", result_c$names[df_c$geo < 3], height_c[df_c$geo < 3], fossilTable_c[df_c$geo < 3,], combinedPlots=TRUE, xmin=0, xmax=10)
+
+cond <- (df_c$geo >= 3 & df_c$geo < 10)
+
+createHistPlotFigures("1c_fossilDatingHist_3to10", result_c$names[cond], height_c[cond], fossilTable_c[cond,], combinedPlots=TRUE, xmin=0, xmax=20)

### 2015-08-25 fossilDating.tex
canids
incorporated Tanja's changes

@@
 % http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3108556/  Bayesian Phylogenetic Method to Estimate Unknown Sequence Ages
+% http://www.pnas.org/content/112/16/4897 Canids
 
@@
 
-\subsection*{Bayes factors}
-
-The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses:
-
-\begin{equation}
-BF = \frac{\Pr\{D|H_1\}}{\Pr\{D|H_2\}} = \frac{p(H_1|D)}{p(H_2|D)}\frac{p(H_2)}{p(H_1)}
-\end{equation}
-
-We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
-In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya.
-So for each fossil the prior probability of $H_1$, $p(H_1)$, is easily computed as the width of the geological range divided by 160 and the prior probability of the alternative is simply $p(H_2) = 1 - p(H_1)$.
-We assume a prior probability of zero that the fossil age is older than 160Mya.
-Likewise $p(H_1 | D)$ is the posterior probability that the age falls in the geological range and $p(H_2 | D) = 1 - p(H_1 | D)$.
-Thus computing the Bayes factor in this instance is easy to do for each fossil once a posterior distribution of the phylogenetic age of the fossil is obtained under a Uniform prior from $(0,160)$ Mya.
-
-\subsection*{Bayes factor - update Tanja}
+\subsection*{Bayes factor}
 

### 2015-08-26 fossilDating.tex
working on introduction

@@
 In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic relationships between fossil and extant taxa.
-We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within a rich and well-characterized data set of fossil and extant penguins.
-We find that we can accurately estimate the age of individual penguin fossils based only on phylogenetic evidence. 
-In fact in the data set we analyze the {\em phylogenetic age} of a fossil species is on average only 2 My from the midpoint age of the geological strata from which the fossil species was excavated. 
-The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the palaeontological and molecular data, and provides striking evidence from a real data set that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
+We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
+In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which the fossil species was excavated.
+The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the palaeontological and molecular data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
 We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on morphological data.
@@
 
-Although there has been much controversy surrounding apparent discrepancies between palaeontological data and molecular phylogenetic inferences it is clear that fossil and molecular data both produce broadly concordant views of evolutionary history\footnote{For example see \url{http://www.actionbioscience.org/evolution/benton.html}}.
-Indeed one of the first examples of endeavouring to apply a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
+Contention between palaeontologists and molecular biologists over which data provides the most accurate inferences about evolutionary history has previously fostered an adversarial relationship between the two fields \autocite{Donoghue2007}.
+Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \autocite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \autocite{BentonAyala2003}.
+%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
 % AJD: The following two sentences are inflammatory to palaeontologists
@@
 %Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
-The continual improvement of models and methods for statistical phylogenetic inference of molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang:2006yu}. 
-However, it may be less widely appreciated by molecular evolutionary biologists that phylogenetic reasoning has also been heavily applied to palaeontological questions including to questions of biogeography, character evolution, and the completeness of the fossil record.
-In particular probabilistic model based approaches to biogeography, ancestral state reconstruction, and rates of morphological character change have been common in recent years.
-But the application of phylogenetic approaches to palaeontological questions extends back for decades.
+The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang:2006yu}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}.
+Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \autocite{Donoghue2007}.
+
+%In particular probabilistic model based approaches to biogeography, ancestral state reconstruction, and rates of morphological character change have been common in recent years.
 Nevertheless there has been a call by some palaeontologists for
@@
 
-Traditionally the practice has been to use a fossil as a ``node calibration'' by associated its geologically-derived age to a particular divergence in a molecular phylogeny.
-The age can be derived by radiometric aging of strata above and/or below the strata, or by biostratigraphy wish employs This node calibration then confers age estimates to the remaining divergences on a phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
-Here we instead focus on what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics.
+Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny.
+The age of the fossil is determined either by radiometric aging of strata above and/or below the strata, or more commonly by biostratigraphy.
+The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved.
+Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
+
+It may be less widely appreciated by molecular evolutionary biologists that the statistical phylogenetic revolution in molecular evolution has also been mirrored in the increasing application of statistical phylogenetic reasoning in macroevolutionary and systematic studies of the fossil record \autocite[e.g.][]{Foote1996,huelsenbeck1997maximum, tavare2002using, WagnerMarcot2013}.
+%This includes applying phylogenetic reasoning to questions of biogeography, character evolution, and the completeness of the fossil record.
+Here we extend this tradition of applying phylogenetic reasoning to the fossil record by focusing on the question of what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics and through them, its phylogenetic and temporal relationships with a set of reference fossils.
+
 \subsection*{Phylogenetic tip-dating}
@@
 This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
-It is distinct from alternative divergence time dating approaches in that is provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
+It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
 This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 

### 2015-08-26 fossilDating.tex
adding canid analysis

@@
 \usepackage{amsmath}
+\usepackage{xcolor}
 \usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
@@
 \newcommand{\Mrelaxed}{{M8}}
+
+\newcommand{\todo}[1]{\textcolor{blue}{{TODO: #1}}}
 
@@
 %Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
-The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang:2006yu}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}.
+The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang2014}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}.
 Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \autocite{Donoghue2007}.
@@
 We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
-The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors \autocite{ksepka2010,ksepka2012}.
+The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
 
@@
 In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
-These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}.
+These three questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of canids \autocite{Slater2015}.
 
@@
 
-For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya.
-We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny.
+For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,T_\text{max})$ Mya.
+For the penguins we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny.
 For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations.
@@
 We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
-In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya.
-The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,160)$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1, and the turnover $r$ and sampling probabiltiy $s$ have uniform priors on $[0,1]$)
+In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,T_\text{max})$ Mya.
+The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,T_\text{max})$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1.5, and the turnover $r$ and sampling probability $s$ have uniform priors on $[0,1]$)
 
@@
 \end{eqnarray*}
-Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,160)$ on a grid.
+Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,T_\text{max})$ on a grid.
 Figuring out now how things change in case we assume uniform prior on $N$, and then evaluate $T=\log(N)/d$.
+
+\subsection*{Applying a morphological clock to Canid evolution}
+
+The second data set that we investigated was a morphological data matrix of 132 canid species \autocite[9 extant and 123 fossil;][]{Slater2015}. 
+The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
+For this data set we did not have a geological age range for each fossil, but instead the author of the study provided a point estimate for each fossil species' geological age.
+
+We applied the M1 model to this data set assuming a fixed geological age for each taxon.
+Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
+As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad uniform prior of $(0,T_\text{max})$ Mya.
 
@@
 
-\subsection*{\Mstrict{}: Strict clock model, no rate variation among sites}
+\subsection*{Penguins conform well to a strict morphological clock}
 Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
-Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates.
-The points in this plot have $R^2 = 0.9$.
-The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr.
-A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
+Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.9$. 
+The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
 
@@
 
-\subsection*{\Mrelaxed{}: Relaxed clock, site partitions, rate variation among sites}
+\subsection*{Relaxing the clock, site partitions, rate variation among sites}
 
@@
 
-\subsection*{Comparison of M1 and M8 results}
+\subsection*{Comparison of simple and complex model results}
 
-Overall the results from the M1 and M8 models were strikingly concordant.
-Figure \ref{fig:compareM1M8} shows three regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant.
+Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+
+\subsection*{Canids conform well to a morphological clock}
+
+The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
+The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
+Only 16 out of 123 fossils (13\%) don’t have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
+The median error is $1.66$My, which is even more accurate than the penguin data set.
+There are less characters than the penguins (122 versus $\sim 250$) so the uncertainty in the individual age estimates is greater (median 99\% HPD range = 6.5 My), but overall it seems to work very nicely.
 
@@
 It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 
-The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of discrete morphological change.
-In both cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
+The median error in age estimates for the two data sets investigated were 2 My and 1.6 My respectively, using either a very simple or more complex models of discrete morphological change.
+In all cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
+
+\todo{How does the number of fossils in the reference affect the precision and accuracy of the fossil age estimates ?}
 

### 2015-08-26 fossilDating.tex
adding canid analysis

@@
 A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}.
-The individual marginal posterior distributions of phylogenetic age under \Mrelaxed{} and corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
-
-\subsection*{Comparison of simple and complex model results}
-
-Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant.
-Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
-
-\subsection*{Canids conform well to a morphological clock}
-
-The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
-The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
-Only 16 out of 123 fossils (13\%) don’t have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
-The median error is $1.66$My, which is even more accurate than the penguin data set.
-There are less characters than the penguins (122 versus $\sim 250$) so the uncertainty in the individual age estimates is greater (median 99\% HPD range = 6.5 My), but overall it seems to work very nicely.
+The individual marginal posterior distributions of phylogenetic age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
 
@@
 
+\subsection*{Comparison of simple and complex model results}
+
+Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant.
+Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+
 \begin{figure}
@@
 \input{run8_5/8_summaryTable.tex}
+
+\subsection*{Canids conform well to a morphological clock}
+
+The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
+The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
+Only 16 out of 123 fossils (13\%) don’t have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
+The median error is $1.66$My, which is even more accurate than the penguin data set.
+There are less characters than the penguins (122 versus $\sim 250$) so the uncertainty in the individual age estimates is greater (median 99\% HPD range = 6.5 My), but overall it seems to work very nicely.
+
+\begin{figure}
+\includegraphics{canids/1c_phyloAgeVsGeoAge.pdf}
+\caption{\label{fig:canidMorphVsGeo}
+The Bayesian phylogenetic age estimate (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}.
+The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval.
+Blue line shows the $x=y$.
+If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age.
+The 16 fossils for which this is the case are labelled.}
+\end{figure}
 

### 2015-08-27 fossilDating.tex
adding canid tree
Tanja priors

@@
 
-The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses and a model $M$:
+The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses and a model $M$.
 
@@
 In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,T_\text{max})$ Mya.
-The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,T_\text{max})$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1.5, and the turnover $r$ and sampling probability $s$ have uniform priors on $[0,1]$)
+The model $M$ specifies the tree generation process (called $M_\mathcal{T}$) - giving rise to the number of observed samples and sampling times - together with the morphological evolution (called $M_m$) - giving rise to the morphological characters for the samples.
+The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) together with the morphological characters for each sample ($D_m$).
+Now, we have,
+We have $\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}$
+since $H_1$ and $H_2$ are independent of $M_m$.
 
 The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
-It remains to calculate the probabilities $p(H_1|M)$ and $p(H_2|M)$.
-In order to do so, we derive the probability density of sampling a fossil at time $t$ in the past, given the model $M$. 
+It remains to calculate the probabilities $p(H_1|M_\mathcal{T})$ and $p(H_2|M_\mathcal{T})$.
+In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with the specified priors on its parameters, in our chosen parameteriztion the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
+We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. 
 For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
@@
 \end{eqnarray*}
-Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,T_\text{max})$ on a grid.
-Figuring out now how things change in case we assume uniform prior on $N$, and then evaluate $T=\log(N)/d$.
+
+In order to determine $p(H_1|D,M)$ and $p(H_2|D,M)$, we need to evaluate $p(t|M_\mathcal{T}) = \int_{T,d,r,s} p(t|T,d,r,s) p(T) p(d) p(r) p(s)$ with $p(T) p(d) p(r) p(s)$ being the prior distributions for the parameters.
+This is done by sampling parameters from the prior distributions and then evaluating $p(t|T,d,r,s)$.
+Note that instead of deriving $p(t|T,d,r,s)$ we could have simulated sampling times for the chosen $T,d,r,s$ however some parameter combinations yield huge trees making a full simulation  approach very slow.
+
+First we determined $p(t|M_\mathcal{T})$ for the prior distributions as in \cite{Gavryushkina2014} ($T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1)$).
+This prior specification leads to a distribution of sampling time with almost all mass close to the present (xxx give precise values, Figure xxx, black line).
+Second we  determined $p(t|M_\mathcal{T})$ for the prior distributions ($T: Unif(0,160), d: lognorm(-3.5,0.5), r: Unif(0,1), s: Unif(0,1)$).
+This produces a prior where larger sampling times have a non-negligable weight (Figure xxx, red line).
+
+blabla on only d affecting the prior  shape $p(t|M_\mathcal{T})$ really.
+
+This simple exercise highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is.
+In our example, the first prior seemed reasonable for the parameters specified, however by allowing very large diversification rates, the resulting huge trees dominate the shape of our prior for our hypothesis  $p(t|M_\mathcal{T})$, namely most samples are very very recent.
+Thus this prior  puts a negligible weight on hypothesis $H_1$ for older fossils.
+CHECK IF THIS IS TRUE: Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.
+
+Discuss stepping stone approach - disadvantage in general that we do not see our prior assumption for the tested hypothesis.
+Here not possible at all, as in stepping stone the $D_\mathcal{T}$ is treated as part of the model, not the data.
+This probably originated as  classic tree model is the coalescent, where sampling is not modelled but conditioned upon.
 
@@
 
+\begin{figure}
+\includegraphics[width=15cm]{canids/1_canids-1440516016976-tree5001.pdf}
+\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.}
+\end{figure}
+
 \section*{Discussion}

### 2015-09-02 fossilDating.tex
small changes

@@
 
-\title{Bayesian phylogenetic estimation of fossil ages reveals remarkable consistency of rocks and clocks}
+\title{Bayesian phylogenetic estimation of fossil ages } %reveals remarkable consistency of rocks and clocks}
 \author{Alexei Drummond, Tanja Stadler}
@@
 In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which the fossil species was excavated.
-The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the palaeontological and molecular data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
-We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on morphological data.
+The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
 }
@@
 \begin{figure}
-\includegraphics[width=5in]{run8_5/8_precisionVsKnownCharacters.pdf}
+\includegraphics[width=5in]{penguins_8/8_precisionVsKnownCharacters.pdf}
 \caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age under \Mrelaxed{} (i.e.
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_5/8_fossilDatingHist_younger.pdf}
+\includegraphics[width=6in]{penguins_8/8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=6in]{run8_5/8_fossilDatingHist_older.pdf}
+\includegraphics[width=6in]{penguins_8/8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
@@
 
-\input{run8_5/8_summaryTable.tex}
+\input{penguins_8/8_summaryTable.tex}
 

### 2015-09-02 penguins_8/8_fossilDating.R

@@ -1,7 +1,7 @@
 require(coda)
 source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
 
-setwd("~/Subversion/compevol/research/fossilDating/run8_5")
+setwd("~/Subversion/compevol/research/fossilDating/penguins_8")
 fossilTable8 <- read.table("~/Subversion/compevol/research/fossilDating/run8_2/fossilTable.txt", sep="\t", header=T)
 regexp = "\\-[0-9]+\\.log"
 seq <- read.table("~/Subversion/compevol/research/fossilDating/sorted_sequences.txt", header=T, sep="\t")

### 2015-09-02 fossilDating.tex
updated figures

@@
 
-\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach to the phylogenetic estimation of time-trees that employed both fossil and molecular sequence data as equal partners under the rule of probability.
+\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach \autocite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
 We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age range based on radiometrically-dated geological strata.
@@
 
-First we determined $p(t|M_\mathcal{T})$ for the prior distributions as in \cite{Gavryushkina2014} ($T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1)$).
+First we determined $p(t|M_\mathcal{T})$ for the prior distributions as in \autocite{gavr2014} ($T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1)$).
 This prior specification leads to a distribution of sampling time with almost all mass close to the present (xxx give precise values, Figure xxx, black line).

### 2015-09-02 fossilDatingFunctions.r

@@ -120,12 +120,13 @@ createPhyloAgeVsGeoAgePDF <- function(filename, df, fossilTable, labelled, label
   par(mar = c(4,4,1,1) + 0.1)
   plot(c(min,max*0.925),c(min,max*0.925), type="n", xlab="paleontological age (Myr)", ylab="Bayesian phylogenetic age estimate (Myr)", ylim=c(min,max), xlim=c(min,max))
 
-  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$est-df$hpd_lower, angle = 0, segment = c(180,360), border ="gray", col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
+  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$est-df$hpd_lower, angle = 0, segment = c(180,360), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
-  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$hpd_upper-df$est, angle = 0, segment = c(0,180), border ="gray", col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
+  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$hpd_upper-df$est, angle = 0, segment = c(0,180), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
   for (i in 1: nrow(df)) {
     lines(c(df$geo[i],df$geo[i]), c(df$hpd_lower[i],df$hpd_upper[i]), col="black")
+    lines(c(fossilTable$Lower[i],fossilTable$Upper[i]), c(df$est[i],df$est[i]), col="black")
   }
 
   lines(c(min,max),c(min,max),col="blue")

### 2015-09-02 penguins_8/8_fossilDating.R

@@ -1,10 +1,10 @@
 require(coda)
-source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
+source("~/Git/fossilDating/fossilDatingFunctions.r")
 
-setwd("~/Subversion/compevol/research/fossilDating/penguins_8")
-fossilTable8 <- read.table("~/Subversion/compevol/research/fossilDating/run8_2/fossilTable.txt", sep="\t", header=T)
+setwd("~/Git/fossilDating/penguins_8")
+fossilTable8 <- read.table("fossilTable.txt", sep="\t", header=T)
 regexp = "\\-[0-9]+\\.log"
-seq <- read.table("~/Subversion/compevol/research/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq <- read.table("~/Git/fossilDating/sorted_sequences.txt", header=T, sep="\t")
 seq$Sequence <- as.character(seq$Sequence)
 
 result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)

### 2015-09-02 fossilDating.tex
updating figure 1

@@
 Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
-Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.9$. 
+Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.905$. 
 The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
@@
 \Mrelaxed{} was the best-fitting model for this data set according to the analysis of \textcite{gavryushkina2015bayesian}.
-As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b).
+As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.927$.
 The median error was 2.1 Myr across all 36 fossils.

### 2015-09-02 fossilDatingFunctions.r

@@ -120,9 +120,9 @@ createPhyloAgeVsGeoAgePDF <- function(filename, df, fossilTable, labelled, label
   par(mar = c(4,4,1,1) + 0.1)
   plot(c(min,max*0.925),c(min,max*0.925), type="n", xlab="paleontological age (Myr)", ylab="Bayesian phylogenetic age estimate (Myr)", ylim=c(min,max), xlim=c(min,max))
 
-  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$est-df$hpd_lower, angle = 0, segment = c(180,360), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
+  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$est-df$hpd_lower, angle = 0, segment = c(180,360), border =rgb(0.5,0.5,0.5,0.2), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
-  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$hpd_upper-df$est, angle = 0, segment = c(0,180), border =rgb(0.5,0.5,0.5,0.0), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
+  draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$hpd_upper-df$est, angle = 0, segment = c(0,180), border =rgb(0.5,0.5,0.5,0.2), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
   for (i in 1: nrow(df)) {
     lines(c(df$geo[i],df$geo[i]), c(df$hpd_lower[i],df$hpd_upper[i]), col="black")

### 2015-09-02 fossilDating.tex
changes Tanja

@@
 This prior specification leads to a distribution of sampling time with almost all mass close to the present (xxx give precise values, Figure xxx, black line).
-Second we  determined $p(t|M_\mathcal{T})$ for the prior distributions ($T: Unif(0,160), d: lognorm(-3.5,0.5), r: Unif(0,1), s: Unif(0,1)$).
+
+Second we  determined $p(t|M_\mathcal{T})$ for the following prior distributions $d: lognorm(-3.5,0.5), r: Unif(0,1)$.
+For $s$ we assume an implicit prior: lognorm(-2,1) for $\psi$, and $s=\psi/(\mu+\psi)$.
+For $T$, we also assume an implicit prior.
+We assume a uniform distribution on $[1,100]$ for the number of present day species, $N$.
+In expectation, we have $N=  \frac{e^{d T}}{1-p_0(T)}$ species after time $T$.
+This leads to $T=\log((1-r)N +r)/d$.
 This produces a prior where larger sampling times have a non-negligable weight (Figure xxx, red line).
-
-blabla on only d affecting the prior  shape $p(t|M_\mathcal{T})$ really.
+The implicit prior on $s$ assures that we do not obtain huge sampling rates (which is the case if $s$ and $r$ close to $1$).
+The implicit prior on $T$ assures that we do not obtain trees with arbitrary large number of extant species (which is the case for large $T$ and large $d$).
 

### 2015-09-03 fossilDating.tex
small edits

--- 2015-09-02 fossilDating.tex
@@
 \abstract{
-Recent advances have allowed for both morphological fossil evidence and molecular sequence data to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability.
-In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic relationships between fossil and extant taxa.
+Recent advances have allowed for both morphological fossil evidence and molecular sequences to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability.
+In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa.
 We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
 We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
-In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which the fossil species was excavated.
+In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which it was excavated. 
 The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
-We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
+We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', and for evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
 }
@@
 The age of the fossil is determined either by radiometric aging of strata above and/or below the strata, or more commonly by biostratigraphy.
-The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved.
-Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.
+The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \autocite{Ho2009, Heled2012, HeledDrummond2015}.
+Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006,drummond2010}.
 
@@
 
-We extend this work into the realm of morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters.
+We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters.
 This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
@@
 The model $M$ specifies the tree generation process (called $M_\mathcal{T}$) - giving rise to the number of observed samples and sampling times - together with the morphological evolution (called $M_m$) - giving rise to the morphological characters for the samples.
-The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) together with the morphological characters for each sample ($D_m$).
-Now, we have,
+The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) together with the morphological characters for each sample ($D_m$). Now, we have,
 We have $\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}$
@@
 $$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
-with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
+with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probability of a single lineage producing $i$ surviving lineages at time $t$.
 We simplify,
@@
 Thus this prior  puts a negligible weight on hypothesis $H_1$ for older fossils.
-CHECK IF THIS IS TRUE: Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.
+\todo{CHECK IF THIS IS TRUE:} Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.
 
@@
 The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. 
-Blue line shows the $x=y$.
-If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD.
+Blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
 If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. 
@@
 The Bayesian phylogenetic age estimate (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}.
-The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval.
-Blue line shows the $x=y$.
-If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age.
-The 16 fossils for which this is the case are labelled.}
+The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval. Blue line shows the $x=y$. 
+If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age. The 16 fossils for which this is the case are labelled.}
 \end{figure}

### 2015-09-03 fossilDating.tex
priors by tanja

--- 2015-09-03 fossilDating.tex
@@
 We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
-In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,T_\text{max})$ Mya.
-The model $M$ specifies the tree generation process (called $M_\mathcal{T}$) - giving rise to the number of observed samples and sampling times - together with the morphological evolution (called $M_m$) - giving rise to the morphological characters for the samples.
-The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) together with the morphological characters for each sample ($D_m$). Now, we have,
-We have $\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}$
+In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range.
+%, but within the broader interval of $(0,T_\text{max})$ Mya. 
+The model $M$ consists of two parts, we call them $M_\mathcal{T}$ and $M_m$.
+The model $M_\mathcal{T}$ specifies the tree generation process  giving rise to the number of observed samples and sampling times.
+The model $M_m$ specifies the morphological evolution giving rise to the morphological characters for the samples.
+The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) and  the morphological characters for each sample ($D_m$).
+We have $$\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}=\frac{1-p(H_1|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}$$
 since $H_1$ and $H_2$ are independent of $M_m$.
@@
 It remains to calculate the probabilities $p(H_1|M_\mathcal{T})$ and $p(H_2|M_\mathcal{T})$.
-In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with the specified priors on its parameters, in our chosen parameteriztion the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
-We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. 
+
+In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
+Using the parameterization in \textcite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
+We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. This probability density will allow us to directly determine $p(H_1|M_\mathcal{T})$.
+
 For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
 $$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
-with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probability of a single lineage producing $i$ surviving lineages at time $t$.
+with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
 We simplify,
@@
 
-In order to determine $p(H_1|D,M)$ and $p(H_2|D,M)$, we need to evaluate $p(t|M_\mathcal{T}) = \int_{T,d,r,s} p(t|T,d,r,s) p(T) p(d) p(r) p(s)$ with $p(T) p(d) p(r) p(s)$ being the prior distributions for the parameters.
+Next, we need to evaluate $p(t|M_\mathcal{T}) = \int_{T,d,r,s} p(t|T,d,r,s) p(T) p(d) p(r) p(s)$ with $p(T) p(d) p(r) p(s)$ being the prior distributions for the parameters.
 This is done by sampling parameters from the prior distributions and then evaluating $p(t|T,d,r,s)$.
-Note that instead of deriving $p(t|T,d,r,s)$ we could have simulated sampling times for the chosen $T,d,r,s$ however some parameter combinations yield huge trees making a full simulation  approach very slow.
+(Note that instead of deriving $p(t|T,d,r,s)$ we could have simulated sampling times for the chosen $T,d,r,s$ however full simulations covering the parameter space sufficiently were computational too slow.)
 
-First we determined $p(t|M_\mathcal{T})$ for the prior distributions as in \autocite{gavr2014} ($T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1)$).
-This prior specification leads to a distribution of sampling time with almost all mass close to the present (xxx give precise values, Figure xxx, black line).
-Second we  determined $p(t|M_\mathcal{T})$ for the prior distributions ($T: Unif(0,160), d: lognorm(-3.5,0.5), r: Unif(0,1), s: Unif(0,1)$).
-This produces a prior where larger sampling times have a non-negligable weight (Figure xxx, red line).
+\begin{figure}
+\includegraphics[width=6in]{priors.pdf}
+\caption{\label{hist8_older}Probability density for the sampling times under the fossilized birth-death process.
+The dot-dashed line uses priors on the parameters as in \textcite{gavryushkina2015bayesian}.
+The solid line uses the new prior with implicit assumptions on $T$ and $s$, the dashed line results from only assuming the implicit prior on $T$, the dotted line results from only assuming the  implicit prior on $s$.}
+\label{Fig:Prior}
+\end{figure}
 
-blabla on only d affecting the prior  shape $p(t|M_\mathcal{T})$ really.
+We determined $p(t|M_\mathcal{T})$ for the prior distributions as in \textcite{gavryushkina2015bayesian}, 
+$$T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1).$$ This prior specification leads to a distribution of sampling time with almost all mass close to the present (Figure \ref{Fig:Prior}, dot-dashed line).
+Thus, essentially $p(H_1|M_\mathcal{T})=0$  which leads to a huge Bayes Factor.
+This means we always reject $H_2$, but not because we necessarily agree with the paleontological range, but because our model has no weight for older sampling times.
 
-This simple exercise highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is. 
-In our example, the first prior seemed reasonable for the parameters specified, however by allowing very large diversification rates, the resulting huge trees dominate the shape of our prior for our hypothesis  $p(t|M_\mathcal{T})$, namely most samples are very very recent.
-Thus this prior  puts a negligible weight on hypothesis $H_1$ for older fossils.
+Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain huge trees with arbitrary many species close to the present, thus we have most of the sampling times close to the present.
+(ii) If we draw $r$ and $s$ close to 1, then we obtain a huge per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
+Thus these parameter combinations govern the probability density curve and cause most prior weight to be close to the present.
+
+Thus we assumed new prior distributions: $d: lognorm(-3.5,0.5)$ (note the smaller standard deviation which avoids to put too much weight on very fast growing trees); $r: Unif(0,1)$ as before.
+
+For $s$ we assume an implicit prior: we assume $lognorm(-2,1)$ for $\psi$, and $$s=\psi/(\mu+\psi)$$ (with extinction rate $\mu = r d /(1-r)$). This avoids very high sampling rates.
+
+For $T$, we also assume an implicit prior.
+We assume a uniform distribution on $[1,100]$ for the number of present day species, $N$.
+In expectation, we have $N=  \frac{e^{d T}}{1-p_0(T)}$ species after time $T$.
+This leads to $$T=\log((1-r)N +r)/d.$$
+Overall, this prior produces a sampling time distribution where larger sampling times have a non-negligable weight (Figure \ref{Fig:Prior}, solid line).
+We need further an implicit prior for both $T$ and $s$: only specifying the implicit prior on $T$ yields the dashed line in Figure \ref{Fig:Prior}, while only specifying the implicit prior on $s$ yields the dotted line in Figure \ref{Fig:Prior}.
+We used the new prior for our final analyses and the Bayes Factor calculation.
+
+This example highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is.
+In our example, the  prior from \textcite{gavryushkina2015bayesian} seemed reasonable for the parameters specified, however this prior puts a negligible weight on hypothesis $H_1$ for older fossils.
+
 \todo{CHECK IF THIS IS TRUE:} Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.

### 2015-09-07 fossilDating.tex
new priors on model 1

@@
 
-Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain huge trees with arbitrary many species close to the present, thus we have most of the sampling times close to the present.
-(ii) If we draw $r$ and $s$ close to 1, then we obtain a huge per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
+Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain very large trees with arbitrary many species close to the present, thus we have most of the sampling times close to the present.
+(ii) If we draw $r$ and $s$ close to 1, then we obtain a very large per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
 Thus these parameter combinations govern the probability density curve and cause most prior weight to be close to the present.

### 2015-09-07 1_penguins_eNprior/1_fossilDating.r

@@ -1,0 +1,34 @@
+require(coda)
+source("~/Git/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Git/fossilDating/1_penguins_eNprior")
+fossilTable1 <- read.table("1_fossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Git/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+h1prob <- read.table("H1prob-final.txt", sep=" ", header=T)
+fossilTable1$prior <- h1prob$fossilH1
+
+result1 <- loadFossilFiles(".", regexp, fossilTable1, 10001)
+height1 <- result1$heights
+df1 <- createSummaryTable(result1$names, height1, fossilTable1, seq)
+
+caption <- "Summary of results for 36 fossil penguins under Model 1. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
+label <- "fossilTable1"
+
+writeSummaryTable("1_summaryTable.tex",caption, label, df1)
+
+ind <- 1:nrow(df1)
+ind <- ind[df1$post < 0.05]
+orient <- rep(4,length(ind))
+orient <- c(4,2,4)
+
+createPhyloAgeVsGeoAgePDF("1_phyloAgeVsGeoAge.pdf", df1, fossilTable1, ind, orient)
+createPrecisionVsKnownCharactersFigure("1_precisionVsKnownCharacters.pdf", df1, ind, orient);
+
+# Need to split younger and older ones
+
+createHistPlotFigures("1_fossilDatingHist_younger", result1$names[df1$geo < 32], height1[df1$geo < 32], fossilTable1[df1$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+
+createHistPlotFigures("1_fossilDatingHist_older", result1$names[df1$geo > 32], height1[df1$geo > 32], fossilTable1[df1$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)

### 2015-09-07 fossilDatingFunctions.r

@@ -44,6 +44,40 @@ loadFossilFiles <- function(dir, regxp, fossilTable, minLogRows) {
 }
 
 ###############################################################################
+# read the names and age estimates from a single fossil file
+###############################################################################
+loadFossilFile <- function(filename, fossilTable, minLogRows) {
+
+  require(coda)
+  logfile <- read.table(filename, sep="\t", header=T)
+  if (nrow(logfile) >= minLogRows) {
+    logfile$X <- NULL
+    lastState <- logfile$Sample[nrow(logfile)]    
+    thinState <- lastState - logfile$Sample[nrow(logfile)-1]
+   
+    logfile <- mcmc(logfile[,2:ncol(logfile)],start=0, end=lastState, thin=thinState)
+  
+    height <- list()
+    names <- character(0)
+    for (i in 1:nrow(fossilTable)) {
+  
+      name <- as.character(fossilTable$Species.name[i])
+
+      varName <- paste("height.",name,".",sep="")
+      varName <- gsub(" ","_",varName)
+
+      height[[i]] <- logfile[,(varnames(logfile) == varName ), drop=FALSE]
+      names[i] <- name
+      varnames(height[[i]]) <- name        
+    } 
+  } else {
+    print(paste("Didn't process",filename[1]," because expected at least ", minLogRows, " rows but found ",nrow(logfile), "rows."))
+  }
+  list(heights=height, names=names)
+}
+
+
+###############################################################################
 # Construct and return main summary data frame of statistics
 ###############################################################################
 createSummaryTable <- function(names, height, fossilTable, sequences, priorRange=160, hpdProb=0.95) {
@@ -65,9 +99,15 @@ createSummaryTable <- function(names, height, fossilTable, sequences, priorRange
     post[i] <- length(height[[i]][height[[i]] < fossilTable$Upper[i] & height[[i]] > fossilTable$Lower[i]])/length(height[[i]])    
   }
   
+  
   prior <- numeric(0)
-  for (i in 1:length(height)) {
-    prior[i] <- (fossilTable$Upper[i] - fossilTable$Lower[i]) / priorRange
+  if ("prior" %in% colnames(fossilTable)) {
+  	prior <- fossilTable$prior
+  } else {
+    
+    for (i in 1:length(height)) {
+      prior[i] <- (fossilTable$Upper[i] - fossilTable$Lower[i]) / priorRange
+    }
   }
   
   probInHPD <- numeric(0)

### 2015-09-07 fossilDating.tex
Bayes factor 1st part checked

@@
 For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,T_\text{max})$ Mya.
+\todo{This needs to change to the new prior!} 
 For the penguins we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny.
@@
 
-The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses and a model $M$.
+The Bayes factor ($BF$) computes the evidence for one hypothesis ($H_1$) over another ($H_2$) as the ratio of the marginal probability of the data under each of the two hypotheses and a model $M$,
 
 \begin{equation}
-BF = \frac{\Pr\{D|H_1,M\}}{\Pr\{D|H_2,M\}} = \frac{p(H_1|D,M)}{p(H_2|D,M)}\frac{p(H_2|M)}{p(H_1|M)}
+BF = \frac{\Pr\{D|H_1,M\}}{\Pr\{D|H_2,M\}} = \frac{p(H_1|D,M)}{p(H_2|D,M)}\frac{p(H_2|M)}{p(H_1|M)}.
 \end{equation}
 
-We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil.
-In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range.
+We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the palaeontological age range for each fossil.
+In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the palaeontological range.
 %, but within the broader interval of $(0,T_\text{max})$ Mya. 

### 2015-09-08 fossilDatingFunctions.r
new table

@@ -144,7 +144,7 @@ writeSummaryTable <- function(filename, caption, label, df) {
   colnames(dfshort) <- c("post", "BF", "phylo age", "lower", "upper", "error", "ESS")
   
   xtab <- xtable(dfshort, 
-    caption=caption, label=label, digits=c(2,2,1,1,1,1,1,0))
+    caption=caption, label=label, digits=c(2,2,2,1,1,1,2,0))
   print(file=filename, xtab)
 }
 

### 2015-09-08 run1_2/1_fossilDating.r

@@ -1,10 +1,10 @@
 require(coda)
-source("~/Subversion/compevol/research/fossilDating/fossilDatingFunctions.r")
+source("~/Git/fossilDating/fossilDatingFunctions.r")
 
-setwd("~/Subversion/compevol/research/fossilDating/run1_2")
+setwd("~/Git/fossilDating/run1_2")
 fossilTable1 <- read.table("1_fossilTable.txt", sep="\t", header=T)
 regexp = "\\-[0-9]+\\.log"
-seq <- read.table("~/Subversion/compevol/research/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq <- read.table("~/Git/fossilDating/sorted_sequences.txt", header=T, sep="\t")
 seq$Sequence <- as.character(seq$Sequence)
 
 result1 <- loadFossilFiles(".", regexp, fossilTable1, 10001)

### 2015-09-08 8_penguins_eNprior/8_fossilDating.R
updating relaxed clock figures

@@ -1,0 +1,31 @@
+require(coda)
+source("~/Git/fossilDating/fossilDatingFunctions.r")
+
+setwd("~/Git/fossilDating/8_penguins_eNprior")
+fossilTable8 <- read.table("fossilTable.txt", sep="\t", header=T)
+regexp = "\\-[0-9]+\\.log"
+seq <- read.table("~/Git/fossilDating/sorted_sequences.txt", header=T, sep="\t")
+seq$Sequence <- as.character(seq$Sequence)
+
+result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)
+height8 <- result8$heights
+df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
+
+caption <- "Summary of results for 36 fossil penguins under Model 8. {\\em post} is the posterior probability that the phylogenetic age is within the paleaontological age range. {\\em BF} is the bayes factor in support of the palaeontogical age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's paleaontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
+label <- "fossilTable8"
+
+writeSummaryTable("8_summaryTable.tex",caption, label, df8)
+
+ind <- 1:nrow(df8)
+ind <- ind[df8$post < 0.05]
+orient <- rep(4,length(ind))
+orient <- c(4,2,4,4,2)
+
+createPhyloAgeVsGeoAgePDF("8_phyloAgeVsGeoAge.pdf", df8, fossilTable8, ind, orient)
+createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, ind, orient);
+
+# Need to split younger and older ones
+
+createHistPlotFigures("8_fossilDatingHist_younger", result8$names[df8$geo < 32], height8[df8$geo < 32], fossilTable8[df8$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+
+createHistPlotFigures("8_fossilDatingHist_older", result8$names[df8$geo > 32], height8[df8$geo > 32], fossilTable8[df8$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)

### 2015-09-08 fossilDating.tex
new table
Bayes factor solve merge problems

@@
 \begin{equation}
-BF = \frac{\Pr\{D|H_1,M\}}{\Pr\{D|H_2,M\}} = \frac{p(H_1|D,M)}{p(H_2|D,M)}\frac{p(H_2|M)}{p(H_1|M)}.
+BF = \frac{p(D|H_1,M)}{p(D|H_2,M)} = \frac{p(H_1|D,M)}{p(H_2|D,M)}\frac{p(H_2|M)}{p(H_1|M)}.
 \end{equation}
@@
 %, but within the broader interval of $(0,T_\text{max})$ Mya. 
-The model $M$ consists of two parts, we call them $M_\mathcal{T}$ and $M_m$.
+The model $M$ consists of two parts, $M=(M_\mathcal{T},M_m)$.
 The model $M_\mathcal{T}$ specifies the tree generation process  giving rise to the number of observed samples and sampling times.
-The model $M_m$ specifies the morphological evolution giving rise to the morphological characters for the samples.
-The data $D$ is the number of samples together with the sampling times ($D_\mathcal{T}$) and  the morphological characters for each sample ($D_m$).
-We have $$\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}=\frac{1-p(H_1|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}$$
-since $H_1$ and $H_2$ are independent of $M_m$.
+The model $M_m$ specifies the morphological evolution along the tree giving rise to the morphological characters for the samples.
+The data $D=(D_\mathcal{T},D_m)$ is the number of samples together with the sampling times ($D_\mathcal{T}$) and  the morphological characters for each sample ($D_m$).
 
-The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
-It remains to calculate the probabilities $p(H_1|M_\mathcal{T})$ and $p(H_2|M_\mathcal{T})$.
+For caclulating the Bayes factor, the probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
+It remains to calculate the probabilities $p(H_1|M)$ and $p(H_2|M)$.
+Since $H_1$ and $H_2$ are independent of $M_m$, we have $$\frac{p(H_2|M)}{p(H_1|M)}=\frac{p(H_2|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}=\frac{1-p(H_1|M_\mathcal{T})}{p(H_1|M_\mathcal{T})}.$$
 
+We now derive  $p(H_1|M_\mathcal{T})$.
 In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
@@
 with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
-We simplify,
+In words, the equation above calculates the following: the left term conditions on survival of the process ($1-p_0(T;d,r)$).
+Then we calculate the probability to have $k$ lineages at time $t$ before the present $( p_k(T-t;d,r))$, multiply by the sampling rate $k \psi$, and weight by the probability that at least one lineage of the $k$ lineages survives to the present $ (1-p_0(t;d,r)^k)$.
+This expression is then summed over $k=1,\ldots, \infty$.
+
+We simplify, using the equations for $p_i(t;d,r)$ given in \cite{Kendall1948},
 \begin{eqnarray*}
 p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
-&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
+&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right],
 \end{eqnarray*}
+with $p_0(t|d,r) = \frac{1-e^{-d*t}}{1/r - e^{-d*t}}$.
 
@@
 This is done by sampling parameters from the prior distributions and then evaluating $p(t|T,d,r,s)$.
-(Note that instead of deriving $p(t|T,d,r,s)$ we could have simulated sampling times for the chosen $T,d,r,s$ however full simulations covering the parameter space sufficiently were computational too slow.)
 
@@
 
-\input{run1_2/1_summaryTable.tex}
+\input{1_penguins_eNprior/1_summaryTable.tex}
 

### 2015-09-08 fossilDating.tex
update comparison figure

--- 2015-09-07 fossilDating.tex
@@
 \begin{figure}
-\includegraphics[width=5in]{penguins_8/8_precisionVsKnownCharacters.pdf}
-\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age under \Mrelaxed{} (i.e.
-the precisions is 1/variance in the marginal posterior distribution of the age).}
+\includegraphics[width=5in]{Figure2.pdf}
+\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
+the precision is 1/variance in the marginal posterior distribution of the age).}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=6in]{penguins_8/8_fossilDatingHist_younger.pdf}
+\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=6in]{penguins_8/8_fossilDatingHist_older.pdf}
+\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
@@
 
-\input{run1_2/1_summaryTable.tex}
+\input{1_penguins_eNprior/1_summaryTable.tex}
 
-\input{penguins_8/8_summaryTable.tex}
+\input{8_penguins_eNprior/8_summaryTable.tex}
 

### 2015-09-08 compareM1M8.r

@@ -1,4 +1,4 @@
-setwd("~/Subversion/compevol/research/fossilDating")
+setwd("~/Git/fossilDating")
 createPhyloAgeVsPhyloAgePDF("compareAgeM1M8.pdf",df8,df1,"Model 8", "Model 1",c(),c())
 
 pdf(file="comparePostM1M8.pdf", width=5, height=5, pointsize=10)
@@ -23,7 +23,7 @@ abline(lm, col="red")
 r2 <- format(summary(lm)$r.squared,digits=3)
 eq <- bquote(bold(R^2 == .(r2)))
     
-text(0.5,5.5, labels=eq,pos=4)
+text(0.5,4.1, labels=eq,pos=4)
 dev.off()
 
 pdf(file="compareErrorM1M8.pdf", width=5, height=5, pointsize=10)

### 2015-09-08 fossilDating.tex

--- 2015-09-08 fossilDating.tex
@@
 
-\input{penguins_8/8_summaryTable.tex}
+\input{8_penguins_eNprior/8_summaryTable.tex}
 

### 2015-09-08 fossilDating.tex
Bayes factor clean up

@@
 
-We now derive  $p(H_1|M_\mathcal{T})$.
+One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $\mathcal{T}$ and record the fraction of sampling times  within a given  paleontological age range.
+However, such a simulation approach turned out to be very time-consuming, and the procedure below provided a much faster evaluation of $p(H_1|M_\mathcal{T})$.
+
+We  derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$.
 In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
@@
 with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
-In words, the equation above calculates the following: the left term conditions on survival of the process ($1-p_0(T;d,r)$).
+The equation above calculates the required probability: The left term conditions on survival of the process ($1-p_0(T;d,r)$).
 Then we calculate the probability to have $k$ lineages at time $t$ before the present $( p_k(T-t;d,r))$, multiply by the sampling rate $k \psi$, and weight by the probability that at least one lineage of the $k$ lineages survives to the present $ (1-p_0(t;d,r)^k)$.
@@
 
-We simplify, using the equations for $p_i(t;d,r)$ given in \cite{Kendall1948},
+We simplify, using the equations for $p_i(t;d,r)$ given in \cite{Kendall1948}, to obtain,
 \begin{eqnarray*}
@@
 
-Next, we need to evaluate $p(t|M_\mathcal{T}) = \int_{T,d,r,s} p(t|T,d,r,s) p(T) p(d) p(r) p(s)$ with $p(T) p(d) p(r) p(s)$ being the prior distributions for the parameters.
-This is done by sampling parameters from the prior distributions and then evaluating $p(t|T,d,r,s)$.
+Next, we need to evaluate $p(t|M_\mathcal{T}) = \int_{T,d,r,s} p(t|T,d,r,s) p(T) p(d) p(r) p(s)$ with $p(T), p(d), p(r)$ and $p(s)$ being the prior distributions for the parameters.
+This is done by sampling parameters from the prior distributions, and then evaluating $p(t|T,d,r,s)$.
 
@@
 The dot-dashed line uses priors on the parameters as in \textcite{gavryushkina2015bayesian}.
-The solid line uses the new prior with implicit assumptions on $T$ and $s$, the dashed line results from only assuming the implicit prior on $T$, the dotted line results from only assuming the  implicit prior on $s$.}
+The solid line uses the new prior with implicit assumptions on $T$ and $s$, the dashed line results from only assuming the implicit prior on $T$, the dotted line results from only assuming the  implicit prior on $s$.
+(Since the dashed, dotted and dot-dashed lines are governed by rare parameter combinations leading to huge trees and huge sample sizes, these lines are very sensitive to drawing another rare event, and thus need to be taken with some caution).}
 \label{Fig:Prior}
@@
 Thus, essentially $p(H_1|M_\mathcal{T})=0$  which leads to a huge Bayes Factor.
-This means we always reject $H_2$, but not because we necessarily agree with the paleontological range, but because our model has no weight for older sampling times.
+This means we always reject $H_2$, but not because we necessarily agree with the paleontological age range, but because our model has no prior weight for the paleontological age range.
 
@@
 
-Thus we assumed new prior distributions: $d: lognorm(-3.5,0.5)$ (note the smaller standard deviation which avoids to put too much weight on very fast growing trees); $r: Unif(0,1)$ as before.
+We thus assumed new prior distributions.
+The net diversification rate $d: lognorm(-3.5,0.5)$ was chosen with a  smaller standard deviation which avoids  too much weight on very fast growing trees.
+The turnover $r: Unif(0,1)$ was set as before.
 
@@
 This leads to $$T=\log((1-r)N +r)/d.$$
-Overall, this prior produces a sampling time distribution where larger sampling times have a non-negligable weight (Figure \ref{Fig:Prior}, solid line).
-We need further an implicit prior for both $T$ and $s$: only specifying the implicit prior on $T$ yields the dashed line in Figure \ref{Fig:Prior}, while only specifying the implicit prior on $s$ yields the dotted line in Figure \ref{Fig:Prior}.
-We used the new prior for our final analyses and the Bayes Factor calculation.
+Overall, this prior produces a sampling time distribution where old sampling times have a non-negligable weight (Figure \ref{Fig:Prior}, solid line). 
+The choice of an implicit prior for both $T$ and $s$ was important: only specifying the implicit prior on $T$ yields the dashed line in Figure \ref{Fig:Prior}, while only specifying the implicit prior on $s$ yields the dotted line in Figure \ref{Fig:Prior}.
+We used this new prior for our  analyses and the Bayes Factor calculation.
 
-This example highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is.
+Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters compared to using the prior in  \textcite{gavryushkina2015bayesian}.
+This investigation of the prior distribution on trees and sampling times highlights that whenever using Bayes factors to test a hypothesis, we have to first investigate what our prior on the hypothesis is.
 In our example, the  prior from \textcite{gavryushkina2015bayesian} seemed reasonable for the parameters specified, however this prior puts a negligible weight on hypothesis $H_1$ for older fossils.
 
-\todo{CHECK IF THIS IS TRUE:} Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.
-
-Discuss stepping stone approach - disadvantage in general that we do not see our prior assumption for the tested hypothesis.
-Here not possible at all, as in stepping stone the $D_\mathcal{T}$ is treated as part of the model, not the data.
-This probably originated as  classic tree model is the coalescent, where sampling is not modelled but conditioned upon.
+Further, we want to note that a classic  stepping stone approach to calculate Bayes factors would have not been possible in our case:  In stepping stone approaches the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
+However, using a birth-death model, the sampling times are part of the data.
+The approach was valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
+In general, even if stepping stone approaches are appropriate, we recommend to investigate $P(H_1|M)$ to investigate the prior assumption on the hypothesis to be tested.
+Such an investigation reveals if the cause of a high or low Bayes factor is due to the prior or due to signal in the data.
 

### 2015-09-08 fossilDating.tex
comparison text

--- 2015-09-08 fossilDating.tex
@@
 \begin{figure}
-\includegraphics[width=6in]{priors.pdf}
+\includegraphics[width=12cm]{priors.pdf}
 \caption{\label{hist8_older}Probability density for the sampling times under the fossilized birth-death process.
@@
 Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
-Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.905$. 
-The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
+Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.903$. 
+The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
 
@@
 \Mrelaxed{} was the best-fitting model for this data set according to the analysis of \textcite{gavryushkina2015bayesian}. 
-As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.927$.
+As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
 The median error was 2.1 Myr across all 36 fossils.
@@
 Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+The main difference is that the relaxed model appears to be a better fit to the data.
+Under \Mrelaxed{} all fossils have positive evidence for their geological age range, whereas under \Mstrict{} there are a handful of fossils with negative evidence for the corresponding geological age range.
+Furthermore, assuming the median geological age is the truth, the variance in the phylogenetic estimation error of the fossil ages is larger under \Mstrict{} than under \Mrelaxed{}.
 
@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model.
+We found that although a strict morphological clock does a surprisingly good job of estimated fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
 There are a diverse array of potential applications for this methodology. 

### 2015-09-08 fossilDating.tex
small changes
many small edits

@@
 We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
-In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which it was excavated. 
+In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
 The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
@@
 Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny.
-The age of the fossil is determined either by radiometric aging of strata above and/or below the strata, or more commonly by biostratigraphy.
-The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \autocite{Ho2009, Heled2012, HeledDrummond2015}.
+The age of the fossil is determined either by radiometric aging of strata above and/or below the fossil, or more commonly by biostratigraphy.
+The difficulty lies in determining the appropriate ancestral divergence in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \autocite{Ho2009, Heled2012, HeledDrummond2015}.
 Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006,drummond2010}.
@@
 
-We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters.
+We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters.
 This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
 It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
-This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 
-We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
+This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa. 
+We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
 The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
@@
 \textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach \autocite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
-We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding palaeontological age range based on radiometrically-dated geological strata.
+We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
 The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.
 
-A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree or chrono-phylogeny. 
+A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree. 
 Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. 
@@
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}. 
-In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
-These three questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of canids \autocite{Slater2015}.
+In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
+These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of canids \autocite{Slater2015}.
 
-\subsection*{Estimating the phylogenetic age of penguin fossils}
-
-For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,T_\text{max})$ Mya.
-\todo{This needs to change to the new prior!} 
-For the penguins we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny.
-For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
-The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
-We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
-
-\subsection*{Bayes factor}
+\subsection*{Computing the phylogenetic evidence for an age range}
 
@@
 
-We simplify, using the equations for $p_i(t;d,r)$ given in \cite{Kendall1948}, to obtain,
+We simplify, using the equations for $p_i(t;d,r)$ given in \autocite{kendall1948}, to obtain,
 \begin{eqnarray*}
@@
 
+\subsection*{Estimating the phylogenetic ages of penguin fossils}
+
+For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad model-based prior on the interval $(0,T_\text{max})$ Mya.
+For the penguin data set we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper prior limit on the origin of the whole phylogeny.
+For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
+The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
+We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
+
 \subsection*{Applying a morphological clock to Canid evolution}
@@
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
-For this data set we did not have a geological age range for each fossil, but instead the author of the study provided a point estimate for each fossil species' geological age.
+For this data set we did not have a geological age range for each fossil.
+Instead used the point estimates for each fossil species' geological age that were employed by the author of the original study.
 
@@
 Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
-As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad uniform prior of $(0,T_\text{max})$ Mya.
+As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad model-based prior of $(0,T_\text{max})$ Mya, where $T_\text{max} = 80$ corresponded again to the upper limit of the prior on the origin of the clade.
 
@@
 
-As judged by Bayes factors, none of the fossils exhibited strong evidence (i.e.
-$|\text{log BF}| > 3.0$) that the phylogenetic age was inconsistent with the geological age range assuming a uniform prior from $(0,160)$ Mya.
-However if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
-These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.01, 0.0011 and 0.0007 respectively.
+As judged by Bayes factors, only one fossil exhibited strong evidence (i.e.
+$\text{log BF} < -3.0$) that the phylogenetic age was inconsistent with the geological age range.
+The log BF for {\em Paraptenodytes antarcticus} was -3.4.
+In fact the majority of the fossils (23/36 = 64\%) had strong positive evidence for the geological age range (i.e.
+$\text{log BF} > 3.0$).
+Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
+These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively.
 All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
@@
 antarcticus}: 29.9 vs 22, {\em S.
-muizoni}: 5.1 vs 9.1.
-The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, suggesting very precise geological knowledge of the ages of these three fossils.
+muizoni}: 5.2 vs 9.1.
+The small posterior probabilities are partially caused in these cases because the corresponding palaeontological age range is narrow, apparently suggesting very precise geological knowledge of the ages of these three fossils.
 
@@
 As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
-The median error was 2.1 Myr across all 36 fossils.
-Again, as with \Mstrict{}, none of the fossils exhibited strong evidence (i.e.
-$|\text{log BF}| > 3.0$) that the phylogenetic age was inconsistent with the geological age range assuming a uniform prior from $(0,160)$ Mya.
+The median error was 2.05 Myr across all 36 fossils.
+In this analysis none of the fossils exhibited any evidence (i.e. $\text{log BF} < 0.0$) that the phylogenetic age was inconsistent with the geological age range. 
 However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
-These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.024, 0.016, 0.048, 0.007, 0.037 respectively.
+These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively.
 All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.5Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
-antarcticus}: 28.1 vs 22, {\em P.
+mirandus}: 6.7Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
+antarcticus}: 28.0 vs 22, {\em P.
 devriesi}: 49.0 vs 40, {\em S.
 muizoni}: 5.1 vs 9.1 and {\em W.
-manneringi}: 56.8 vs 61.05.
+manneringi}: 56.7 vs 61.05.
 A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}. 
@@
 The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
-Only 16 out of 123 fossils (13\%) don’t have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
+Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
 The median error is $1.66$My, which is even more accurate than the penguin data set.
-There are less characters than the penguins (122 versus $\sim 250$) so the uncertainty in the individual age estimates is greater (median 99\% HPD range = 6.5 My), but overall it seems to work very nicely.
+This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision (median 99\% HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
 
@@
 We found that although a strict morphological clock does a surprisingly good job of estimated fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
+We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
+This may suggest that the larger reference set of fossils in the canid data set (122 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates.
+
 There are a diverse array of potential applications for this methodology. 
@@
 
-\todo{How does the number of fossils in the reference affect the precision and accuracy of the fossil age estimates ?}
+Future work may investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
+
+In the mean time we anticipate that this approach will help to promote the application of a consistent probabilistic framework to consider both molecular and fossil evidence.
+Our results are encouraging in suggesting that the statistical models presented are adequate for inference of phylogenetic time-trees from morphological fossil data.
 

### 2015-09-08 fossilDating.tex
acknowledgements

@@
 
+\section*{Acknowledgements}
+
+The authors would like to thank Daniel Ksepka and Graeme T.
+Lloyd for insightful comments and suggestions on early drafts of this manuscript.
+AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).
+AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).
+
 \printbibliography

### 2015-09-08 fossilDating.tex
git address availability
added ronquist citation

@@
 This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa.
+A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \autocite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa of the other taxa in the tree \textcite{gavr2014}.
 We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
@@
 
+\section*{Availability}
+
+All BEAST2 xml input files and R analysis scripts required to reproduce the results in this paper are available at \url{https://github.com/alexeid/fossilDating}.
+
 \section*{Acknowledgements}

### 2015-09-08 fossilDating.tex
Tanja read through everything

--- 2015-09-08 fossilDating.tex
@@
 
-\title{Bayesian phylogenetic estimation of fossil ages } %reveals remarkable consistency of rocks and clocks}
-\author{Alexei Drummond, Tanja Stadler}
-\date{\today{}}
-\maketitle
+\bigskip
+\medskip
+\begin{center}
+
+\noindent{\Large \bf Bayesian phylogenetic estimation of fossil ages}
+
+\bigskip
+
+\noindent {\normalsize \sc  Alexei J.\ Drummond$^1$ and Tanja Stadler$^{2,3}$}\\
+\noindent {\small \it 
+$^1$Department of Computer Science, University of Auckland, Auckland, 1010, New Zealand;\\
+$^2$Department of Biosystems Science \& Engineering, Eidgen\"{o}ssische Technische Hochschule Z\"{u}rich, 4058 Basel, Switzerland;\\
+$^3$Swiss Institute of Bioinformatics (SIB), Switzerland.}
+\end{center}
+\medskip
+\noindent{\bf Corresponding author:} Alexei J. Drummond, Department of Computer Science, University of 
+Auckland, 38 Princes Street, Auckland, 1010, New Zealand; E-mail: alexei@cs.auckland.ac.nz\\
+
+%\title{Bayesian phylogenetic estimation of fossil ages } %reveals remarkable consistency of rocks and clocks}
+%\author{Alexei Drummond$^1$, Tanja Stadler$^{2,3}$}
+%\date{\today{}}
+%\maketitle
+%
 
@@
 The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
-We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', and for evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
+We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 
+%TS I do not understand this last subsentence
+and for evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses. %TStodo
 }
@@
 
-Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny.
+Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associating their geologically-derived age to a particular divergence in a molecular phylogeny.
 The age of the fossil is determined either by radiometric aging of strata above and/or below the fossil, or more commonly by biostratigraphy.
@@
 
+%TS: I suggest to remove the following heading
 \subsection*{Phylogenetic tip-dating}
@@
 This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa.
+A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \autocite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \textcite{gavr2014}.
 We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
-The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
+The method is applied to two rich and well-characterized morphological data sets: (i) 
+%TS how many extant penguins?
+extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
 
@@
 We refer to this estimate of a fossil's age as its {\em phylogenetic age}. 
-In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
-These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of canids \autocite{Slater2015}.
+In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
+These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \autocite{Slater2015}.
 
@@
 
-One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $\mathcal{T}$ and record the fraction of sampling times  within a given  paleontological age range.
-However, such a simulation approach turned out to be very time-consuming, and the procedure below provided a much faster evaluation of $p(H_1|M_\mathcal{T})$.
+One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $M_\mathcal{T}$ and record the fraction of sampling times  within a given  paleontological age range.
+However, such a simulation approach turns out to be very time-consuming, and the procedure below provided a much faster evaluation of $p(H_1|M_\mathcal{T})$.
 
-We  derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$.
+We  will now derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$.
 In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
@@
 We determined $p(t|M_\mathcal{T})$ for the prior distributions as in \textcite{gavryushkina2015bayesian}, 
-$$T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1).$$ This prior specification leads to a distribution of sampling time with almost all mass close to the present (Figure \ref{Fig:Prior}, dot-dashed line).
+$$T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1).$$ This prior specification leads to a distribution of sampling time with almost all probability mass close to the present (Figure \ref{Fig:Prior}, dot-dashed line).
 Thus, essentially $p(H_1|M_\mathcal{T})=0$  which leads to a huge Bayes Factor.
@@
 (ii) If we draw $r$ and $s$ close to 1, then we obtain a very large per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
-Thus these parameter combinations govern the probability density curve and cause most prior weight to be close to the present.
+Thus these parameter combinations govern the probability density curve and cause again most prior weight to be close to the present.
 
@@
 
-Further, we want to note that a classic  stepping stone approach to calculate Bayes factors would have not been possible in our case:  In stepping stone approaches the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
+We want to note that a classic  stepping stone approach
+%TS which paper to cite here?
+ to calculate Bayes factors would have not been possible in our case:  In stepping stone approaches the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
 However, using a birth-death model, the sampling times are part of the data.
@@
 In general, even if stepping stone approaches are appropriate, we recommend to investigate $P(H_1|M)$ to investigate the prior assumption on the hypothesis to be tested.
-Such an investigation reveals if the cause of a high or low Bayes factor is due to the prior or due to signal in the data.
+Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.
 
@@
 
-For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad model-based prior on the interval $(0,T_\text{max})$ Mya.
-For the penguin data set we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper prior limit on the origin of the whole phylogeny.
+%TS: check:
+For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by 
+the fossilized birth-death process prior as described above.
+%a broad model-based prior on the interval $(0,T_\text{max})$ Mya. 
+%For the penguin data set we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper prior limit on the origin of the whole phylogeny.
 For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
@@
 For this data set we did not have a geological age range for each fossil.
-Instead used the point estimates for each fossil species' geological age that were employed by the author of the original study.
+Instead we used the point estimates for each fossil species' geological age that were employed by the author of the original study.
 
@@
 As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad model-based prior of $(0,T_\text{max})$ Mya, where $T_\text{max} = 80$ corresponded again to the upper limit of the prior on the origin of the clade.
+%TS: which priors on d, r, s, T did we use now?
 
@@
 These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively.
-All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
+All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M.
@@
 
-\Mrelaxed{} was the best-fitting model for this data set according to the analysis of \textcite{gavryushkina2015bayesian}. 
+\Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \textcite{gavryushkina2015bayesian}. 
 As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
@@
 These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively.
-All other fossils has posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
+All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M.
@@
 The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. 
-Blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
+The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
 If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. 
-The three labelled fossils show significant inconsistency.}
+The three labelled fossils show significant inconsistency.
+%TS do these three fossils not intersect in horizontal AND vertical?
+}
 \end{figure}
@@
 Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+%TS is the following quantified? otherwise I'd delete the next sentence!?
 The main difference is that the relaxed model appears to be a better fit to the data. 
@@
 Comparison of \Mstrict{} against \Mrelaxed{}.
-(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue) to guide eye (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
+(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
 \end{figure}
@@
 The median error is $1.66$My, which is even more accurate than the penguin data set.
-This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision (median 99\% HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
+This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision 
+%TS why 99\% for candids and 95\% for penguins?
+(median 99\% HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
 
@@
 If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age. The 16 fossils for which this is the case are labelled.}
+%TS I find this plot very hard to read as many fossils are overlaying...can we do something there for better visualization?
 \end{figure}
@@
 AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).
+TS was supported in part by the European Research Council under the 7th Framework Programme of the European Commission (PhyPD; grant agreement 335529).
 

### 2015-10-29 TotalEvidenceTalk.tex
talk

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
%handout, if no \pause

\usepackage{tabularx}
\usepackage[]{algorithm2e}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
\bibliography{fossilDating.bib}

\newcommand{\Mstrict}{{M1}}
\newcommand{\Mrelaxed}{{M8}}

%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
%\usepackage{pgfpages}
%\pgfpagesuselayout{resize to}[a4paper, landscape]

\definecolor{headerColour}{RGB}{180, 230, 245}
\definecolor{headerTitleColour}{RGB}{0, 60, 80}
\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}

\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
\setbeamercolor{title}{fg=headerTitleColour, bg=white}
\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}

\usetheme{Goettingen}
\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother

%% to suppress subsections in sidebar:
%\setbeamertemplate{subsection in sidebar shaded}
%{\vspace*{-\baselineskip}}
%\setbeamertemplate{subsubsection in sidebar shaded}
%{\vspace*{-\baselineskip}}

\usepackage{amsmath, amssymb}
\usepackage{fancyvrb}

% Font modification
\usefonttheme{professionalfonts}
\usepackage{cmbright}
\usepackage{eulervm}
\newfont{\Ss}{cmcsc12 scaled 1600}

% no navigation symbols at the bottom of frames
\beamertemplatenavigationsymbolsempty
\setbeamertemplate{footline}[frame number]

\usepackage{multirow} % allows entries across multipe rows in tables
\usepackage{booktabs} % makes fancier rulers in tables
%\usepackage{setspace} % spacing between lines

%%%%%%%%%%%%%%%%%%
%% Bibliography related stuff:
% control space between lines
  \let\oldthebibliography=\thebibliography
  \let\endoldthebibliography=\endthebibliography
  \renewenvironment{thebibliography}[1]{
    \begin{oldthebibliography}{#1}
      \setlength{\parskip}{-0.5ex}
      \setlength{\itemsep}{-0.5ex}
  }{ \end{oldthebibliography} }
% Force entry to be in one line
\setbeamertemplate{bibliography entry title}{}
\setbeamertemplate{bibliography entry location}{}
\setbeamertemplate{bibliography entry note}{}
% Set bullet point shape:
\setbeamertemplate{bibliography item}{-}
%%%%%%%%%%%%%%

\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
\newcommand{\head}[1]
{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
\parskip0.3ex

\newcommand{\cT}{{\mathcal T}}

%\defbeamertemplate*{title page}{customized}[1][]
%{
%  \titlepage
%  \usebeamerfont{title}\inserttitle\par
%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
%  \bigskip
%  \usebeamerfont{author}\insertauthor\par
%  \usebeamerfont{institute}\insertinstitute\par
%  \usebeamerfont{date}\insertdate\par
%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
%}

\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
\date{The Royal Society, London \\ 9th November 2015}
\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
%}

%%%%%%%%%%%%%% Figure caption setup
\usepackage[compatibility=false]{caption}
\captionsetup[figure]{labelsep=space,justification=centering}
\renewcommand{\figurename}{\scriptsize Figure adapted from}

% syntax: \figureCaption{Citation caption}{Second (real) caption}
% second caption can be skipped, first one has to be suppressed if you want to skip it
\newcommand\figureCaption[2]{%
  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
  \caption{\scriptsize #1}
  \caption*{#2}
}

%%%%%%%%%%%%%% Fixme and comment setup¤
\definecolor{red}{HTML}{C92D39}
\definecolor{green}{HTML}{498A44}

% syntax: \fixme{What needs to be fixed}
\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
% Hide all fixmes by switching the line above to this:
%\newcommand{\fixme}[1]{}

% syntax: \comment{Name of commenter}{C omment}
\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
% Hide all comments by switching the line above to this:
%\newcommand{\comment}[2]{}

\begin{document}

%Slide1
\frame{\titlepage}

\section{Fossilized birth-death process}

\begin{frame}
\frametitle{The fossilized birth-death process}

\end{frame}

\section{Penguin dating}

\begin{frame}
\begin{figure}
\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
\label{saPostPr}
\end{figure}
\small{Probabilities of each fossil to be a sampled ancestor across models.

{\it P} stands for the partitioned model, {\it G} for gamma variation across sites, {\it Mkv} for conditioning on 
variable characters, {\it R} for relaxed clock model, {\it dns} for $d$, $\nu$, and $s$ tree prior parameterisation, {\it lmp} for $
\lambda$, $\mu$, and $\psi$ tree prior parameterisation.} 
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
\label{postPredictive}
\end{figure}
\small{The posterior and posterior predictive distributions for (left) tree length, $T$, and genealogical $D_F$ statistics and (right) $B_1$ tree imbalance statistic and Colless's tree imbalance index, $I_c$, for model 8.

The posterior trees are on the unbalanced end of the predictive distribution.}
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
\label{penguinTree}
\end{figure}
\footnotesize{A maximum sampled ancestor clade credibility tree for the total-evidence analysis. % with mean node ages. 
The numbers in blue show the posterior supports of clades. The filled red and black circles represent sampled 
ancestors. Fossils with positive evidence of being sampled ancestors are shown in red.}

\end{frame}

\begin{frame}
\begin{figure}
\begin{center}
\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
\label{fig:BF}
\end{center}
\end{figure}
\small{The evidence fossil sampled ancestors. The samples above the shaded area have positive evidence to be sampled ancestors and below the shaded area have positive evidence to be terminal nodes.}

\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
\label{fig:divDates}
\end{figure}
\small{The posterior summary tree of extant penguins with 95\% HPD intervals and
posterior supports of clades after removing fossils.}
\end{frame}

\begin{frame}
\begin{figure}
\begin{center}
\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
\label{fig:rootAges}
\end{center}
\end{figure}
\small{The age of root of extant penguins across models including total-evidence analysis under relaxed (8+DNA R) and strict (8+DND S) molecular clocks and 
total-evidence analysis with crown fossils only.}
\end{frame}

\begin{frame}
\frametitle{Dating the evolutionary history of penguins}
\begin{figure}
\includegraphics[width=\textwidth]{penguinTree.pdf}
\end{figure}
\end{frame}

\section{Fossil dating}

\begin{frame}
\begin{figure}
\includegraphics[width=0.6\textwidth]{priors.pdf}
\label{Fig:Prior}
\end{figure}

Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{Figure1.pdf}
\end{figure}
Bayesian phylogenetic age estimates for each of 36 penguin fossils plotted against their palaeontological age estimates, under two alternative evolutionary models. 
The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
\end{figure}
Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
\end{figure}
Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}. Red boxes are the superimposed age ranges derived from geological data.
\end{frame}

\begin{frame}
\begin{figure}
\includegraphics[width=\textwidth]{Figure2.pdf}
\end{figure}
A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e. the precision is 1/variance in the marginal posterior distribution of the age).
\end{frame}

\begin{frame}
\frametitle{Comparison of age \& BF estimates between models}
\begin{figure}%
\centering
\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
\label{fig:compareM1M8ab}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Comparison of probability \& error between models}
\begin{figure}%
\centering
\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
\label{fig:compareM1M8cd}
\end{figure}
\end{frame}

\end{document}

### 2015-10-29 fossilDating.tex
talk

@@
 
-\noindent {\normalsize \sc  Alexei J.\ Drummond$^1$ and Tanja Stadler$^{2,3}$}\\
+\noindent {\normalsize \sc  Alexei J.\ Drummond$^{1,2}$ and Tanja Stadler$^{2,3}$}\\
 \noindent {\small \it 
@@
 Auckland, 38 Princes Street, Auckland, 1010, New Zealand; E-mail: alexei@cs.auckland.ac.nz\\
-
-%\title{Bayesian phylogenetic estimation of fossil ages } %reveals remarkable consistency of rocks and clocks}
-%\author{Alexei Drummond$^1$, Tanja Stadler$^{2,3}$}
-%\date{\today{}}
-%\maketitle
-%
 
@@
 We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 
-%TS I do not understand this last subsentence
-and for evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses. %TStodo
+and for uncovering potential model misspecification and/or data errors when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
 }
@@
 %TS: I suggest to remove the following heading
-\subsection*{Phylogenetic tip-dating}
+%\subsection*{Phylogenetic tip-dating}
 
@@
 %TS how many extant penguins?
-extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
+19 extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
 
@@
 These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \autocite{Slater2015}.
+
+\subsection*{Estimating the phylogenetic ages of penguin fossils}
+
+%TS: check:
+For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by 
+the fossilized birth-death process prior as described below.  
+For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
+The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
+We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
+
+\subsection*{Estimating the phylogenetic ages of Canid fossils}
+
+The second data set that we investigated was a morphological data matrix of 132 canid species \autocite[9 extant and 123 fossil;][]{Slater2015}. 
+The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
+For this data set we did not have a geological age range for each fossil.
+Instead we used the point estimates for each fossil species' geological age that were employed by the author of the original study.
+
+We applied the M1 model to this data set assuming a fixed geological age for each taxon.
+Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
+As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn.
+Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \textcite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
+To estimate the phylogenetic age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
+%TS: which priors on d, r, s, T did we use now?
 
@@
 
-We want to note that a classic  stepping stone approach
+We want to note that the stepping stone sampling approach \autocite{Xie2011}
 %TS which paper to cite here?
- to calculate Bayes factors would have not been possible in our case:  In stepping stone approaches the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
+ to calculate Bayes factors would not have been possible in our case:  In stepping stone sampling the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
 However, using a birth-death model, the sampling times are part of the data.
@@
 Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.
-
-\subsection*{Estimating the phylogenetic ages of penguin fossils}
-
-%TS: check:
-For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by 
-the fossilized birth-death process prior as described above.
-%a broad model-based prior on the interval $(0,T_\text{max})$ Mya. 
-%For the penguin data set we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper prior limit on the origin of the whole phylogeny.
-For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
-The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
-We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
-
-\subsection*{Applying a morphological clock to Canid evolution}
-
-The second data set that we investigated was a morphological data matrix of 132 canid species \autocite[9 extant and 123 fossil;][]{Slater2015}. 
-The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
-For this data set we did not have a geological age range for each fossil.
-Instead we used the point estimates for each fossil species' geological age that were employed by the author of the original study.
-
-We applied the M1 model to this data set assuming a fixed geological age for each taxon.
-Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
-As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad model-based prior of $(0,T_\text{max})$ Mya, where $T_\text{max} = 80$ corresponded again to the upper limit of the prior on the origin of the clade.
-%TS: which priors on d, r, s, T did we use now?
 
@@
 If the horizontal line doesn't cross $x=y$, then the median phylogenetic estimate is not contained in the palaeontological age range. 
-The three labelled fossils show significant inconsistency.
+The three labelled fossils have posterior probability of less than 0.05 for their age being within the palaeontological age interval.
 %TS do these three fossils not intersect in horizontal AND vertical?
@@
 %TS is the following quantified? otherwise I'd delete the next sentence!?
-The main difference is that the relaxed model appears to be a better fit to the data. 
 Under \Mrelaxed{} all fossils have positive evidence for their geological age range, whereas under \Mstrict{} there are a handful of fossils with negative evidence for the corresponding geological age range.
 Furthermore, assuming the median geological age is the truth, the variance in the phylogenetic estimation error of the fossil ages is larger under \Mstrict{} than under \Mrelaxed{}.
+%AJD added
+This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \autocite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.
 
@@
 Comparison of \Mstrict{} against \Mrelaxed{}.
-(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
+(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
 \end{figure}
@@
 %TS why 99\% for candids and 95\% for penguins? 
-(median 99\% HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
+%AJD I will check this later.
+(median HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
 
@@
 TS was supported in part by the European Research Council under the 7th Framework Programme of the European Commission (PhyPD; grant agreement 335529).
+AJD and TS also thank ETH Z\"{u}rich for funding.
 

### 2015-10-29 8_penguins_eNprior/8_fossilDating.R

@@ -7,6 +7,9 @@ regexp = "\\-[0-9]+\\.log"
 seq <- read.table("~/Git/fossilDating/sorted_sequences.txt", header=T, sep="\t")
 seq$Sequence <- as.character(seq$Sequence)
 
+h1prob <- read.table("~/Git/fossilDating/1_penguins_eNprior/H1prob-final.txt", sep=" ", header=T)
+fossilTable8$prior <- h1prob$fossilH1
+
 result8 <- loadFossilFiles(".", regexp, fossilTable8, 10001)
 height8 <- result8$heights
 df8 <- createSummaryTable(result8$names, height8, fossilTable8, seq)
@@ -26,6 +29,10 @@ createPrecisionVsKnownCharactersFigure("8_precisionVsKnownCharacters.pdf", df8, 
 
 # Need to split younger and older ones
 
-createHistPlotFigures("8_fossilDatingHist_younger", result8$names[df8$geo < 32], height8[df8$geo < 32], fossilTable8[df8$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+createHistPlotFigures("8_fossilDatingHist_younger", result8$names[df8$geo < 32], height8[df8$geo < 32], fossilTable8[df8$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45, numrow=6, numcol=3)
 
-createHistPlotFigures("8_fossilDatingHist_older", result8$names[df8$geo > 32], height8[df8$geo > 32], fossilTable8[df8$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)
+createHistPlotFigures("8_fossilDatingHist_older", result8$names[df8$geo > 32], height8[df8$geo > 32], fossilTable8[df8$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70, numrow=6, numcol=3)
+
+createHistPlotFigures("8_fossilDatingHist_younger_wide", result8$names[df8$geo < 32], height8[df8$geo < 32], fossilTable8[df8$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45, numrow=3, numcol=6, pdfWidth=12, pdfHeight=6, histCol=rgb(0.1, 0.1, 0.9, 1.0))
+
[…]

### 2015-10-29 compareM1M8.r

@@ -1,4 +1,8 @@
 setwd("~/Git/fossilDating")
+
+# before running this script you must run 8_penguins_eNprior/8_fossilDating.R and 
+# 1_penguins_eNprior/8_fossilDating.r, so get df1 and df8 variables
+
 createPhyloAgeVsPhyloAgePDF("compareAgeM1M8.pdf",df8,df1,"Model 8", "Model 1",c(),c())
 
 pdf(file="comparePostM1M8.pdf", width=5, height=5, pointsize=10)
@@ -10,6 +14,15 @@ r2 <- format(summary(lm)$r.squared,digits=3)
 eq <- bquote(bold(R^2 == .(r2)))
     
 text(0.05,0.925, labels=eq,pos=4)
+
+ind <- 1:nrow(df8)
+ind <- ind[abs(df8$post - df1$post)>0.25]
+orient <- rep(4,length(ind))
+
+# Label outliers
+if (length(ind) > 0) {
+  text(df8$post[ind], df1$post[ind], labels=df1$names[ind],pos=orient, offset=0.4, cex=0.7)
+}
 dev.off()
 
 logbf1 <- log(df1$bf)
@@ -20,10 +33,20 @@ par(mar = c(4,4,1,1) + 0.1)
 plot(logbf8, logbf1, xlab="log(BF) in Model 8", ylab="log(BF) in Model 1", pch=16, col="blue")
 lm <- lm(logbf1 ~ logbf8)
 abline(lm, col="red")
+#lines(c(-1,7), c(-1,7), col="blue")
 r2 <- format(summary(lm)$r.squared,digits=3)
 eq <- bquote(bold(R^2 == .(r2)))
     
 text(0.5,4.1, labels=eq,pos=4)
+
+ind <- 1:nrow(df8)
+ind <- ind[abs(logbf8 - logbf1)>2]
+orient <- rep(4,length(ind))
+
+# Label outliers
+if (length(ind) > 0) {
+  text(logbf8[ind], logbf1[ind], labels=df1$names[ind],pos=orient, offset=0.4, cex=0.7)
+}
 dev.off()
 
 pdf(file="compareErrorM1M8.pdf", width=5, height=5, pointsize=10)
@@ -35,4 +58,14 @@ r2 <- format(summary(lm)$r.squared,digits=3)
 eq <- bquote(bold(R^2 == .(r2)))
     
 text(0.5,15.5, labels=eq,pos=4)
+
[…]

### 2015-10-29 fossilDatingFunctions.r

@@ -294,12 +294,12 @@ createDensityPlotFigures <- function (filePrefix, names, height, fossilTable, co
 ####################################################################
 # produce the hist plots and write to pdf files
 ####################################################################
-createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combinedPlots, xmin=0, xmax=66,binsPerMY=1) {
+createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combinedPlots, xmin=0, xmax=66,binsPerMY=1,numcol=3, numrow=6, pdfWidth=8, pdfHeight=11.5, histCol="dark gray") {
 
   if (combinedPlots) {
   
-    pdf(file=paste(filePrefix, ".pdf", sep=""), width=8, height=11.5)
-    par(mfrow = c(7,3),
+    pdf(file=paste(filePrefix, ".pdf", sep=""), width=pdfWidth, height=pdfHeight)
+    par(mfrow = c(numrow,numcol),
         oma = c(4,4,0,0) + 0.1,
         mar = c(1,1,2,1) + 0.1,
         mgp=c(2.5,0.65, 0),lwd=0.5)
@@ -312,7 +312,7 @@ createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combi
       pdf(file=paste(fossilTable$File.name[i],".pdf",sep=""))
     }
     
-    hist <- hist(height[[i]], xlim=c(xmin,xmax),breaks=c((0:(xmax*binsPerMY)/binsPerMY),10*xmax),freq=FALSE, xlab="",main="",col="dark gray", border="white")
+    hist <- hist(height[[i]], xlim=c(xmin,xmax),breaks=c((0:(xmax*binsPerMY)/binsPerMY),10*xmax),freq=FALSE, xlab="",main="",col=histCol, border="black")
     title(main = list(names[i], cex = 1.0))
     #rect(lower, 0.0, upper, density, col=rgb(1.0,0.0,0.0,0.5))
     
@@ -327,7 +327,10 @@ createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combi
     y <- c(0.0, density, density, 0.0)
     
     
-    lines(x,y, col="red",lwd=1)
+    rect(lower, 0.0, upper, density, col=rgb(1,0.25,0.25,0.5),border=FALSE)
+    lines(x,y, col="black",lwd=1)
+    
+    #plot(hist, add=T)
 
     if (!combinedPlots) {
       dev.off()

### 2015-10-29 TotalEvidenceTalk.tex
homo naledi

@@
 \usepackage{tabularx}
+\usepackage{tikz}
 \usepackage[]{algorithm2e}
@@
 \newcommand{\Mrelaxed}{{M8}}
+% turnover = birth / death
+\newcommand{\turnover}{\nu}
+
+% sampling proportion = \psi / \psi + \mu
+\newcommand{\fosp}{s}
+\newcommand{\sampleprop}{sampling proportion}
+
+\definecolor{newblue}{rgb}{0.3, 0.5, 0.9}
+\definecolor{newblue1}{rgb}{0.3, 0.3, 0.7}
+\definecolor{newred}{RGB}{200,0,0}
 
@@
 
-\begin{frame}
-\frametitle{The fossilized birth-death process}
+\begin{frame}{The problem}
+
+\begin{itemize}
+
+\item We have 
+\begin{itemize}
+  \item \normalsize{{\bf molecular data} of extant species,} 
+  \item \normalsize{{\bf morphological data} of extant and fossil species and }
+  \item \normalsize{{\bf Fossil ages} (or geological time intervals).}
+\end{itemize} 
+\item We want to utilise all this data to learn about {\bf evolutionary history} and {\bf macroevolutionary processes}.
+\item {\bf Bayesian statistical inference} is our preferred method of learning about the world.
+
+\end{itemize}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Node dating}
+
+\begin{itemize}
+\item {\bf Divergence time dating is tricky :) }
+\item {\bf A common practice is node dating}. One or more divergences in a phylogeny are assigned ages by identifying fossils of known geological age that correspond* to each divergence. 
+\item This method has a few drawbacks:
+\begin{itemize}
+\item Often only the oldest fossil in a calibrated clade is used,
+\item Calibration densities are often {\it ad hoc},
+\item Difficult* to combine calibrations with tree priors,
+\item Computational complexity of some methods mean that only a few nodes can be calibrated (Heled \& Drummond 2012).
+\end{itemize}
+\end{itemize}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Tip dating and total evidence}
+
+Two main features of new method for divergence-time dating:
+\begin{itemize}
+\item Explicit modelling of fossilization events as a part of the tree branching process (Pyron 2011, Heath {\it et al} 2014),
+\item Utilising all existing data \textcolor{newblue1}{(total evidence)} in a joint inference (Pyron 2011, Ronquist {\it et al} 2012).
+\end{itemize}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\begin{frame}{Total evidence}
+
+%\begin{itemize}
+
+%\item Pyron 2011 uses Lewis MK model to model evolution of morphological characters.  
+%\item Ronquist et al 2012, Wood et al 2012. 
+%\item They used Yule or Uniform tree prior. 
+%\end{itemize}
+
+%\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Birth-death-sampling models}
+
+\begin{figure}
+  \begin{center}
+               \input{pics/SampledTree1}
+            \end{center}
+
+         \end{figure}    
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Birth-death-sampling models}
+
+ \begin{figure}
+            \begin{center}
+               \input{pics/SampledTree2}
+            \end{center}
+
+         \end{figure}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame} {Fossillized birth-death model (FBD)}
+
+\begin{columns}[t]
+\begin{column}{.5\textwidth}
+
+Stadler 2010, Heath {\it et al} 2014. 
+\vspace{0.2cm}
+
+The process starts at time $t_{or}>0$ and ends at time zero (present time).  
+  \vspace{0.01cm}
+   \begin{itemize}
+\item birth rate $\lambda$
+\item death rate $\mu$
+\item sampling rate $\psi$
+\item sampling at present probability $\rho$ 
+   \end{itemize}
+\vspace{0.01cm}
+
+   \end{column}
+   \begin{column}{.5\textwidth}
+
+            \begin{center}
+               \input{pics/SATree2}
+            \end{center}
+
+         \begin{center}\emph{Sampled tree} \end{center}
+
+   \end{column}    
+\end{columns}
+
+\vspace{0.2cm}
+
+Model parameters: \color{newblue1} $\eta = (t_{or}, \lambda, \mu, \psi, \rho)$.
+
+\color{black}
+All the parameters are identifiable.
+
+\vspace{1.5cm}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame} {Fossilized birth-death (FBD) skyline model}
+
+Stadler \& K\"uhnert et al 2012, Gavryushkina {\it et al} 2014.
+
+\vspace{0.2cm}
+
+There are $k$ time intervals and parameters remain constants within the intervals but may vary from one interval to another
+
+\begin{columns}[c]
+\begin{column}{.5\textwidth}
+\begin{itemize}
+\item birth rates $\lambda_1, \ldots, \lambda_k$
+\item death rates $\mu_1, \ldots, \mu_k$
+\item sampling rates $\psi_1, \ldots, \psi_k$
+\item sampling at time $t_k$ (present) probability $\rho$
+\end{itemize}
+
+\begin{center}
+Model parameters:
+\color{newblue1} $\eta = (t_{or}, \bar \lambda, \bar \mu, \bar \psi, \rho)$
+
+\end{center}
+  \end{column}
+   \begin{column}{.5\textwidth}
+\begin{center}
+\input{pics/SkylineTree}
+\end{center}
+\begin{center}\emph{\small Sampled tree} \end{center}   
+ \end{column}    
+\end{columns}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\iffalse
+
+\begin{frame}{Variation in the choice of the tree model parameters}
+
+The $t_{or}$ parameter can be replaced with $t_{root}$ parameter assuming the process started with a branching event.
+
+\vskip2mm
+
+Instead of $\lambda$, $\mu$, and $\psi$ we can also use an alternative parameterisation:
+
+\renewcommand{\arraystretch}{1.2}
+
+\begin{equation*} 
+\begin{tabular}{ll} 
+\text{net diversification rate} & $d = \lambda-\mu$  \\ 
+\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
+\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
+\end{tabular} \end{equation*}
+
+\end{frame}
+
+\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Heath {\it et al} approach and its extensions}
+
+ Heath {\it et al} 2014 first used FBD model to infer divergence times of bears in a Bayesian MCMC framework:
+
+\begin{itemize}
+\item The topology of extant species is fixed
+\item Only used occurrence dates of fossils, and no morphology 
+\item Fixed $\rho$ to the truth in the inference
+\end{itemize}
+
+We extend this model in two ways:
+
+\begin{itemize}
+\item Sampling sampled ancestor trees (Gavryushkina {\it et al} 2014)
+\item Incorparating morphological data
+\end{itemize}
+
+Implemented in SA and Morph-models packages for BEAST2 (\url{www.beast2.org})
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Comparative data of fossil tips vs only occurrence dates}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{pics/Height_SA/Rplots.pdf}
+\end{figure}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Comparative data of fossil tips vs only occurrence dates}
+
+\setlength{\oddsidemargin}{0.1in} % adjust as necessary
+
+\begin{columns}[c]
+
+\begin{column}{.7\textwidth}
+\begin{center}
+\begin{figure}
+\includegraphics[height=0.65\textheight]{pics/TurnoverHPD/Rplots.pdf}
+\end{figure}
+\end{center}
+ \end{column}
+
+\begin{column}{.3\textwidth}
+\begin{center}
+turnover~rate~=~$\frac \mu \lambda$ 
+\end{center}
+ \end{column}    
+\end{columns}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Biased estimates when not modelling sampled ancestors}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{pics/SAvsNoSA/Rplots.pdf}
+\end{figure}
+
+\begin{center}
+diversification rate = $\lambda-\mu$ 
+\end{center}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Analysis of penguin morphological data}
+
+Penguin dataset (Ksepka et al 2012) consisting of morphological data of:
+\begin{itemize}
+\item all extant penguins (19 species)
+\item 36 fossil species assigned to stratigraphic intervals
+\end{itemize}
+
+\vskip2mm
+
+Models:
+
+\begin{itemize}
+\item Lewis MK vs MKv (Lewis 2001)
+\item Partitions vs single alignment
+\item Rate variation across sites and across partitions
+\item Tree prior: FBD and Skyline FBD
+\item Different tree model parameterisations
+\end{itemize}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Parameterisations}
+
+\renewcommand{\arraystretch}{1.2}
+
+lmp:
+
+\color{newblue1}
+
+\begin{equation*} 
+\begin{tabular}{ll} 
+\text{birth rate} & $\lambda$  \\ 
+\text{death rate} & $\mu$ \\ 
+\text{sampling rate} & $\psi$  \\ 
+\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
+\end{tabular} \end{equation*}
+
+\textcolor{black}{dvs:}
+
+\begin{equation*} 
+\begin{tabular}{ll} 
+\text{net diversification rate} & $d = \lambda-\mu$  \\ 
+\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
+\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
+\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
+\end{tabular} \end{equation*}
 
@@
 
-\begin{frame}
+\begin{frame}{But what about those hyper parameter priors?}
 \begin{figure}
@@
 
-Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
+\small{Prior density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
+Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.}
+\end{frame}
+
+\begin{frame}{And what about this guy?}
+\includegraphics[width=0.5\textwidth]{Homo_naledi.jpg}
+\includegraphics[width=0.5\textwidth]{Atlantic_Homo_naledi.jpg}
+
+\bigskip{}
+
+{\bf Can we date fossil remains using Bayesian phylogenetics? }
+
 \end{frame}

### 2015-10-30 TotalEvidenceTalk.tex
yes we can
talk
hominins and canids

@@
 \title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
+\author[]{Alexei Drummond§}
 \date{The Royal Society, London \\ 9th November 2015}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Penguin dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Penguin dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}
@@
 \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Fossil dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fossil dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}{And what about this guy?}
@@
 
-{\bf Can we date fossil remains using Bayesian phylogenetics? }
+\end{frame}
+
+\begin{frame}{Can we date fossils using Bayesian phylogenetics? }
+
+Given:
+\begin{itemize}
+\item A reference set of fossils of known age, and
+\item A morphological characterization for reference fossils, and
+\item The fossilized birth-death model of sampled ancestor trees
+\item A ``clock-like'' model of morphological evolution 
+\item A morphological characterization of a related focal fossil of unknown age 
+\end{itemize}
+
+{\bf Can the age of the focal fossil be estimated using Bayesian phylogenetic inference?}
+
+\medskip{}
+
+We tested this hypothesis by using the penguin data set.
+For each of the 36 fossils in turn, we discarded the age information of the focal fossil to mimic the scenario where the age was unknown, and sought to estimate its age using the reference 35 fossils as the reference set.
 
@@
 
-\begin{frame}
+\begin{frame}{Estimating the phylogenetic age of a fossil}
 \begin{figure}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Canids}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\begin{figure}
+\includegraphics[width=0.70\textwidth]{canids/1c_phyloAgeVsGeoAge.pdf}
+\end{figure}
+{\small Bayesian phylogenetic ages for each of 123 canid fossils plotted against palaeontological ages, under \Mstrict{}. 
+Bayesian estimates are represented by the median and the 99\% HPD interval of posterior. Blue line shows the $x=y$. 
+16 fossils with inconsistent estimates are labelled.}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{canids/1_canids-1440516016976-tree5001.pdf}
+\end{figure}
+A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Hominins}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Controversies in human evolution: {\it Ardipithecus ramidus}}
+\begin{figure}
+\includegraphics[width=\textwidth]{ardi_crop.jpg}
+\end{figure}
+`Ardi', 4.4M years old was described as a the oldest fossil relative on the human lineage, a {\it hominin}, and thus more closely related to the human than the chimpanzee. 
+\end{frame}
+
+\begin{frame}{Hominin phylogeny inferred using FBD tree prior \& morphological clock.}
+\begin{figure}
+\includegraphics[width=0.75\textwidth]{homininSummaryTree.pdf}
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Conclusions and Future Directions}
+
+\begin{itemize}
+\item FBD (skyline) is an elegant solution to the divergence dating problem
+\item Bayesian inference implementations of FBD on sampled ancestor trees are already quite computationally tractable
+\item The effects of hyper parameter priors need to be examined carefully
+\item The {\bf morphological clock hypothesis seems adequate} to estimate fossil ages in a range of well-characterized fossil data sets.
+\item Summarizing sampled ancestor trees is more complicated and more work on summary and visualization are needed.
+\item Improved models of fossilization, speciation and extinction are obvious candidates for future work.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Acknowledgements}
+\begin{itemize}
+\item Sasha Gavryushkina, University of Auckland
+\item Tanja Stadler, ETH
+\item David Welch, University of Auckland
+\item Tracy Heath, Iowa State University
+\item Daniel Ksepka, Bruce Museum, Greenwich
+\end{itemize}
+\end{frame}
+
 \end{document}

### 2015-10-30 pics/SampledTree1.tex
sampled ancestor tree pics

\usetikzlibrary{shapes,snakes}

\begin{tikzpicture}[thick, scale=0.6, color=gray]

\begin{scope}

%terminal nodes
\node at (-3.5, 2.5)(s2){};
\node at (0, 1)(s3){};
\node at (-2.5, 1)(s4){};
\node at (4, 3.1)(s5){};
\node at (1.4, 3.5)(s7){};
\node at (-1, 1)(s8){};
\node at (3.2, 1)(s9){};
\node at (2, 1)(s10){};

%branches
\draw (0, 7.5) -- (0,7) -- (-2, 7) --(-2, 5.5) --(-3, 5.5) -- (-3, 4.9) -- (-3.5, 4.9) -- (s2);
\draw (-3, 4.9) -- (-2.5, 4.9) -- (s4); 
\draw (-2, 5.5) -- (-1, 5.5) -- (s8); 
\draw (0,7) -- (2,7) -- (2,5.8) -- (0.7, 5.8) -- (0.7, 4.5) -- (0, 4.5) -- (s3); 
\draw (0.7, 4.5) -- (1.4, 4.5) -- (s7); 
\draw (2, 5.8) -- (3.3, 5.8) -- (3.3, 4) -- (2.6, 4) -- (2.6, 2.1) -- (2, 2.1) -- (s10); 
\draw (2.6, 2.1) -- (3.2, 2.1) -- (s9);
\draw (3.3, 4) -- (4, 4) -- (s5);

\node at (0, -1) {Full tree};

\pause

%sampled nodes
\node[fill,circle, inner sep=2pt] at (-2.5, 1)(sA){};
\node[anchor=north] (Alet) at (sA) {\textcolor{black}{\footnotesize A}};
\node[fill,circle, inner sep=2pt] at (0, 1)(sB){};
\node[anchor=north] (Alet) at (sB) {\textcolor{black}{\footnotesize B}};
\node[fill,circle, inner sep=2pt] at (2, 1)(sC){};
\node[anchor=north] (Alet) at (sC) {\textcolor{black}{\footnotesize C}};
\node[fill,circle, inner sep=2pt] at (3.2, 1)(sD){};
\node[anchor=north] (Alet) at (sD) {\textcolor{black}{\footnotesize D}};

\pause

\node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
\node[anchor=west] (Alet) at (sE) {\textcolor{black}{\footnotesize E}};
\node[fill,circle, inner sep=2pt] at (-1, 2.9)(sF){};
\node[anchor=west] (Alet) at (sF) {\textcolor{black}{\footnotesize F}};

\pause

\color{newblue1}

%branches
\draw (0, 7.5) -- (0,7) -- (-2, 7) --(-2, 5.5) --(-3, 5.5) -- (-3, 4.9);
\draw (-3, 4.9) -- (-2.5, 4.9) -- (s4); 
\draw (-2, 5.5) -- (-1, 5.5) -- (sF); 
\draw (0,7) -- (2,7) -- (2,5.8) -- (0.7, 5.8) -- (0.7, 4.5) -- (0, 4.5) -- (s3); 
\draw (2, 5.8) -- (3.3, 5.8) -- (3.3, 4) -- (2.6, 4) -- (2.6, 2.1) -- (2, 2.1) -- (s10); 
\draw (2.6, 2.1) -- (3.2, 2.1) -- (s9);

%sampled nodes
\node[fill,circle, inner sep=2pt] at (-2.5, 1)(sA){};
\node[anchor=north] (Alet) at (sA) {\textcolor{black}{\footnotesize A}};
\node[fill,circle, inner sep=2pt] at (0, 1)(sB){};
\node[anchor=north] (Alet) at (sB) {\textcolor{black}{\footnotesize B}};
\node[fill,circle, inner sep=2pt] at (2, 1)(sC){};
\node[anchor=north] (Alet) at (sC) {\textcolor{black}{\footnotesize C}};
\node[fill,circle, inner sep=2pt] at (3.2, 1)(sD){};
\node[anchor=north] (Alet) at (sD) {\textcolor{black}{\footnotesize D}};

\node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
\node[anchor=west] (Alet) at (sE) {\textcolor{black}{\footnotesize E}};
\node[fill,circle, inner sep=2pt] at (-1, 2.9)(sF){};
\node[anchor=west] (Alet) at (sF) {\textcolor{black}{\footnotesize F}};

\end{scope}

\begin{scope}[xshift=8.5cm]

\color{white}

%terminal nodes
\node[fill,circle, inner sep=2pt] at (-3, 1)(sA){};
\node[anchor=north] (Alet) at (sA) {\textcolor{white}{\footnotesize A}};
\node[fill,circle, inner sep=2pt] at (0.7, 1)(sB){};
\node[anchor=north] (Alet) at (sB) {\textcolor{white}{\footnotesize B}};
\node[fill,circle, inner sep=2pt] at (2.7, 1)(sC){};
\node[anchor=north] (Alet) at (sC) {\textcolor{white}{\footnotesize C}};
\node[fill,circle, inner sep=2pt] at (3.9, 1)(sD){};
\node[anchor=north] (Alet) at (sD) {\textcolor{white}{\footnotesize D}};
\node[fill,circle, inner sep=2pt] at (-1, 2.9)(sF){};
\node[anchor=west] (Alet) at (sF) {\textcolor{white}{\footnotesize F}};
\node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
\node[anchor=west] (Alet) at (sE) {\textcolor{white}{\footnotesize E}};

%branches
\draw (0, 7.5) -- (0,7) -- (-2, 7) --(-2, 5.5) --(-3, 5.5)  -- (sA);
\draw (-2, 5.5) -- (-1, 5.5) -- (sF); 
\draw (0,7) -- (2,7) -- (2,5.8) -- (0.7, 5.8) -- (0.7, 4.5) -- (sB); 
\draw (2, 5.8) -- (3.3, 5.8) -- (3.3, 4) -- (3.3, 2.1) -- (2.7, 2.1) -- (sC); 
\draw (3.3, 2.1) -- (3.9, 2.1) -- (sD);

\end{scope}

\end{tikzpicture}

### 2015-10-30 pics/Height_SA/combined.r

@@ -1,0 +1,228 @@
+#replace from here
+
+meantreeHeight <- c(2.720451,2.868717,3.576647,2.432841,1.391817,2.752808,2.78589,0.914045,3.748587,3.147452,2.780463,3.582103,2.687788,4.118663,3.814223,3.448147,1.630521,0.368796,1.976184,2.896259,4.17942,2.159888,3.096881,3.23024,2.325281,3.366376,4.23116,2.89524,3.563949,2.678202,2.07655,3.970413,3.972019,4.391398,2.47988,3.025562,2.589085,4.03658,3.062896,3.286309,2.924163,3.06016,3.573206,4.410655,3.903991,2.381922,3.211783,2.312159,2.923029,2.658005,2.482019,2.951835,3.471832,3.174322,3.312778,3.550283,3.476999,2.942892,3.210365,2.570121,3.579743,3.783381,2.456991,3.800511,2.373685,2.87861,3.909957,1.485166,3.17623,3.938906,3.742874,3.018804,3.451836,1.632896,2.207927,4.071134,3.927119,4.012017,3.271883,4.132294,3.064027,2.456588,3.569598,2.999907,2.919383,2.918403,3.286708,3.547361,2.18788,3.715435,3.512911,2.834833)
[…]

### 2015-10-30 pics/SAvsNoSA/combined.r

@@ -1,0 +1,178 @@
+meanorigin <- c(3.097556,2.891752,3.036793,3.054854,3.332335,3.372771,3.104252,3.108089,3.177228,3.286148,3.18811,3.129275,2.883658,3.172307,3.231096,2.09729,3.449032,3.237759,2.759723,3.130432,3.095301,3.300194,3.174906,3.457544,3.005747,3.262978,2.614712,3.092993,3.313446,3.116054,2.57023,3.455167,3.511363,3.178809,3.144754,2.83437,3.109395,3.685789,3.01077,3.024385,3.235081,2.943418,2.740384,2.839844,3.193746,2.77599,3.056693,3.033801,2.760648,2.363517,2.735236,3.115353,2.99748,3.343816,2.860882,2.940676,3.54013,3.310451,3.199327,2.910229,3.07609,3.579939,3.157838,3.150537,3.006207,3.227735,3.058106,3.031693,3.034131,3.127346,2.874999,3.268004,3.13627,3.145229,2.944998,3.026176,2.735312,2.596134,3.172846,3.200926,3.314302,2.579966,3.041626,2.912639,2.75602,3.347383,3.58038,2.448919,2.934101,3.002292,3.202442,3.210588,2.097001,3.337739,2.358294,3.38713,4.112744,3.331386,2.29925,2.906776)
[…]

### 2015-10-30 pics/TurnoverHPD/combined.r

@@ -1,0 +1,62 @@
+tipsFos <- c(110,22,35,30,19,71,32,13,14,54,139,123,18,104,93,4,112,43,9,38,163,45,31,127,41,14,23,9,56,69,86,6,22,23,22,17,10,27,61,112,56,53,132,74,12,8,27,5,108,39,21,96,12,42,18,6,21,45,95,118,91,28,69,122,117,82,12,36,51,33,35,44,22,119,25,31,80,63,14,5,45,155,28,15,75,5,26,11,35,49,93,146,23,155,30,5,80)
+meanturnoverFos1 <- c(0.237753,0.282807,0.406237,0.325908,0.318481,0.482941,0.441836,0.388898,0.316985,0.411895,0.339383,0.282116,0.354442,0.302688,0.298226,0.387423,0.162655,0.250391,0.311645,0.400448,0.322474,0.305538,0.283538,0.264003,0.491897,0.317938,0.401526,0.387903,0.450142,0.498687,0.361352,0.474671,0.381083,0.562027,0.258626,0.516092,0.516234,0.255333,0.383665,0.407373,0.237667,0.408168,0.190386,0.343227,0.36504,0.635617,0.446687,0.510821,0.373153,0.390367,0.5032,0.287836,0.341761,0.349008,0.256275,0.495444,0.234588,0.449309,0.266914,0.32554,0.43248,0.463163,0.287841,0.265211,0.235868,0.308995,0.499442,0.356889,0.277843,0.26912,0.469959,0.286075,0.450844,0.403487,0.299836,0.245508,0.235599,0.326164,0.422383,0.456544,0.443678,0.308638,0.478939,0.347856,0.406673,0.270111,0.430847,0.456058,0.580468,0.356516,0.168834,0.318948,0.362852,0.34897,0.347366,0.262352,0.255882)
[…]

### 2015-10-30 geoscale_plot.R
hominin geoscaled

@@ -1,0 +1,85 @@
+setwd("~/Git/fossilDating")
+
+library("phytools")
+library("phyloch")
+library("strap")
+library("coda")
+
+## user epochs
+Name <- c("Ps","Po","Miocene","Oligocene","Eocene","Paleocene")
+Start <- c(2.58,5.333,23.03,33.9,56.0,66)
+End <- c(0.0117,2.58,5.333,23.03,33.9,56.0)
+user_ep <- data.frame(Start, End, Name)
+
+t <- read.beast("hominin1r_cog/1r_hominin_ranges_cog-1440423074150-mcc.tree")
+t$root.time <- t$height[1]
+#t$posterior[50] <- 0.0958
+#t$height[50] <- 37.5857
+#t$"height_95%_HPD_MIN"[50] <- 34.2724
+#t$"height_95%_HPD_MAX"[50] <- 40.0015
+#t$posterior[21] <- 0.154
+#t$height[21] <- 15.6909
+#t$"height_95%_HPD_MIN"[21] <- 12.824
+#t$"height_95%_HPD_MAX"[21] <- 18.6343
+
+#t$posterior[22] <- 0.481
+#t$height[22] <- 17.6519
+#t$"height_95%_HPD_MIN"[22] <- 16.0011
+#t$"height_95%_HPD_MAX"[22] <- 19.7024
+
+tre <- read.nexus("hominin1r_cog/1r_hominin_ranges_cog-1440423074150-mcc.tree")
+
+num_taxa <- length(t$tip.label)
+
+display_all_node_bars <- TRUE
+
+names_list <-vector()
+for (name in t$tip){
+  v <- strsplit(name,"_")[[1]]
+  if(display_all_node_bars){
+    names_list = c(names_list,name)
+  }
+  else if(v[length(v)]=="0"){
+    names_list = c(names_list,name)
+  }
+}
+
+nids <- vector()
+pos <- 1
+len_nl <- length(names_list)
+for(n in names_list){
+  for(nn in names_list[pos:len_nl]){
+    if(n != nn){
+      m <- getMRCA(t,c(n,nn))
+      if(m %in% nids == FALSE){
+        nids <- c(nids,m)
+      }
+    }
+  }
+  pos<-pos+1
+}
+
+root_max <- t$"height_95%_HPD_MAX"[1]
[…]

### 2015-11-02 totalEvidenceTalk/pics/SATree2.tex
reorganize to separate talk and paper

\usetikzlibrary{shapes,snakes}

\begin{tikzpicture}[thick,scale=0.45]

\begin{scope}

\draw[ -> ,color=gray] (5, 1) -- (5, 9.5);
\draw[thin,dashed,color=gray] (5, 8.5) -- (0, 8.5);

\node[anchor=west]  at (5, 8.5) (or) {\textcolor{gray}{\footnotesize $t_{or}$}};

\draw[thin,dashed, color=gray] (-4, 1) -- (5, 1);
\node[anchor=west]  at (5, 1) (zero) {\textcolor{gray}{\footnotesize $0$}};

%sampled nodes
\node[fill,circle, inner sep=2pt] at (-3, 3.9)(s2){};
\node[fill,circle, inner sep=2pt] at (0.7, 4.5)(s1){};
\node[fill,circle, inner sep=2pt] at (0.7, 1)(s3){};
\node[fill,circle, inner sep=2pt] at (-2, 6.2)(s4){};
\node[fill,circle, inner sep=2pt] at (4, 3.1)(s5){};
\node[fill,circle, inner sep=2pt] at (-1, 1)(s8){};
\node[fill,circle, inner sep=2pt] at (-1, 2.5)(s9){};
\node[fill,circle, inner sep=2pt] at (2.6, 1)(s10){};

%branches
\draw (0, 8.5) -- (0,7) -- (-2, 7) --(-2, 5.5) --(-3, 5.5) -- (-3, 4.9) -- (s2);
\draw (-2, 5.5) -- (-1, 5.5) -- (s8); 
\draw (0,7) -- (2,7) -- (2,5.8) -- (0.7, 5.8) -- (0.7, 4.5) -- (s3); 
\draw (2, 5.8) -- (3.3, 5.8) -- (3.3, 4) -- (2.6, 4) --  (s10); 
\draw (3.3, 4) -- (4, 4) -- (s5);

\end{scope}

\end{tikzpicture}

### 2015-11-02 totalEvidenceTalk/pics/SampledTree2.tex
reorganize to separate talk and paper

--- 2015-10-30 pics/SampledTree1.tex
@@
 
-\pause
-
 %sampled nodes
@@
 \node[anchor=north] (Alet) at (sD) {\textcolor{black}{\footnotesize D}};
-
-\pause
-
 \node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
@@
 \node[anchor=west] (Alet) at (sF) {\textcolor{black}{\footnotesize F}};
-
-\pause
 
@@
 \node[anchor=north] (Alet) at (sD) {\textcolor{black}{\footnotesize D}};
-
 \node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
@@
 
-\color{white}
+\color{newblue1}
 
@@
 \node[fill,circle, inner sep=2pt] at (-3, 1)(sA){};
-\node[anchor=north] (Alet) at (sA) {\textcolor{white}{\footnotesize A}};
+\node[anchor=north] (Alet) at (sA) {\textcolor{black}{\footnotesize A}};
 \node[fill,circle, inner sep=2pt] at (0.7, 1)(sB){};
-\node[anchor=north] (Alet) at (sB) {\textcolor{white}{\footnotesize B}};
+\node[anchor=north] (Alet) at (sB) {\textcolor{black}{\footnotesize B}};
 \node[fill,circle, inner sep=2pt] at (2.7, 1)(sC){};
-\node[anchor=north] (Alet) at (sC) {\textcolor{white}{\footnotesize C}};
+\node[anchor=north] (Alet) at (sC) {\textcolor{black}{\footnotesize C}};
 \node[fill,circle, inner sep=2pt] at (3.9, 1)(sD){};
-\node[anchor=north] (Alet) at (sD) {\textcolor{white}{\footnotesize D}};
+\node[anchor=north] (Alet) at (sD) {\textcolor{black}{\footnotesize D}};
 \node[fill,circle, inner sep=2pt] at (-1, 2.9)(sF){};
-\node[anchor=west] (Alet) at (sF) {\textcolor{white}{\footnotesize F}};
+\node[anchor=west] (Alet) at (sF) {\textcolor{black}{\footnotesize F}};
 \node[fill,circle, inner sep=2pt] at (3.3, 5)(sE){};
-\node[anchor=west] (Alet) at (sE) {\textcolor{white}{\footnotesize E}};
+\node[anchor=west] (Alet) at (sE) {\textcolor{black}{\footnotesize E}};
 
 %branches
+
 \draw (0, 7.5) -- (0,7) -- (-2, 7) --(-2, 5.5) --(-3, 5.5)  -- (sA);
@@
 
+\node at (0.6, -1) {\textcolor{newblue1}{Sampled tree}};
+
 \end{scope}

### 2015-11-02 totalEvidenceTalk/Section_Canids.tex
broke talk into sections

--- 2015-10-29 TotalEvidenceTalk.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
-%handout, if no \pause
-
-\usepackage{tabularx}
-\usepackage[]{algorithm2e}
-\usepackage{listings}
-\usepackage{lstautogobble}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{subfigure}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
-
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
-
-%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
-%\usepackage{pgfpages}
-%\pgfpagesuselayout{resize to}[a4paper, landscape]
-
-\definecolor{headerColour}{RGB}{180, 230, 245}
-\definecolor{headerTitleColour}{RGB}{0, 60, 80}
-\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
-\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
-\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}
-
-\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
-\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
-\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
-\setbeamercolor{title}{fg=headerTitleColour, bg=white}
-\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}
-
-\usetheme{Goettingen}
-\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother
-
-%% to suppress subsections in sidebar:
-%\setbeamertemplate{subsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-%\setbeamertemplate{subsubsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-
-\usepackage{amsmath, amssymb}
-\usepackage{fancyvrb}
-
-% Font modification
-\usefonttheme{professionalfonts}
-\usepackage{cmbright}
-\usepackage{eulervm}
-\newfont{\Ss}{cmcsc12 scaled 1600}
-
-% no navigation symbols at the bottom of frames
-\beamertemplatenavigationsymbolsempty
-\setbeamertemplate{footline}[frame number]
-
-\usepackage{multirow} % allows entries across multipe rows in tables
-\usepackage{booktabs} % makes fancier rulers in tables
-%\usepackage{setspace} % spacing between lines
-
-%%%%%%%%%%%%%%%%%%
-%% Bibliography related stuff:
-% control space between lines
-  \let\oldthebibliography=\thebibliography
-  \let\endoldthebibliography=\endthebibliography
-  \renewenvironment{thebibliography}[1]{
-    \begin{oldthebibliography}{#1}
-      \setlength{\parskip}{-0.5ex}
-      \setlength{\itemsep}{-0.5ex}
-  }{ \end{oldthebibliography} }
-% Force entry to be in one line
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-% Set bullet point shape:
-\setbeamertemplate{bibliography item}{-}
-%%%%%%%%%%%%%%
-
-\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
-\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
-\newcommand{\head}[1]
-{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
-{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
-\parskip0.3ex
-
-\newcommand{\cT}{{\mathcal T}}
-
-%\defbeamertemplate*{title page}{customized}[1][]
-%{
-%  \titlepage
-%  \usebeamerfont{title}\inserttitle\par
-%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
-%  \bigskip
-%  \usebeamerfont{author}\insertauthor\par
-%  \usebeamerfont{institute}\insertinstitute\par
-%  \usebeamerfont{date}\insertdate\par
-%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
-%}
-
-\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
-\date{The Royal Society, London \\ 9th November 2015}
-\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
-%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
-%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
-%}
-
-%%%%%%%%%%%%%% Figure caption setup
-\usepackage[compatibility=false]{caption}
-\captionsetup[figure]{labelsep=space,justification=centering}
-\renewcommand{\figurename}{\scriptsize Figure adapted from}
-
-% syntax: \figureCaption{Citation caption}{Second (real) caption}
-% second caption can be skipped, first one has to be suppressed if you want to skip it
-\newcommand\figureCaption[2]{%
-  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
-  \caption{\scriptsize #1}
-  \caption*{#2}
-}
-
-%%%%%%%%%%%%%% Fixme and comment setup¤
-\definecolor{red}{HTML}{C92D39}
-\definecolor{green}{HTML}{498A44}
-
-% syntax: \fixme{What needs to be fixed}
-\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
-% Hide all fixmes by switching the line above to this:
-%\newcommand{\fixme}[1]{}
-
-% syntax: \comment{Name of commenter}{C omment}
-\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
-% Hide all comments by switching the line above to this:
-%\newcommand{\comment}[2]{}
-
-\begin{document}
-
-%Slide1
-\frame{\titlepage}
-
-\section{Fossilized birth-death process}
-
-\begin{frame}
-\frametitle{The fossilized birth-death process}
-
-\end{frame}
-
-\section{Penguin dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Canids}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
-\label{saPostPr}
+\includegraphics[width=0.70\textwidth]{../canids/1c_phyloAgeVsGeoAge.pdf}
 \end{figure}
-\small{Probabilities of each fossil to be a sampled ancestor across models.
+{\small Bayesian phylogenetic ages for each of 123 canid fossils plotted against palaeontological ages, under \Mstrict{}. 
+Bayesian estimates are represented by the median and the 99\% HPD interval of posterior. Blue line shows the $x=y$. 
+16 fossils with inconsistent estimates are labelled.}
+\end{frame}
 
-{\it P} stands for the partitioned model, {\it G} for gamma variation across sites, {\it Mkv} for conditioning on 
-variable characters, {\it R} for relaxed clock model, {\it dns} for $d$, $\nu$, and $s$ tree prior parameterisation, {\it lmp} for $
-\lambda$, $\mu$, and $\psi$ tree prior parameterisation.} 
-\end{frame}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
-\label{postPredictive}
+\includegraphics[width=0.5\textwidth]{../canids/1_canids-1440516016976-tree5001.pdf}
 \end{figure}
-\small{The posterior and posterior predictive distributions for (left) tree length, $T$, and genealogical $D_F$ statistics and (right) $B_1$ tree imbalance statistic and Colless's tree imbalance index, $I_c$, for model 8.
-
-The posterior trees are on the unbalanced end of the predictive distribution.}
+A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.
 \end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
-\label{penguinTree}
-\end{figure}
-\footnotesize{A maximum sampled ancestor clade credibility tree for the total-evidence analysis. % with mean node ages. 
-The numbers in blue show the posterior supports of clades. The filled red and black circles represent sampled 
-ancestors. Fossils with positive evidence of being sampled ancestors are shown in red.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
-\label{fig:BF}
-\end{center}
-\end{figure}
-\small{The evidence fossil sampled ancestors. The samples above the shaded area have positive evidence to be sampled ancestors and below the shaded area have positive evidence to be terminal nodes.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
-\label{fig:divDates}
-\end{figure}
-\small{The posterior summary tree of extant penguins with 95\% HPD intervals and
-posterior supports of clades after removing fossils.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
-\label{fig:rootAges}
-\end{center}
-\end{figure}
-\small{The age of root of extant penguins across models including total-evidence analysis under relaxed (8+DNA R) and strict (8+DND S) molecular clocks and 
-total-evidence analysis with crown fossils only.}
-\end{frame}
-
-\begin{frame}
-\frametitle{Dating the evolutionary history of penguins}
-\begin{figure}
-\includegraphics[width=\textwidth]{penguinTree.pdf}
-\end{figure}
-\end{frame}
-
-\section{Fossil dating}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.6\textwidth]{priors.pdf}
-\label{Fig:Prior}
-\end{figure}
-
-Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure1.pdf}
-\end{figure}
-Bayesian phylogenetic age estimates for each of 36 penguin fossils plotted against their palaeontological age estimates, under two alternative evolutionary models. 
-The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure2.pdf}
-\end{figure}
-A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
-the precision is 1/variance in the marginal posterior distribution of the age).
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of age \& BF estimates between models}
-\begin{figure}%
-\centering
-\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
-\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
-\label{fig:compareM1M8ab}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of probability \& error between models}
-\begin{figure}%
-\centering
-\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
-\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
-\label{fig:compareM1M8cd}
-\end{figure}
-\end{frame}
-
-\end{document}

### 2015-11-02 totalEvidenceTalk/Section_FBD.tex
broke talk into sections

--- 2015-10-29 TotalEvidenceTalk.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
-%handout, if no \pause
-
-\usepackage{tabularx}
-\usepackage{tikz}
-\usepackage[]{algorithm2e}
-\usepackage{listings}
-\usepackage{lstautogobble}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{subfigure}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
-
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
-% turnover = birth / death
-\newcommand{\turnover}{\nu}
-
-% sampling proportion = \psi / \psi + \mu
-\newcommand{\fosp}{s}
-\newcommand{\sampleprop}{sampling proportion}
-
-\definecolor{newblue}{rgb}{0.3, 0.5, 0.9}
-\definecolor{newblue1}{rgb}{0.3, 0.3, 0.7}
-\definecolor{newred}{RGB}{200,0,0}
-
-%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
-%\usepackage{pgfpages}
-%\pgfpagesuselayout{resize to}[a4paper, landscape]
-
-\definecolor{headerColour}{RGB}{180, 230, 245}
-\definecolor{headerTitleColour}{RGB}{0, 60, 80}
-\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
-\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
-\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}
-
-\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
-\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
-\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
-\setbeamercolor{title}{fg=headerTitleColour, bg=white}
-\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}
-
-\usetheme{Goettingen}
-\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother
-
-%% to suppress subsections in sidebar:
-%\setbeamertemplate{subsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-%\setbeamertemplate{subsubsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-
-\usepackage{amsmath, amssymb}
-\usepackage{fancyvrb}
-
-% Font modification
-\usefonttheme{professionalfonts}
-\usepackage{cmbright}
-\usepackage{eulervm}
-\newfont{\Ss}{cmcsc12 scaled 1600}
-
-% no navigation symbols at the bottom of frames
-\beamertemplatenavigationsymbolsempty
-\setbeamertemplate{footline}[frame number]
-
-\usepackage{multirow} % allows entries across multipe rows in tables
-\usepackage{booktabs} % makes fancier rulers in tables
-%\usepackage{setspace} % spacing between lines
-
-%%%%%%%%%%%%%%%%%%
-%% Bibliography related stuff:
-% control space between lines
-  \let\oldthebibliography=\thebibliography
-  \let\endoldthebibliography=\endthebibliography
-  \renewenvironment{thebibliography}[1]{
-    \begin{oldthebibliography}{#1}
-      \setlength{\parskip}{-0.5ex}
-      \setlength{\itemsep}{-0.5ex}
-  }{ \end{oldthebibliography} }
-% Force entry to be in one line
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-% Set bullet point shape:
-\setbeamertemplate{bibliography item}{-}
-%%%%%%%%%%%%%%
-
-\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
-\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
-\newcommand{\head}[1]
-{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
-{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
-\parskip0.3ex
-
-\newcommand{\cT}{{\mathcal T}}
-
-%\defbeamertemplate*{title page}{customized}[1][]
-%{
-%  \titlepage
-%  \usebeamerfont{title}\inserttitle\par
-%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
-%  \bigskip
-%  \usebeamerfont{author}\insertauthor\par
-%  \usebeamerfont{institute}\insertinstitute\par
-%  \usebeamerfont{date}\insertdate\par
-%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
-%}
-
-\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
-\date{The Royal Society, London \\ 9th November 2015}
-\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
-%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
-%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
-%}
-
-%%%%%%%%%%%%%% Figure caption setup
-\usepackage[compatibility=false]{caption}
-\captionsetup[figure]{labelsep=space,justification=centering}
-\renewcommand{\figurename}{\scriptsize Figure adapted from}
-
-% syntax: \figureCaption{Citation caption}{Second (real) caption}
-% second caption can be skipped, first one has to be suppressed if you want to skip it
-\newcommand\figureCaption[2]{%
-  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
-  \caption{\scriptsize #1}
-  \caption*{#2}
-}
-
-%%%%%%%%%%%%%% Fixme and comment setup¤
-\definecolor{red}{HTML}{C92D39}
-\definecolor{green}{HTML}{498A44}
-
-% syntax: \fixme{What needs to be fixed}
-\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
-% Hide all fixmes by switching the line above to this:
-%\newcommand{\fixme}[1]{}
-
-% syntax: \comment{Name of commenter}{C omment}
-\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
-% Hide all comments by switching the line above to this:
-%\newcommand{\comment}[2]{}
-
-\begin{document}
-
-%Slide1
-\frame{\titlepage}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Fossilized birth-death process}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}{Analysis of penguin morphological data}
-
-Penguin dataset (Ksepka et al 2012) consisting of morphological data of:
-\begin{itemize}
-\item all extant penguins (19 species)
-\item 36 fossil species assigned to stratigraphic intervals
-\end{itemize}
-
-\vskip2mm
-
-Models:
-
-\begin{itemize}
-\item Lewis MK vs MKv (Lewis 2001)
-\item Partitions vs single alignment
-\item Rate variation across sites and across partitions
-\item Tree prior: FBD and Skyline FBD
-\item Different tree model parameterisations
-\end{itemize}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}{Parameterisations}
-
-\renewcommand{\arraystretch}{1.2}
-
-lmp:
-
-\color{newblue1}
-
-\begin{equation*} 
-\begin{tabular}{ll} 
-\text{birth rate} & $\lambda$  \\ 
-\text{death rate} & $\mu$ \\ 
-\text{sampling rate} & $\psi$  \\ 
-\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
-\end{tabular} \end{equation*}
-
-\textcolor{black}{dvs:}
-
-\begin{equation*} 
-\begin{tabular}{ll} 
-\text{net diversification rate} & $d = \lambda-\mu$  \\ 
-\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
-\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
-\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
-\end{tabular} \end{equation*}
-
-\end{frame}
-
-\section{Penguin dating}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
-\label{saPostPr}
-\end{figure}
-\small{Probabilities of each fossil to be a sampled ancestor across models.
-
-{\it P} stands for the partitioned model, {\it G} for gamma variation across sites, {\it Mkv} for conditioning on 
-variable characters, {\it R} for relaxed clock model, {\it dns} for $d$, $\nu$, and $s$ tree prior parameterisation, {\it lmp} for $
-\lambda$, $\mu$, and $\psi$ tree prior parameterisation.} 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
-\label{postPredictive}
-\end{figure}
-\small{The posterior and posterior predictive distributions for (left) tree length, $T$, and genealogical $D_F$ statistics and (right) $B_1$ tree imbalance statistic and Colless's tree imbalance index, $I_c$, for model 8.
-
-The posterior trees are on the unbalanced end of the predictive distribution.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
-\label{penguinTree}
-\end{figure}
-\footnotesize{A maximum sampled ancestor clade credibility tree for the total-evidence analysis. % with mean node ages. 
-The numbers in blue show the posterior supports of clades. The filled red and black circles represent sampled 
-ancestors. Fossils with positive evidence of being sampled ancestors are shown in red.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
-\label{fig:BF}
-\end{center}
-\end{figure}
-\small{The evidence fossil sampled ancestors. The samples above the shaded area have positive evidence to be sampled ancestors and below the shaded area have positive evidence to be terminal nodes.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
-\label{fig:divDates}
-\end{figure}
-\small{The posterior summary tree of extant penguins with 95\% HPD intervals and
-posterior supports of clades after removing fossils.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
-\label{fig:rootAges}
-\end{center}
-\end{figure}
-\small{The age of root of extant penguins across models including total-evidence analysis under relaxed (8+DNA R) and strict (8+DND S) molecular clocks and 
-total-evidence analysis with crown fossils only.}
-\end{frame}
-
-\begin{frame}
-\frametitle{Dating the evolutionary history of penguins}
-\begin{figure}
-\includegraphics[width=\textwidth]{penguinTree.pdf}
-\end{figure}
-\end{frame}
-
-\section{Fossil dating}
-
-\begin{frame}{But what about those hyper parameter priors?}
-\begin{figure}
-\includegraphics[width=0.6\textwidth]{priors.pdf}
-\label{Fig:Prior}
-\end{figure}
-
-\small{Prior density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.}
-\end{frame}
-
-\begin{frame}{And what about this guy?}
-\includegraphics[width=0.5\textwidth]{Homo_naledi.jpg}
-\includegraphics[width=0.5\textwidth]{Atlantic_Homo_naledi.jpg}
-
-\bigskip{}
-
-{\bf Can we date fossil remains using Bayesian phylogenetics? }
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure1.pdf}
-\end{figure}
-Bayesian phylogenetic age estimates for each of 36 penguin fossils plotted against their palaeontological age estimates, under two alternative evolutionary models. 
-The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure2.pdf}
-\end{figure}
-A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
-the precision is 1/variance in the marginal posterior distribution of the age).
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of age \& BF estimates between models}
-\begin{figure}%
-\centering
-\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
-\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
-\label{fig:compareM1M8ab}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of probability \& error between models}
-\begin{figure}%
-\centering
-\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
-\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
-\label{fig:compareM1M8cd}
-\end{figure}
-\end{frame}
-
-\end{document}

### 2015-11-02 totalEvidenceTalk/Section_FossilDating.tex
broke talk into sections

--- 2015-10-29 TotalEvidenceTalk.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
-%handout, if no \pause
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fossil dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage{tabularx}
-\usepackage[]{algorithm2e}
-\usepackage{listings}
-\usepackage{lstautogobble}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{subfigure}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
+\begin{frame}{And what about this guy?}
+\includegraphics[width=0.5\textwidth]{Homo_naledi.jpg}
+\includegraphics[width=0.5\textwidth]{Atlantic_Homo_naledi.jpg}
 
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
-
-%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
-%\usepackage{pgfpages}
-%\pgfpagesuselayout{resize to}[a4paper, landscape]
-
-\definecolor{headerColour}{RGB}{180, 230, 245}
-\definecolor{headerTitleColour}{RGB}{0, 60, 80}
-\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
-\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
-\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}
-
-\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
-\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
-\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
-\setbeamercolor{title}{fg=headerTitleColour, bg=white}
-\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}
-
-\usetheme{Goettingen}
-\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother
-
-%% to suppress subsections in sidebar:
-%\setbeamertemplate{subsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-%\setbeamertemplate{subsubsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-
-\usepackage{amsmath, amssymb}
-\usepackage{fancyvrb}
-
-% Font modification
-\usefonttheme{professionalfonts}
-\usepackage{cmbright}
-\usepackage{eulervm}
-\newfont{\Ss}{cmcsc12 scaled 1600}
-
-% no navigation symbols at the bottom of frames
-\beamertemplatenavigationsymbolsempty
-\setbeamertemplate{footline}[frame number]
-
-\usepackage{multirow} % allows entries across multipe rows in tables
-\usepackage{booktabs} % makes fancier rulers in tables
-%\usepackage{setspace} % spacing between lines
-
-%%%%%%%%%%%%%%%%%%
-%% Bibliography related stuff:
-% control space between lines
-  \let\oldthebibliography=\thebibliography
-  \let\endoldthebibliography=\endthebibliography
-  \renewenvironment{thebibliography}[1]{
-    \begin{oldthebibliography}{#1}
-      \setlength{\parskip}{-0.5ex}
-      \setlength{\itemsep}{-0.5ex}
-  }{ \end{oldthebibliography} }
-% Force entry to be in one line
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-% Set bullet point shape:
-\setbeamertemplate{bibliography item}{-}
-%%%%%%%%%%%%%%
-
-\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
-\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
-\newcommand{\head}[1]
-{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
-{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
-\parskip0.3ex
-
-\newcommand{\cT}{{\mathcal T}}
-
-%\defbeamertemplate*{title page}{customized}[1][]
-%{
-%  \titlepage
-%  \usebeamerfont{title}\inserttitle\par
-%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
-%  \bigskip
-%  \usebeamerfont{author}\insertauthor\par
-%  \usebeamerfont{institute}\insertinstitute\par
-%  \usebeamerfont{date}\insertdate\par
-%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
-%}
-
-\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
-\date{The Royal Society, London \\ 9th November 2015}
-\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
-%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
-%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
-%}
-
-%%%%%%%%%%%%%% Figure caption setup
-\usepackage[compatibility=false]{caption}
-\captionsetup[figure]{labelsep=space,justification=centering}
-\renewcommand{\figurename}{\scriptsize Figure adapted from}
-
-% syntax: \figureCaption{Citation caption}{Second (real) caption}
-% second caption can be skipped, first one has to be suppressed if you want to skip it
-\newcommand\figureCaption[2]{%
-  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
-  \caption{\scriptsize #1}
-  \caption*{#2}
-}
-
-%%%%%%%%%%%%%% Fixme and comment setup¤
-\definecolor{red}{HTML}{C92D39}
-\definecolor{green}{HTML}{498A44}
-
-% syntax: \fixme{What needs to be fixed}
-\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
-% Hide all fixmes by switching the line above to this:
-%\newcommand{\fixme}[1]{}
-
-% syntax: \comment{Name of commenter}{C omment}
-\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
-% Hide all comments by switching the line above to this:
-%\newcommand{\comment}[2]{}
-
-\begin{document}
-
-%Slide1
-\frame{\titlepage}
-
-\section{Fossilized birth-death process}
-
-\begin{frame}
-\frametitle{The fossilized birth-death process}
+\bigskip{}
 
@@
 
-\section{Penguin dating}
+\begin{frame}{Can we date fossils using Bayesian phylogenetics? }
 
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
-\label{saPostPr}
-\end{figure}
-\small{Probabilities of each fossil to be a sampled ancestor across models.
+Given:
+\begin{itemize}
+\item A reference set of fossils of known age, and
+\item A morphological characterization for reference fossils, and
+\item The fossilized birth-death model of sampled ancestor trees
+\item A ``clock-like'' model of morphological evolution 
+\item A morphological characterization of a related focal fossil of unknown age 
+\end{itemize}
 
-{\it P} stands for the partitioned model, {\it G} for gamma variation across sites, {\it Mkv} for conditioning on 
-variable characters, {\it R} for relaxed clock model, {\it dns} for $d$, $\nu$, and $s$ tree prior parameterisation, {\it lmp} for $
-\lambda$, $\mu$, and $\psi$ tree prior parameterisation.} 
-\end{frame}
+{\bf Can the age of the focal fossil be estimated using Bayesian phylogenetic inference?}
 
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
-\label{postPredictive}
-\end{figure}
-\small{The posterior and posterior predictive distributions for (left) tree length, $T$, and genealogical $D_F$ statistics and (right) $B_1$ tree imbalance statistic and Colless's tree imbalance index, $I_c$, for model 8.
+\medskip{}
 
-The posterior trees are on the unbalanced end of the predictive distribution.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
-\label{penguinTree}
-\end{figure}
-\footnotesize{A maximum sampled ancestor clade credibility tree for the total-evidence analysis. % with mean node ages. 
-The numbers in blue show the posterior supports of clades. The filled red and black circles represent sampled 
-ancestors. Fossils with positive evidence of being sampled ancestors are shown in red.}
+We tested this hypothesis by using the penguin data set.
+For each of the 36 fossils in turn, we discarded the age information of the focal fossil to mimic the scenario where the age was unknown, and sought to estimate its age using the reference 35 fossils as the reference set.
 
@@
 
-\begin{frame}
+\begin{frame}{Estimating the phylogenetic age of a fossil}
 \begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
-\label{fig:BF}
-\end{center}
-\end{figure}
-\small{The evidence fossil sampled ancestors. The samples above the shaded area have positive evidence to be sampled ancestors and below the shaded area have positive evidence to be terminal nodes.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
-\label{fig:divDates}
-\end{figure}
-\small{The posterior summary tree of extant penguins with 95\% HPD intervals and
-posterior supports of clades after removing fossils.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
-\label{fig:rootAges}
-\end{center}
-\end{figure}
-\small{The age of root of extant penguins across models including total-evidence analysis under relaxed (8+DNA R) and strict (8+DND S) molecular clocks and 
-total-evidence analysis with crown fossils only.}
-\end{frame}
-
-\begin{frame}
-\frametitle{Dating the evolutionary history of penguins}
-\begin{figure}
-\includegraphics[width=\textwidth]{penguinTree.pdf}
-\end{figure}
-\end{frame}
-
-\section{Fossil dating}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.6\textwidth]{priors.pdf}
-\label{Fig:Prior}
-\end{figure}
-
-Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure1.pdf}
+\includegraphics[width=\textwidth]{../Figure1.pdf}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
+\includegraphics[width=\textwidth]{../8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
+\includegraphics[width=\textwidth]{../8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{Figure2.pdf}
+\includegraphics[width=\textwidth]{../Figure2.pdf}
 \end{figure}
@@
 \centering
-\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
-\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
+\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{../compareAgeM1M8.pdf}}\qquad
+\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{../compareBFM1M8.pdf}}\\
 \label{fig:compareM1M8ab}
@@
 \centering
-\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
-\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
+\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{../comparePostM1M8.pdf}}\qquad
+\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{../compareErrorM1M8.pdf}}\\
 \label{fig:compareM1M8cd}
@@
 \end{frame}
-
-\end{document}

### 2015-11-02 totalEvidenceTalk/Section_Hominins.tex
broke talk into sections

--- 2015-10-29 TotalEvidenceTalk.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
-%handout, if no \pause
-
-\usepackage{tabularx}
-\usepackage[]{algorithm2e}
-\usepackage{listings}
-\usepackage{lstautogobble}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{subfigure}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
-
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
-
-%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
-%\usepackage{pgfpages}
-%\pgfpagesuselayout{resize to}[a4paper, landscape]
-
-\definecolor{headerColour}{RGB}{180, 230, 245}
-\definecolor{headerTitleColour}{RGB}{0, 60, 80}
-\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
-\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
-\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}
-
-\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
-\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
-\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
-\setbeamercolor{title}{fg=headerTitleColour, bg=white}
-\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}
-
-\usetheme{Goettingen}
-\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother
-
-%% to suppress subsections in sidebar:
-%\setbeamertemplate{subsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-%\setbeamertemplate{subsubsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-
-\usepackage{amsmath, amssymb}
-\usepackage{fancyvrb}
-
-% Font modification
-\usefonttheme{professionalfonts}
-\usepackage{cmbright}
-\usepackage{eulervm}
-\newfont{\Ss}{cmcsc12 scaled 1600}
-
-% no navigation symbols at the bottom of frames
-\beamertemplatenavigationsymbolsempty
-\setbeamertemplate{footline}[frame number]
-
-\usepackage{multirow} % allows entries across multipe rows in tables
-\usepackage{booktabs} % makes fancier rulers in tables
-%\usepackage{setspace} % spacing between lines
-
-%%%%%%%%%%%%%%%%%%
-%% Bibliography related stuff:
-% control space between lines
-  \let\oldthebibliography=\thebibliography
-  \let\endoldthebibliography=\endthebibliography
-  \renewenvironment{thebibliography}[1]{
-    \begin{oldthebibliography}{#1}
-      \setlength{\parskip}{-0.5ex}
-      \setlength{\itemsep}{-0.5ex}
-  }{ \end{oldthebibliography} }
-% Force entry to be in one line
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-% Set bullet point shape:
-\setbeamertemplate{bibliography item}{-}
-%%%%%%%%%%%%%%
-
-\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
-\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
-\newcommand{\head}[1]
-{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
-{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
-\parskip0.3ex
-
-\newcommand{\cT}{{\mathcal T}}
-
-%\defbeamertemplate*{title page}{customized}[1][]
-%{
-%  \titlepage
-%  \usebeamerfont{title}\inserttitle\par
-%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
-%  \bigskip
-%  \usebeamerfont{author}\insertauthor\par
-%  \usebeamerfont{institute}\insertinstitute\par
-%  \usebeamerfont{date}\insertdate\par
-%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
-%}
-
-\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
-\date{The Royal Society, London \\ 9th November 2015}
-\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
-%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
-%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
-%}
-
-%%%%%%%%%%%%%% Figure caption setup
-\usepackage[compatibility=false]{caption}
-\captionsetup[figure]{labelsep=space,justification=centering}
-\renewcommand{\figurename}{\scriptsize Figure adapted from}
-
-% syntax: \figureCaption{Citation caption}{Second (real) caption}
-% second caption can be skipped, first one has to be suppressed if you want to skip it
-\newcommand\figureCaption[2]{%
-  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
-  \caption{\scriptsize #1}
-  \caption*{#2}
-}
-
-%%%%%%%%%%%%%% Fixme and comment setup¤
-\definecolor{red}{HTML}{C92D39}
-\definecolor{green}{HTML}{498A44}
-
-% syntax: \fixme{What needs to be fixed}
-\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
-% Hide all fixmes by switching the line above to this:
-%\newcommand{\fixme}[1]{}
-
-% syntax: \comment{Name of commenter}{C omment}
-\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
-% Hide all comments by switching the line above to this:
-%\newcommand{\comment}[2]{}
-
-\begin{document}
-
-%Slide1
-\frame{\titlepage}
-
-\section{Fossilized birth-death process}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Hominins}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{frame}
-\frametitle{The fossilized birth-death process}
-
+\frametitle{Controversies in human evolution: {\it Ardipithecus ramidus}}
+\begin{figure}
+\includegraphics[width=\textwidth]{ardi_crop.jpg}
+\end{figure}
+`Ardi', 4.4M years old was described as a the oldest fossil relative on the human lineage, a {\it hominin}, and thus more closely related to the human than the chimpanzee. 
 \end{frame}
 
-\section{Penguin dating}
-
-\begin{frame}
+\begin{frame}{Hominin phylogeny inferred using FBD tree prior \& morphological clock.}
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
-\label{saPostPr}
-\end{figure}
-\small{Probabilities of each fossil to be a sampled ancestor across models.
-
-{\it P} stands for the partitioned model, {\it G} for gamma variation across sites, {\it Mkv} for conditioning on 
-variable characters, {\it R} for relaxed clock model, {\it dns} for $d$, $\nu$, and $s$ tree prior parameterisation, {\it lmp} for $
-\lambda$, $\mu$, and $\psi$ tree prior parameterisation.} 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
-\label{postPredictive}
-\end{figure}
-\small{The posterior and posterior predictive distributions for (left) tree length, $T$, and genealogical $D_F$ statistics and (right) $B_1$ tree imbalance statistic and Colless's tree imbalance index, $I_c$, for model 8.
-
-The posterior trees are on the unbalanced end of the predictive distribution.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
-\label{penguinTree}
-\end{figure}
-\footnotesize{A maximum sampled ancestor clade credibility tree for the total-evidence analysis. % with mean node ages. 
-The numbers in blue show the posterior supports of clades. The filled red and black circles represent sampled 
-ancestors. Fossils with positive evidence of being sampled ancestors are shown in red.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
-\label{fig:BF}
-\end{center}
-\end{figure}
-\small{The evidence fossil sampled ancestors. The samples above the shaded area have positive evidence to be sampled ancestors and below the shaded area have positive evidence to be terminal nodes.}
-
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
-\label{fig:divDates}
-\end{figure}
-\small{The posterior summary tree of extant penguins with 95\% HPD intervals and
-posterior supports of clades after removing fossils.}
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
-\label{fig:rootAges}
-\end{center}
-\end{figure}
-\small{The age of root of extant penguins across models including total-evidence analysis under relaxed (8+DNA R) and strict (8+DND S) molecular clocks and 
-total-evidence analysis with crown fossils only.}
-\end{frame}
-
-\begin{frame}
-\frametitle{Dating the evolutionary history of penguins}
-\begin{figure}
-\includegraphics[width=\textwidth]{penguinTree.pdf}
+\includegraphics[width=0.75\textwidth]{homininSummaryTree.pdf}
 \end{figure}
 \end{frame}
-
-\section{Fossil dating}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=0.6\textwidth]{priors.pdf}
-\label{Fig:Prior}
-\end{figure}
-
-Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure1.pdf}
-\end{figure}
-Bayesian phylogenetic age estimates for each of 36 penguin fossils plotted against their palaeontological age estimates, under two alternative evolutionary models. 
-The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure2.pdf}
-\end{figure}
-A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
-the precision is 1/variance in the marginal posterior distribution of the age).
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of age \& BF estimates between models}
-\begin{figure}%
-\centering
-\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
-\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
-\label{fig:compareM1M8ab}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of probability \& error between models}
-\begin{figure}%
-\centering
-\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
-\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
-\label{fig:compareM1M8cd}
-\end{figure}
-\end{frame}
-
-\end{document}

### 2015-11-02 totalEvidenceTalk/Section_PenguinDating.tex
broke talk into sections

--- 2015-10-29 TotalEvidenceTalk.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[xcolor=pdftex,dvipsnames,table,10pt]{beamer}
-%handout, if no \pause
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Penguin dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage{tabularx}
-\usepackage[]{algorithm2e}
-\usepackage{listings}
-\usepackage{lstautogobble}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{xcolor}
-\usepackage{subfigure}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
+\begin{frame}{Analysis of penguin morphological data}
 
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
+Penguin dataset (Ksepka et al 2012) consisting of morphological data of:
+\begin{itemize}
+\item all extant penguins (19 species)
+\item 36 fossil species assigned to stratigraphic intervals
+\end{itemize}
 
-%\documentclass[handout,xcolor=pdftex,dvipsnames,table]{beamer} % USE THIS WITH pdfpages STUFF
-%\usepackage{pgfpages}
-%\pgfpagesuselayout{resize to}[a4paper, landscape]
+\vskip2mm
 
-\definecolor{headerColour}{RGB}{180, 230, 245}
-\definecolor{headerTitleColour}{RGB}{0, 60, 80}
-\definecolor{sectionShadedColour}{RGB}{50, 110, 130}
-\definecolor{subsectionHighlightColour}{RGB}{50, 50, 50}
-\definecolor{subsectionShadedColour}{RGB}{100, 100, 100}
+Models:
 
-\setbeamercolor{subsection in sidebar}{fg=subsectionHighlightColour}
-\setbeamercolor{subsection in sidebar shaded}{fg=subsectionShadedColour}
-\setbeamercolor{structure}{fg=headerTitleColour, bg=headerColour}
-\setbeamercolor{title}{fg=headerTitleColour, bg=white}
-\setbeamercolor{section in sidebar shaded}{fg=sectionShadedColour}
-
-\usetheme{Goettingen}
-\makeatletter\setbeamertemplate{sidebar canvas \beamer@sidebarside}[vertical shading][top=headerColour,bottom=white]\makeatother
-
-%% to suppress subsections in sidebar:
-%\setbeamertemplate{subsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-%\setbeamertemplate{subsubsection in sidebar shaded}
-%{\vspace*{-\baselineskip}}
-
-\usepackage{amsmath, amssymb}
-\usepackage{fancyvrb}
-
-% Font modification
-\usefonttheme{professionalfonts}
-\usepackage{cmbright}
-\usepackage{eulervm}
-\newfont{\Ss}{cmcsc12 scaled 1600}
-
-% no navigation symbols at the bottom of frames
-\beamertemplatenavigationsymbolsempty
-\setbeamertemplate{footline}[frame number]
-
-\usepackage{multirow} % allows entries across multipe rows in tables
-\usepackage{booktabs} % makes fancier rulers in tables
-%\usepackage{setspace} % spacing between lines
-
-%%%%%%%%%%%%%%%%%%
-%% Bibliography related stuff:
-% control space between lines
-  \let\oldthebibliography=\thebibliography
-  \let\endoldthebibliography=\endthebibliography
-  \renewenvironment{thebibliography}[1]{
-    \begin{oldthebibliography}{#1}
-      \setlength{\parskip}{-0.5ex}
-      \setlength{\itemsep}{-0.5ex}
-  }{ \end{oldthebibliography} }
-% Force entry to be in one line
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-% Set bullet point shape:
-\setbeamertemplate{bibliography item}{-}
-%%%%%%%%%%%%%%
-
-\newenvironment{items}{\begin{list}{$\bullet$}{\itemsep0ex plus 0.2ex
-\parsep0ex plus 0.2ex \topsep0ex \parskip0ex}}{\end{list}}
-\newcommand{\head}[1]
-{\slide{\begin{center}\textbf{#1}\vspace*{-0.5\baselineskip}
-{\color{red}\rule{\textwidth}{1mm}}\end{center}}}
-\parskip0.3ex
-
-\newcommand{\cT}{{\mathcal T}}
-
-%\defbeamertemplate*{title page}{customized}[1][]
-%{
-%  \titlepage
-%  \usebeamerfont{title}\inserttitle\par
-%  \usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par
-%  \bigskip
-%  \usebeamerfont{author}\insertauthor\par
-%  \usebeamerfont{institute}\insertinstitute\par
-%  \usebeamerfont{date}\insertdate\par
-%  \usebeamercolor[fg]{titlegraphic}\inserttitlegraphic
-%}
-
-\title[Total-evidence dating]{`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages}
-\author[]{Alexei Drummond, alexei@cs.auckland.ac.nz}
-\date{The Royal Society, London \\ 9th November 2015}
-\institute{Professor of Computational Biology \\ Department of Computer Science \\ University of Auckland}
-%\titlegraphic{\hspace*{8cm}\includegraphics[height=1.5cm]{figures/cEvo_logo_transparent.png}}
-%{\small Link to Script\&Slides:  \url{http://www.tb.ethz.ch/education} 
-%}
-
-%%%%%%%%%%%%%% Figure caption setup
-\usepackage[compatibility=false]{caption}
-\captionsetup[figure]{labelsep=space,justification=centering}
-\renewcommand{\figurename}{\scriptsize Figure adapted from}
-
-% syntax: \figureCaption{Citation caption}{Second (real) caption}
-% second caption can be skipped, first one has to be suppressed if you want to skip it
-\newcommand\figureCaption[2]{%
-  \captionsetup{aboveskip=0.1cm,belowskip=0cm}
-  \caption{\scriptsize #1}
-  \caption*{#2}
-}
-
-%%%%%%%%%%%%%% Fixme and comment setup¤
-\definecolor{red}{HTML}{C92D39}
-\definecolor{green}{HTML}{498A44}
-
-% syntax: \fixme{What needs to be fixed}
-\newcommand{\fixme}[1]{\textcolor{red}{\texttt{{\bf FIX ME:} #1}}}
-% Hide all fixmes by switching the line above to this:
-%\newcommand{\fixme}[1]{}
-
-% syntax: \comment{Name of commenter}{C omment}
-\newcommand{\comment}[2]{\textcolor{green}{{\bf Comment by {#1}}: #2}}
-% Hide all comments by switching the line above to this:
-%\newcommand{\comment}[2]{}
-
-\begin{document}
-
-%Slide1
-\frame{\titlepage}
-
-\section{Fossilized birth-death process}
-
-\begin{frame}
-\frametitle{The fossilized birth-death process}
+\begin{itemize}
+\item Lewis MK vs MKv (Lewis 2001)
+\item Partitions vs single alignment
+\item Rate variation across sites and across partitions
+\item Tree prior: FBD and Skyline FBD
+\item Different tree model parameterisations
+\end{itemize}
 
@@
 
-\section{Penguin dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Parameterisations}
+
+\renewcommand{\arraystretch}{1.2}
+
+lmp:
+
+\color{newblue1}
+
+\begin{equation*} 
+\begin{tabular}{ll} 
+\text{birth rate} & $\lambda$  \\ 
+\text{death rate} & $\mu$ \\ 
+\text{sampling rate} & $\psi$  \\ 
+\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
+\end{tabular} \end{equation*}
+
+\textcolor{black}{dvs:}
+
+\begin{equation*} 
+\begin{tabular}{ll} 
+\text{net diversification rate} & $d = \lambda-\mu$  \\ 
+\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
+\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
+\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
+\end{tabular} \end{equation*}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{../ag_morpho_data/figures/sa.pdf}
+\includegraphics[width=0.7\textwidth]{../../ag_morpho_data/figures/sa.pdf}
 \label{saPostPr}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}
 \begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/pp.pdf}
+\includegraphics[width=\textwidth]{../../ag_morpho_data/figures/pp.pdf}
 \label{postPredictive}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}
 \begin{figure}
-\includegraphics[width=\textwidth]{../ag_morpho_data/figures/tree.pdf}
+\includegraphics[width=\textwidth]{../../ag_morpho_data/figures/tree.pdf}
 \label{penguinTree}
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/BayesFactors.pdf}
+\includegraphics[width=0.75\textwidth]{../../ag_morpho_data/figures/BayesFactors.pdf}
 \label{fig:BF}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{../ag_morpho_data/figures/extant_tree.pdf}
+\includegraphics[width=0.95\textwidth]{../../ag_morpho_data/figures/extant_tree.pdf}
 \label{fig:divDates}
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{../ag_morpho_data/figures/root_ages.pdf}
+\includegraphics[width=0.75\textwidth]{../../ag_morpho_data/figures/root_ages.pdf}
 \label{fig:rootAges}
@@
 
-\section{Fossil dating}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{frame}
+\begin{frame}{But what about those hyper parameter priors?}
 \begin{figure}
-\includegraphics[width=0.6\textwidth]{priors.pdf}
+\includegraphics[width=0.6\textwidth]{../priors.pdf}
 \label{Fig:Prior}
@@
 
-Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
-Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
+\small{Prior density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
+Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.}
 \end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure1.pdf}
-\end{figure}
-Bayesian phylogenetic age estimates for each of 36 penguin fossils plotted against their palaeontological age estimates, under two alternative evolutionary models. 
-The blue line shows the $x=y$. If the vertical line doesn't cross $x=y$, then the midpoint of the geological range is not in the phylogenetic 95\% HPD. 
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_younger_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{8_penguins_eNprior/8_fossilDatingHist_older_wide.pdf}
-\end{figure}
-Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
-Red boxes are the superimposed age ranges derived from geological data.
-\end{frame}
-
-\begin{frame}
-\begin{figure}
-\includegraphics[width=\textwidth]{Figure2.pdf}
-\end{figure}
-A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
-the precision is 1/variance in the marginal posterior distribution of the age).
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of age \& BF estimates between models}
-\begin{figure}%
-\centering
-\subfigure[Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareAgeM1M8.pdf}}\qquad
-\subfigure[Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareBFM1M8.pdf}}\\
-\label{fig:compareM1M8ab}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Comparison of probability \& error between models}
-\begin{figure}%
-\centering
-\subfigure[Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{comparePostM1M8.pdf}}\qquad
-\subfigure[Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}.]{\includegraphics[width=0.45\textwidth]{compareErrorM1M8.pdf}}\\
-\label{fig:compareM1M8cd}
-\end{figure}
-\end{frame}
-
-\end{document}

### 2016-01-27 fossilDating.tex
changes for arXiv submission

@@
-% !BIB TS-program = biber
 \documentclass[11pt]{article}
@@
 \usepackage{graphicx}
-\usepackage{hyperref}
+%\usepackage{hyperref}
 \usepackage{amsmath}
 \usepackage{xcolor}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
+\usepackage[style=authoryear,backend=bibtex,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
 \bibliography{fossilDating.bib}
@@
 \begin{figure}
-\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
+\includegraphics[width=6in]{8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
+\includegraphics[width=6in]{8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
@@
 
-\input{1_penguins_eNprior/1_summaryTable.tex}
+\input{1_summaryTable.tex}
 
-\input{8_penguins_eNprior/8_summaryTable.tex}
+\input{8_summaryTable.tex}
 
@@
 \begin{figure}
-\includegraphics{canids/1c_phyloAgeVsGeoAge.pdf}
+\includegraphics{1c_phyloAgeVsGeoAge.pdf}
 \caption{\label{fig:canidMorphVsGeo}
@@
 \begin{figure}
-\includegraphics[width=15cm]{canids/1_canids-1440516016976-tree5001.pdf}
+\includegraphics[width=15cm]{1_canids-1440516016976-tree5001.pdf}
 \caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.}

### 2016-04-05 fossilDating-revision.tex
start of revision

--- 2015-10-29 fossilDating.tex
@@
 \usepackage{xcolor}
-\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
-\bibliography{fossilDating.bib}
+
+\usepackage[numbers]{natbib}
+
+%\usepackage[style=numeric,
+%            firstinits=true,
+%            sorting=none,
+%            backend=biber,
+%            maxbibnames=10]{biblatex}
+%\renewbibmacro{in:}{%
+%  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
+
+%\bibliography{fossilDating.bib}
 
@@
 \newcommand{\todo}[1]{\textcolor{blue}{{TODO: #1}}}
+
+%\newcommand{\cite}[1]{\cite{#1}}
 
@@
 
-Contention between palaeontologists and molecular biologists over which data provides the most accurate inferences about evolutionary history has previously fostered an adversarial relationship between the two fields \autocite{Donoghue2007}.
-Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \autocite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \autocite{BentonAyala2003}.
-%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}.
+Contention between palaeontologists and molecular biologists over which data provides the most accurate inferences about evolutionary history has previously fostered an adversarial relationship between the two fields \cite{Donoghue2007}.
+Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \cite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \cite{BentonAyala2003}.
+%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \cite{tavare2002using}.
 % AJD: The following two sentences are inflammatory to palaeontologists
 %Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. 
-%Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
-The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang2014}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}.
-Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \autocite{Donoghue2007}.
+%Models of molecular evolution have been refined over many decades \cite{Felsenstein2004,Yang:2006yu}. 
+The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \cite{Felsenstein2004,Yang2014}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \cite{Donoghue2007}.
+Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \cite{Donoghue2007}.
 
@@
 The age of the fossil is determined either by radiometric aging of strata above and/or below the fossil, or more commonly by biostratigraphy.
-The difficulty lies in determining the appropriate ancestral divergence in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \autocite{Ho2009, Heled2012, HeledDrummond2015}.
-Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006,drummond2010}.
+The difficulty lies in determining the appropriate ancestral divergence in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \cite{Ho2009, Heled2012, HeledDrummond2015}.
+Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \cite{Thorne1998,thorne2005,yang2006,Drummond2006,drummond2010}.
 
-It may be less widely appreciated by molecular evolutionary biologists that the statistical phylogenetic revolution in molecular evolution has also been mirrored in the increasing application of statistical phylogenetic reasoning in macroevolutionary and systematic studies of the fossil record \autocite[e.g.][]{Foote1996,huelsenbeck1997maximum, tavare2002using, WagnerMarcot2013}.
+It may be less widely appreciated by molecular evolutionary biologists that the statistical phylogenetic revolution in molecular evolution has also been mirrored in the increasing application of statistical phylogenetic reasoning in macroevolutionary and systematic studies of the fossil record \cite{Foote1996,huelsenbeck1997maximum, tavare2002using, WagnerMarcot2013}.
 %This includes applying phylogenetic reasoning to questions of biogeography, character evolution, and the completeness of the fossil record.
@@
 
-The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \autocite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa.
-For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \autocite{meyer2014mitochondrial}. 
-The same technique has also been used to estimate the age of viral samples based on molecular sequence data \autocite[e.g.][]{gray2013evolutionary}.
+The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \cite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa.
+For example, this technique has been successfully employed to estimate the age of human subfossil remains based on an ancient mitochondrial genome sequence \cite{meyer2014mitochondrial}. 
+The same technique has also been used to estimate the age of viral samples based on molecular sequence data (e.g. \cite{gray2013evolutionary}).
 
 We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters.
-This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
+This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}. 
 It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
 This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa.
-A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \autocite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \textcite{gavr2014}.
+A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
 We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
@@
 %TS how many extant penguins?
-19 extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.
+19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \cite{Slater2015}.
 
@@
 
-\textcite{gavryushkina2015bayesian} described a ``total-evidence'' approach \autocite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
+Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach \cite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
 We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
-The model of time-tree phylogeny employed is the so-called fossilized birth-death process \autocite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \autocite{Gavr2013}.
+The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.
 
@@
 In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
-These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \autocite{Slater2015}.
+These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \cite{Slater2015}.
 
@@
 the fossilized birth-death process prior as described below.  
-For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
-The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}.
-We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.
+For the choices of morphological substitution model and tree prior we followed \cite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
+The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \cite{Lewis2001}.
+We also employed \Mrelaxed{} from \cite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \cite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}.
 
@@
 
-The second data set that we investigated was a morphological data matrix of 132 canid species \autocite[9 extant and 123 fossil;][]{Slater2015}. 
+The second data set that we investigated was a morphological data matrix of 132 canid species \cite[9 extant and 123 fossil;][]{Slater2015}. 
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
@@
 We applied the M1 model to this data set assuming a fixed geological age for each taxon.
-Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
+Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
 As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn.
-Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \textcite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
+Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
 To estimate the phylogenetic age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
@@
 In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
-Using the parameterization in \textcite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
+Using the parameterization in \cite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
 We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. This probability density will allow us to directly determine $p(H_1|M_\mathcal{T})$.
@@
 
-We simplify, using the equations for $p_i(t;d,r)$ given in \autocite{kendall1948}, to obtain,
+We simplify, using the equations for $p_i(t;d,r)$ given in \cite{kendall1948}, to obtain,
 \begin{eqnarray*}
@@
 \caption{\label{hist8_older}Probability density for the sampling times under the fossilized birth-death process.
-The dot-dashed line uses priors on the parameters as in \textcite{gavryushkina2015bayesian}.
+The dot-dashed line uses priors on the parameters as in \cite{gavryushkina2015bayesian}.
 The solid line uses the new prior with implicit assumptions on $T$ and $s$, the dashed line results from only assuming the implicit prior on $T$, the dotted line results from only assuming the  implicit prior on $s$.
@@
 
-We determined $p(t|M_\mathcal{T})$ for the prior distributions as in \textcite{gavryushkina2015bayesian}, 
+We determined $p(t|M_\mathcal{T})$ for the prior distributions as in \cite{gavryushkina2015bayesian}, 
 $$T: Unif(0,160), d: lognorm(-3.5,1.5), r: Unif(0,1), s: Unif(0,1).$$ This prior specification leads to a distribution of sampling time with almost all probability mass close to the present (Figure \ref{Fig:Prior}, dot-dashed line).
@@
 
-Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters compared to using the prior in  \textcite{gavryushkina2015bayesian}.
+Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters compared to using the prior in  \cite{gavryushkina2015bayesian}.
 This investigation of the prior distribution on trees and sampling times highlights that whenever using Bayes factors to test a hypothesis, we have to first investigate what our prior on the hypothesis is.
-In our example, the  prior from \textcite{gavryushkina2015bayesian} seemed reasonable for the parameters specified, however this prior puts a negligible weight on hypothesis $H_1$ for older fossils.
+In our example, the  prior from \cite{gavryushkina2015bayesian} seemed reasonable for the parameters specified, however this prior puts a negligible weight on hypothesis $H_1$ for older fossils.
 
-We want to note that the stepping stone sampling approach \autocite{Xie2011}
+We want to note that the stepping stone sampling approach \cite{Xie2011}
 %TS which paper to cite here?
@@
 
-\Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \textcite{gavryushkina2015bayesian}. 
+\Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \cite{gavryushkina2015bayesian}. 
 As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
@@
 %AJD added
-This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \autocite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.
+This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \cite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.
 
@@
 
-\printbibliography
+\bibliographystyle{vancouver}
+
+\bibliography{fossilDating}
 

### 2016-04-05 response-to-reviewers.tex
updated abstract

\documentclass[11pt]{article}

\usepackage{xcolor}

\newcommand{\response}[1]{\medskip{}\textcolor{blue}{{Response: #1}}\medskip{}}

\begin{document}

\title{Response to reviewers' comments of RSTB-2015-0129 entitled "Bayesian phylogenetic estimation of fossil ages}
\author{Alexei Drummond and Tanja Stadler}
\date{}
\maketitle

Dear Dr Drummond,

Thank you for submitting your article RSTB-2015-0129 entitled ``Bayesian phylogenetic estimation of fossil ages'' for publication in the upcoming theme issue of Philosophical Transactions B.

I am pleased to inform you that the reviews were positive and have just recommended some minor changes prior to final approval for publication. Please find the referees' comments below.

I glanced over your ms. and have two small comments.

First, please check the journal style and fix the formatting issues of the References section.  Right now some references use first names, some use first initials, and so on.

\response{We have changed the reference style to Vancouver, as per author instructions for Philosophical Transactions B.}

The second is that all of us (the two referees and myself) are surprised at the optimistic tone in the Abstract, and we can expect that an average reader will be as well.  As referee 1 commented on, your results do not really support this level of optimism.  Many people are uncomfortable with the idea of a morphological clock, because morphological characters are well known to have highly variable rates and undergo convergent evolution.  Perhaps you can be more explicit about the results, saying something like that the estimation accuracy is high in the sense that the HPD intervals include the true ages, but the precision is low in that the intervals are wide.  (I think for true ages younger than 30 myrs, an interval of 10-15 myrs is wide.)

\response{We have changed the abstract along the lines suggested. Specifically we have added the sentence ``We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision.''}

Yours sincerely,

Dr Ziheng Yang
z.yang@ucl.ac.uk

Referee reports:

\section*{Referee: 1}

Comments to Author(s)
This paper is timely, as there has been a recent surge in interest in modelling morphological evolution to date species divergence times. The topic is also very interesting. However, many papers have obtained very ancient divergence time estimates under the method. For example, Beck and Lee (2014, PRSB, 281:20141278) found very ancient estimates of mammal divergence times, and estimates had also very large uncertainties. Other works report similar patters (e.g. Arcila et al. 2014, MPE, 82:131). People are very puzzled about whether these ancient date estimates are reliable, or whether the Mk model (which is simply the Jukes-Cantor extended to k states) is too simplistic. I therefore was a bit surprised when I read the abstract of the present ms, as it portrays the method very optimistically. However, after having examined the ms, I can see that levels of uncertainty in estimated fossil ages are very large (e.g. Figure 7). I therefore think the optimistic tone needs to be moderated.

Overall I find the paper interesting. One strength is the modelling of the fossil ages as distributions based on stratigraphy, an important point we have recently highlighted (O'Reilly et al. 2015, TIGs, in press). I also find interesting the comparison between the stratigraphic age of fossils and the Bayesian estimates using the birth-death process.

I think the paper can be improved in certain areas. A bit of text needs to be added to clarify certain things (perhaps the ms is too short). Below I make several suggestions.

p.1. Abstract: change "phylogenetic age" by "phylogenetic estimates of fossil age". Please change throughout the ms.

p.4. Data set descriptions: For each dataset please add a small paragraph providing additional details, i.e. total number of characters, amount of missing data, fossil age ranges (say, penguin fossils range from xx Mya to yy Mya), and character state space (i.e. characters ranged from binary to k=x).

Maybe add a small "phylogentics analysis" section, between p.4 and p.5:

Here you can describe the analysis steps that are common to both datasets. Please provide brief descriptions of models M1 and M8. I note that in Lewis' Mk model, the k stands for the dimension of the character substitution matrix, so M2 means a 2x2 substitution matrix (binary characters). This does not seem to be the case here. The authors could either change their notation slightly (suggestion: Mk-1, Mk-8), or explicitly state that their nomenclature does not follow Lewis.

Provide a better description of the stepwise analysis of fossil ages. My impression (but I am not totally sure) is that in one analysis, fossil ages are unknown parameters and have uniform prior distributions set according to the stratigraphic bounds. Then, in stepwise analysis, each fossil age is in turn represented not by the stratigraphic uniform distribution, but by a prior density derived from the fossilised birth-death process. Give some details please. I was trying to work out the general form of the posterior distribution, I was guessing something like:

$$p(tip\_ages) p(node\_ages) L(D|ages,tree)$$

where $p(tip\_ages)$ is the prior density on fossil ages using uniform distributions, $p(node\_ages)$ is the prior on internal node ages given the BD process, and $L(D|ages,tree)$ is the likelihood of the morphological data D, given the ages and the topology calculated using Lewis (2001) formula. However, it seems from the text in p.5 that another likelihood term may be involved. For the stepwise deletion procedure, I am guessing the priors are something like $p(tip\_ages\hat{}) p(node\_ages, tip\_age*)$, where now $tip\_ages\hat{}$ includes all the fossil ages except the one under analysis, which has been moved into the other BD-based prior term $(tip\_age*)$. Please provide general forms for the posteriors and make the connection with equation (1).

Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.

Please give the prior on the morhpological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model? I am also confused, do sampling times refer to the ages of fossil tips? Aren't these unknown parameters being estimated in the MCMC? Then they would appear as terms in the prior and the likelihood. (?)

p.9 Results section. For both data sets the authors state that the relaxed clock model fits the data better, but I can't find the BF for this. Also, for both data sets, the authors have sections titled "Penguins conform well to a strict morphological clcok" and "Canids conform well to a morphological clock", but these headings seem to contradict the statement in lines 48-49 p.12 of discussion. I am also puzzled as the authors have not quantified here departure from the clock.

Please provide a table with prior and posterior mean estimates of the age of the root for both data sets under all models used, and 95% CI. Similarly, in the same table provide prior and posterior estimates of mean morphological substitution rate for all data sets and models (relaxed and strict clock). For the relaxed-clocks, provide prior and posterior estimates of rate-variation parameter. Then, when the authors discuss whether the data fit or not the strict clock, they should refer to the estimates in the table.

Provide axis labels and axis titles in Figures 2, 6 and 7.

I think Figures 4 and 5 should be merged into one figure.

Figure 3 and 6: Mention the dataset in the legend (canids? penguins?).

Figure 6, panels (b), (c) and (d) seem empty.

Figure 7: Add dots indicating posterior means.

Figure 8: Please replace this with the consensus tree and provide 95% CIs of internal node ages.

I could not find references to Figures 3 and 8 in the text.

In the results section, please provide some quantification of the uncertainty in time estimates. For example,  the authors could provide estimates of the precision as (95% CI width)/(posterior mean of fossil age) and express this as a percentage. Roughly, looking at figure 7, it seems that for some nodes the 95% CI limits would range from anywhere between 50% to 150% of the point stratigraphic estimate. Clearly, uncertainty here is very, very large.

Considering the point above (and the large CIs observed in figures 2 and 7), it is clear that estimates are not precise (note, I refer to accuracy as bias of estimates and precision as 1/variance of estimates). So, please remove any references in the ms about having obtained precise estimates. Estimates do seem to be accurate, as the posterior means are close to stratigraphic mid-points, but the authors must stress, in the abstract, results and discussion, that uncertainties are large.

As I pointed out at the beginning of this review, many papers report ancient estimates of node ages using morphological data. How do the posterior age of the root for canids and penguins obtained in this paper compare with estimates from traditional methods? Please mention briefly in discussion. For example, some of the previous estimates could be added to the table I suggested.

I think the insistence by the authors to use the term "morhpological clock" will cause controversy. Please provide some evidence that departure of the clock is small (for example, posterior estimates of the rate variation parameter are, say, close to zero). Otherwise I suggest removing the term.

As a side note I would like to point out that there is an extensive literature on morphological rates of evolution, indeed, before molecular data became available in the sixties, most work on evolution was effectively work on phenotypic (morphological) evolution. One of the reasons the molecular clock was so controversial when it was initially proposed, was precisely because early workers were skeptical that molecular rates should be constant when morphological rates were so variable. For example, in his 1985 book on the neutral theory, Kimura has a whole chapter contrasting morphological vs. molecular rates (I strongly encourage the authors to read this), and he provides lots of citation to early works (for example, Haldane's) on morphological rates. Michael Lynch has some theoretical work where he shows that under neutral mutation and random drift, morphological clocks can be achieved (can't remember the reference) and he gives examples in humans. It may be worth looking this up if the authors want to add a balance discussion of this in their ms. There are also lots of papers throughout the 80s and 90s on morphological rates. One thing that surprises me with the new morphological papers is that a lot of this literature has been completely ignored. I encourage the authors to read and cite some of these works.

I am troubled about the sensitivity of the BF to the prior in this case. Can the authors give some clear, easy to follow, recommendations for biologists that may wish to use the method in their own works? Another issue is that when the model is changed, the BF change. The authors should point out to the reader that rejecting the stratigraphic estimate of a fossil with BF does not prove that the stratigraphic age is wrong. If the morphological or clock model are misspecified (i.e. they are  bad description of the data), then I would expect the BF to fail. For example, if a lineage has shown explosively fast rates of evolution, then I would expect the Bayesian phylogenetic estimates of the fossil age to contradict the stratigraphic range if a strict clock is imposed.

Best wishes,
Mario dos Reis.

\section*{Referee: 2}

Comments to Author(s)
This manuscript provides evidence that Bayesian modelling of morphological evolution in a phylogenetic context can predict the age of a fossil taxon with a considerable degree of accuracy. Frankly, I find the results surprising and I must say that this has helped to assuage many of the doubts I have had about the efficacy of such methods, and I expect the same will be said about many others in this field. Certainly this has the potential to settle some of the longstanding debates between molecular and morphological systematists and has a number of potential applications. I look forward to further testing across different taxonomic datasets and deeper divergence times to see if the predictive power of this method holds up.

I only recommend minimal revisions to the manuscript.

General comments:
Perhaps another reason for the discrepancy between phylogenetic and geological age is due to stasis of morphology through the fossil record. For example, imagine the anatomy of a specific fossil taxon appeared 30 Ma but underwent no morphological evolution through the next 10 million years. If this taxon is recovered from strata dating to 20 Ma, and the models you are using sufficiently capture morphological evolutionary rates, the phylogenetic age estimate should predate that of the fossil's appearance by, on average, 10 million years. This may also be a useful application of this method: making predictions of where a taxon should appear prior to the known fossil record for that particular species.

One concern I have with this method, which perhaps doesn't apply to these datasets, is how it handles situations where there is rampant morphological convergence across species in the dataset. For example, previous analyses have shown that large morphological datasets group together mammals with convergent ecotypes (e.g., ant-eating, tree-dwelling, hoofed; Springer et a. 2013 Science) and squamates without legs (e.g., snakes, amphisbaenians, anniellids; Gauthier et al. 2012 Bulletin of the Peabody Museum of Natural History), respectively, in sharp contrast to both molecular data and, frequently, the fossil record. Presumably it should lead to highly inaccurate phylogenetic age estimates, though this might also be a potential application of the method (i.e., highly inaccurate dates may be suggestive of homoplasy appropriately grouping taxa together).

This raises a further question of whether this method can be used in conjunction with molecular data to correct for morphological homoplasy. Beyond this, perhaps one could test how the molecular dataset can influence the phylogenetic age of these fossils, and could provide further insights regarding whether divergence times derived from molecular clocks are accurate. I'm not suggesting more analyses, but perhaps further discussion of these issues.

Specific comments:
Inconsistent with British spelling of palaeontology (=paleontology) and palaentological in text/figures/legends

Page 3 line 27: reference is not completely within parentheses

Page 6 line 53: T and large d, we obtain very large trees with arbitrary <--(Typo?) many species

Page 8 line 54 (poor wording or maybe typo): In general, even if stepping stone approaches are appropriate, we recommend to investigate*** P(H1|M) to investigate*** the prior assumption on the hypothesis to be tested. Such an investigation***

Page 9: for the species that fell far beyond the regression line, do you have any hypotheses for these discrepancies? Were there insufficient coded characters (including nonoverlapping characters), homoplasy or reason to doubt the geological age?

Figure 3: I did not see a reference for this in the text

Figures 4 and 5: It appears that the figure legends were cut off. Both ended with "red boxes"

Figure 8: I did not see a reference for this in the text

Tables 1 and 2: I think it would be useful to readers if you add the geological age ranges of the fossils, along with upper and lower bounds of the strata.

\end{document}

### 2016-04-05 fossilDating-revision.tex
updated abstract

@@
 We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
-We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
+We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision. 
 In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
-The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
 We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 

### 2016-04-05 response-to-reviewers.tex
removed instances of phylogenetic age

@@
 
+\response{We have changed the abstract to moderate the tone.
+Specifically we have added the sentence ``We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision.''}
+
 Overall I find the paper interesting.
@@
 
-p.1. Abstract: change "phylogenetic age" by "phylogenetic estimates of fossil age". Please change throughout the ms.
+p.1. Abstract: change ``phylogenetic age'' by ``phylogenetic estimates of fossil age''. Please change throughout the ms.
+
+\response{We have replaced all instances of ``phylogenetic age'' with variants of ``phylogenetic estimates of fossil age''.}
 
@@
 
-Please give the prior on the morhpological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).
+Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).
 
@@
 
-Please provide a table with prior and posterior mean estimates of the age of the root for both data sets under all models used, and 95% CI.
+Please provide a table with prior and posterior mean estimates of the age of the root for both data sets under all models used, and 95\% CI.
 Similarly, in the same table provide prior and posterior estimates of mean morphological substitution rate for all data sets and models (relaxed and strict clock).
@@
 
-Figure 8: Please replace this with the consensus tree and provide 95% CIs of internal node ages.
+Figure 8: Please replace this with the consensus tree and provide 95\% CIs of internal node ages.
 
@@
 In the results section, please provide some quantification of the uncertainty in time estimates.
-For example,  the authors could provide estimates of the precision as (95% CI width)/(posterior mean of fossil age) and express this as a percentage.
-Roughly, looking at figure 7, it seems that for some nodes the 95% CI limits would range from anywhere between 50% to 150% of the point stratigraphic estimate.
+For example,  the authors could provide estimates of the precision as (95\% CI width)/(posterior mean of fossil age) and express this as a percentage.
+Roughly, looking at figure 7, it seems that for some nodes the 95\% CI limits would range from anywhere between 50\% to 150\% of the point stratigraphic estimate.
 Clearly, uncertainty here is very, very large.

### 2016-04-05 fossilDating-revision.tex
removed instances of phylogenetic age

@@
 In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa.
-We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
+We exploit this statistical framework to investigate the internal consistency of these models by producing phylogenetic estimates of the age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
 We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision. 
-In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
+In fact in the two data sets we analyze the phylogenetic estimates of fossil age is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
 The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
@@
 %This includes applying phylogenetic reasoning to questions of biogeography, character evolution, and the completeness of the fossil record.
-Here we extend this tradition of applying phylogenetic reasoning to the fossil record by focusing on the question of what phylogenetic inference techniques can tell us about the ``phylogenetic age'' of a fossil, based solely on its morphological characteristics and through them, its phylogenetic and temporal relationships with a set of reference fossils.
+Here we extend this tradition of applying phylogenetic reasoning to the fossil record by focusing on the question of what phylogenetic inference techniques can tell us about the age of a fossil, based solely on its morphological characteristics and through them, its phylogenetic and temporal relationships with a set of reference fossils.
 
@@
 A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
-We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
+We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
 The method is applied to two rich and well-characterized morphological data sets: (i) 
@@
 Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach \cite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
-We extend their work further by investigating the consistency between the estimated phylogenetic age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
+We extend their work further by investigating the consistency between the phylogenetic estimate of the age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
 The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.
@@
 This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree.
-We refer to this estimate of a fossil's age as its {\em phylogenetic age}. 
-In estimating each of the fossils' phylogenetic ages in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
+We refer to this as the phylogenetic estimate of the fossil's age. 
+In phylogenetically estimating the age of each of the fossils in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
 These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \cite{Slater2015}.
 
-\subsection*{Estimating the phylogenetic ages of penguin fossils}
+\subsection*{Phylogenetic estimates of the ages of penguin fossils}
 
@@
 
-\subsection*{Estimating the phylogenetic ages of Canid fossils}
+\subsection*{Phylogenetic estimates of the ages of Canids fossils}
 
@@
 Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
-To estimate the phylogenetic age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
+To phylogenetically estimate the age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
 %TS: which priors on d, r, s, T did we use now?
@@
 \subsection*{Penguins conform well to a strict morphological clock}
-Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
-Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.903$. 
+Although \Mstrict{} is a very simple model, the phylogenetic estimates of the ages of the penguin fossils were remarkably consistent with their palaeontological age ranges. 
+Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic estimates of fossil age. The points in this plot have $R^2 = 0.903$. 
 The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
@@
 As judged by Bayes factors, only one fossil exhibited strong evidence (i.e.
-$\text{log BF} < -3.0$) that the phylogenetic age was inconsistent with the geological age range.
+$\text{log BF} < -3.0$) that the phylogenetic estimate of fossil age was inconsistent with the geological age range.
 The log BF for {\em Paraptenodytes antarcticus} was -3.4.
@@
 $\text{log BF} > 3.0$).
-Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
-These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively.
+Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range.
+These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively.
 All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.3Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
+mirandus}: 6.3Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
 antarcticus}: 29.9 vs 22, {\em S.
@@
 \Mrelaxed{} was the best-fitting model for the penguin data set according to the analysis of \cite{gavryushkina2015bayesian}. 
-As with \Mstrict{} this model produced phylogenetic age estimates that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
+As with \Mstrict{} this model produced phylogenetic estimates of fossil age that were very concordant with the geological age ranges of the fossils (Figure \ref{fig:phyloAgeVsGeoAge}b), with an overall $R^2 = 0.924$.
 The median error was 2.05 Myr across all 36 fossils.
-In this analysis none of the fossils exhibited any evidence (i.e. $\text{log BF} < 0.0$) that the phylogenetic age was inconsistent with the geological age range. 
-However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic age being within the palaeontological age range.
-These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively.
+In this analysis none of the fossils exhibited any evidence (i.e. $\text{log BF} < 0.0$) that the phylogenetic estimate of fossil age was inconsistent with the geological age range. 
+However if we consider the posterior probability that the fossil is in the geological age range then five of the 36 fossils had a posterior probability $< 0.05$ for \Mrelaxed{}, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range.
+These five fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus}, {\em Perudyptes devriesi}, {\em Sphenicus muizoni} and {\em Waimanu manneringi} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.035, 0.018, 0.046, 0.004, 0.037 respectively.
 All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.7Myr vs 10Myr (phylogenetic age versus palaeontological age), {\em P.
+mirandus}: 6.7Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
 antarcticus}: 28.0 vs 22, {\em P.
@@
 A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}. 
-The individual marginal posterior distributions of phylogenetic age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
+The individual marginal posterior distributions of phylogenetic estimates of fossil age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
 
@@
 \caption{\label{fig:phyloAgeVsGeoAge}
-The Bayesian phylogenetic age estimate (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates, under two alternative site and molecular clock models.
+The Bayesian phylogenetic estimate of fossil age (median of marginal posterior) for each of the 36 penguin fossils plotted against their palaeontological age estimates, under two alternative site and molecular clock models.
 The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits. 
@@
 \includegraphics[width=5in]{Figure2.pdf}
-\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
+\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic estimate of fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
 the precision is 1/variance in the marginal posterior distribution of the age).}
@@
 \includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
-\caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
+\caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
 Red boxes are the superimposed age ranges derived from geological data.}
@@
 \includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
-\caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic age estimate of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
+\caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
 Red boxes are the superimposed age ranges derived from geological data.}
@@
 Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant. 
-Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
+Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.
 %TS is the following quantified? otherwise I'd delete the next sentence!?
@@
 Comparison of \Mstrict{} against \Mrelaxed{}.
-(a) Estimated phylogenetic age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
+(a) Phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
 \end{figure}
@@
 
-The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
+The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
 The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
-Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
+Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
 The median error is $1.66$My, which is even more accurate than the penguin data set.
@@
 \caption{\label{fig:canidMorphVsGeo}
-The Bayesian phylogenetic age estimate (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}.
+The Bayesian phylogenetic estimate of fossil age (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}.
 The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval. Blue line shows the $x=y$. 
-If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic age. The 16 fossils for which this is the case are labelled.}
+If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic estimate of fossil age. The 16 fossils for which this is the case are labelled.}
 %TS I find this plot very hard to read as many fossils are overlaying...can we do something there for better visualization?
@@
 Lloyd for insightful comments and suggestions on early drafts of this manuscript.
+We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.
 AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).

### 2016-04-06 response-to-reviewers.tex
phylogenetic analyses

@@
 
+\usepackage[numbers]{natbib}
+
 \newcommand{\response}[1]{\medskip{}\textcolor{blue}{{Response: #1}}\medskip{}}
+\newcommand{\respquote}[1]{\medskip{}\textcolor{blue}{``{\it #1}''}\medskip{}}
 
@@
 characters ranged from binary to k=x).
+
+\response{We added the following paragraph:}
+
+\respquote{We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
+species. 
+We used the same subset of this original data as in \cite{gavryushkina2015bayesian}. 
+Additionally we discarded the molecular sequence data from the living species. 
+The resulting morphological data matrix we used contains 36 fossil species, 19 extant species and the morphological matrix contains 202 characters. 
+The morphological characters ranged from binary to k=7. 
+The majority of these (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
+See \cite{gavryushkina2015bayesian} for further details.}
 
@@
 
+\bibliographystyle{vancouver}
+
+\bibliography{fossilDating}
+
 \end{document}

### 2016-04-06 fossilDating-revision.tex
phylogenetic analyses

@@
 
-\newcommand{\Mstrict}{{M1}}
-\newcommand{\Mrelaxed}{{M8}}
+\newcommand{\Mstrict}{{Mk-1}}
+\newcommand{\Mrelaxed}{{Mk-8}}
 
@@
 
-%TS: check:
+We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
+species. 
+We used the same subset of this original data as in \cite{gavryushkina2015bayesian}. 
+Additionally we discarded the molecular sequence data from the living species. 
+The resulting morphological data matrix we used contains 36 fossil species, 19 extant species and the morphological matrix contains 202 characters. 
+The morphological characters ranged from binary to k=7. 
+The majority of these (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
+See \cite{gavryushkina2015bayesian} for further details.
+
 For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by 
@@
 %TS: which priors on d, r, s, T did we use now?
+
+\subsection*{Phylogenetic analyses}
+
+Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model 
+\citep{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
+As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \citep{ksepka2012} as unordered.
+
+We applied two phylogenetic models to each data set: Mk-1 and Mk-8.
+Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
+Model Mk-8 partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
 

### 2016-04-12 canids/stratRanges/createCanidXMLs.sh
canid analysis update

@@ -1,0 +1,3 @@
+#!/bin/bash
+java -cp ../../out/production/fossilDating/ FossilDating 1_canids_ranges_3_subset.xml 80 ../canidFossilTaxa_subset.txt idref 
+

### 2016-04-12 src/FossilDating.java

@@ -73,7 +73,7 @@ public class FossilDating {
                 writer.write(l);
                 writer.write("\n");
                 if (l.matches(".*INSERT SINGLE SAMPLINGDATE.*")) {
-                    writer.write("    <operator spec='SampledNodeDateRandomWalker' windowSize=\"1.\"  tree=\"@tree\" weight=\"3.\">\n" +
+                    writer.write("    <operator spec='SampledNodeDateRandomWalkerForZeroBranchSATrees' windowSize=\"1.\"  tree=\"@tree\" weight=\"1.\">\n" +
                             "        <taxonset spec=\"TaxonSet\">\n" +
                             "            <taxon " + idString + "=\"" + taxon + "\" spec=\"beast.evolution.alignment.Taxon\"/>\n" +
                             "        </taxonset>\n" +

### 2016-04-12 src/FossilDating.java
longer XMLs random walk for focal taxon age

@@ -73,7 +73,7 @@ public class FossilDating {
                 writer.write(l);
                 writer.write("\n");
                 if (l.matches(".*INSERT SINGLE SAMPLINGDATE.*")) {
-                    writer.write("    <operator spec='SampledNodeDateRandomWalkerForZeroBranchSATrees' windowSize=\"1.\"  tree=\"@tree\" weight=\"1.\">\n" +
+                    writer.write("    <operator spec='SampledNodeDateRandomWalkerForZeroBranchSATrees' windowSize=\"2.\"  tree=\"@tree\" weight=\"0.5\" useWindowSizeWithSamplingDates=\"true\">\n" +
                             "        <taxonset spec=\"TaxonSet\">\n" +
                             "            <taxon " + idString + "=\"" + taxon + "\" spec=\"beast.evolution.alignment.Taxon\"/>\n" +
                             "        </taxonset>\n" +

### 2016-04-14 response-to-reviewers.tex

--- 2016-04-05 response-to-reviewers.tex
@@
 characters ranged from binary to k=x).
+
+\response{We added the following paragraph:}
+
+\respquote{We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
+species. 
+We used the same subset of the morphological data as in \cite{gavryushkina2015bayesian} but we did not use the molecular sequence data from the living species. 
+The morphological data matrix we used contains 36 fossil species, 19 extant species and 202 characters (ranging from binary to k=7). 
+The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
+See \cite{gavryushkina2015bayesian} for further details of data selection.}
 

### 2016-04-14 fossilDating-revision.tex

@@
 species. 
-We used the same subset of this original data as in \cite{gavryushkina2015bayesian}. 
-Additionally we discarded the molecular sequence data from the living species. 
-The resulting morphological data matrix we used contains 36 fossil species, 19 extant species and the morphological matrix contains 202 characters. 
-The morphological characters ranged from binary to k=7. 
-The majority of these (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
-See \cite{gavryushkina2015bayesian} for further details.
+We used the same subset of the morphological data as in \cite{gavryushkina2015bayesian} but we did not use the molecular sequence data from the living species. 
+The morphological data matrix we used contains 36 fossil species, 19 extant species and 202 characters (ranging from binary to k=7). 
+The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
+The fossil ages ranged from 
+See \cite{gavryushkina2015bayesian} for further details of data selection.
 
@@
 
-The second data set that we investigated was a morphological data matrix of 132 canid species \cite[9 extant and 123 fossil;][]{Slater2015}. 
+The second data set that we investigated was a morphological data matrix of 132 canid species \cite[9 extant and 123 fossil;][]{Slater2015} with 123 characters (ranging from binary to k=5). 
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
@@
 
-We applied the M1 model to this data set assuming a fixed geological age for each taxon.
+We applied the \Mstrict{} model to this data set assuming a fixed geological age for each taxon.
 Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
@@
 
-Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant. 
+Overall the results of analyzing the penguin data set with the \Mstrict{} and \Mrelaxed{} models were strikingly concordant. 
 Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic estimates of fossil ages of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.

### 2016-04-15 fossilDating-revision.tex
update of canid analysis

@@
 \newcommand{\Mrelaxed}{{Mk-8}}
+\newcommand{\ncanidfossils}{{116}}
 
@@
 %TS how many extant penguins?
-19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \cite{Slater2015}.
+19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.
 
@@
 In phylogenetically estimating the age of each of the fossils in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
-These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of 123 canid fossils and their extant relatives \cite{Slater2015}.
+These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of \ncanidfossils{} canid fossils and their extant relatives \cite{Slater2015}.
 
@@
 
-The second data set that we investigated was a morphological data matrix of 132 canid species \cite[9 extant and 123 fossil;][]{Slater2015} with 123 characters (ranging from binary to k=5). 
+The second data set that we investigated was a morphological data matrix of 125 canid species \cite[9 extant and \ncanidfossils{} fossil;][]{Slater2015} with 122 characters (ranging from binary to k=5).
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
-For this data set we did not have a geological age range for each fossil.
-Instead we used the point estimates for each fossil species' geological age that were employed by the author of the original study.
+We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms).
 
-We applied the \Mstrict{} model to this data set assuming a fixed geological age for each taxon.
-Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils.
-As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn.
-Since we can't calculate Bayes factors without a fossil age range (see following section) we simply used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
+We applied the \Mstrict{} model to this data and assumed a uniform prior on each fossil's age within their respective stratigraphic ranges.
+Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the fossils.
+As with the penguin data set we also performed an analysis for each of the \ncanidfossils{} canid fossils in turn.
+We used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
 To phylogenetically estimate the age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
@@
 
-We applied two phylogenetic models to each data set: Mk-1 and Mk-8.
-Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
-Model Mk-8 partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
+We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}.
+\Mstrict{} assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
+Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
 
@@
 
-The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
-The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
-Only 16 out of 123 fossils (13\%) don't have the paleontological age in the credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
-The median error is $1.66$My, which is even more accurate than the penguin data set.
-This data set contains half as many morphological characters as does the penguin data set (122 versus $245$) but despite that the individual age estimates have more precision 
-%TS why 99\% for candids and 95\% for penguins? 
-%AJD I will check this later.
-(median HPD range = 6.5 My for canids as opposed to 9.6My for penguins).
+The canid data set shows remarkably good consistency between stratigraphic age ranges and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
+The $R^2 = 0.897$ between the phylogenetic and stratigraphic ages ranges (see Figure \ref{fig:canidMorphVsGeo}). 
+Only 13 out of \ncanidfossils{} fossils (11\%) don't have the mean stratigraphic age in the 95\% credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
+The median error is $1.56$My, which is even more accurate than the penguin data set.
+This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$) but despite that the individual age estimates have about double the precision (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
+This may well be because the average age of the penguin fossils is considerably.
 
 \begin{figure}
-\includegraphics{canids/1c_phyloAgeVsGeoAge.pdf}
+\includegraphics{canids/stratRanges/1c_phyloAgeVsGeoAge.pdf}
 \caption{\label{fig:canidMorphVsGeo}
-The Bayesian phylogenetic estimate of fossil age (median and credible interval of marginal posterior) for each of the 123 canid fossils plotted against their palaeontological age point estimates, under a strict morphological clock model \Mstrict{}.
-The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 99\% HPD interval. Blue line shows the $x=y$. 
-If the vertical line doesn't cross $x=y$, then the geological age is not in the credible interval of the phylogenetic estimate of fossil age. The 16 fossils for which this is the case are labelled.}
+The Bayesian phylogenetic estimate of fossil age (median and 95\% credible interval of marginal posterior) for each of the \ncanidfossils{} canid fossils plotted against their stratigraphic age ranges, under a strict morphological clock model \Mstrict{}.
+The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. Blue line shows the $x=y$. 
+If the vertical line doesn't cross $x=y$, then the mean of the stratigraphic age range is not in the credible interval of the phylogenetic estimate of fossil age.
+The 13 fossils for which this is the case are labelled.}
 %TS I find this plot very hard to read as many fossils are overlaying...can we do something there for better visualization?
@@
 \begin{figure}
-\includegraphics[width=15cm]{canids/1_canids-1440516016976-tree5001.pdf}
+\includegraphics[width=15cm]{canids/stratRanges/1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
 \caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.}
@@
 Lloyd for insightful comments and suggestions on early drafts of this manuscript.
+We are very grateful to Graham Slater for providing stratigraphic age ranges for the Canid fossils.
 We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.

### 2016-04-15 fossilDatingFunctions.r

@@ -151,14 +151,14 @@ writeSummaryTable <- function(filename, caption, label, df) {
 ###############################################################################
 # produce the phylo age versus geo age plot and write to pdf file
 ###############################################################################
-createPhyloAgeVsGeoAgePDF <- function(filename, df, fossilTable, labelled, labelPos, min=0, max=65) {
+createPhyloAgeVsGeoAgePDF <- function(filename, df, fossilTable, labelled, labelPos, min=0, max=65, xlabel="paleontological age (Myr)") {
 
   require(plotrix)
 
   pdf(file=filename, width=5, height=5, pointsize=10)
 
   par(mar = c(4,4,1,1) + 0.1)
-  plot(c(min,max*0.925),c(min,max*0.925), type="n", xlab="paleontological age (Myr)", ylab="Bayesian phylogenetic age estimate (Myr)", ylim=c(min,max), xlim=c(min,max))
+  plot(c(min,max*0.925),c(min,max*0.925), type="n", xlab=xlabel, ylab="Bayesian phylogenetic age estimate (Myr)", ylim=c(min,max), xlim=c(min,max))
 
   draw.ellipse(df$geo, df$est, df$geo-fossilTable$Lower, df$est-df$hpd_lower, angle = 0, segment = c(180,360), border =rgb(0.5,0.5,0.5,0.2), col=rgb(0.5,0.5,0.5,0.1), arc.only=FALSE)
 
@@ -346,20 +346,30 @@ createHistPlotFigures <- function (filePrefix, names, height, fossilTable, combi
 ####################################################################
 # produce the precision versus number of known characters plot
 ####################################################################
-createPrecisionVsKnownCharactersFigure <- function(filename, df, labelled, labelPos) {
+createPrecisionVsKnownCharactersFigure <- function(filename, df, labelled, labelPos, relativeHPD=FALSE) {
 
   pdf(file=filename, width=5, height=5, pointsize=10)
-  plot(df$characters,df$precision, pch=16, xlab="Number of known characters", ylab="Precision of phylogenetic age estimate (1/variance)", xlim=c(0,max(df$characters)*1.05))
   
-  text(df$characters[labelled], df$precision[labelled], labels=df$names[labelled],pos=labelPos,offset=0.45, cex=0.7)
+  pr <- df$precision
[…]

### 2016-04-16 fossilDating-revision.tex
graham slater's comments

@@
 \usepackage{amsmath}
-\usepackage{xcolor}
+\usepackage[dvipsnames]{xcolor}
 
@@
 \newcommand{\todo}[1]{\textcolor{blue}{{TODO: #1}}}
+
+\newcommand{\alexei}[1]{\textcolor{NavyBlue}{{#1}}}
 
@@
 
-We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
+\alexei{We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
 species. 
@@
 The fossil ages ranged from 
-See \cite{gavryushkina2015bayesian} for further details of data selection.
+See \cite{gavryushkina2015bayesian} for further details of data selection.}
 
@@
 
-The second data set that we investigated was a morphological data matrix of 125 canid species \cite[9 extant and \ncanidfossils{} fossil;][]{Slater2015} with 122 characters (ranging from binary to k=5).
+\alexei{The second data set that we investigated was a morphological data matrix of 125 canid species \cite[9 extant and \ncanidfossils{} fossil;][]{Slater2015} with 122 characters (ranging from binary to k=5).
+}
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
-We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms).
+\alexei{We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms). }
 
@@
 Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model 
-\citep{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
-As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \citep{ksepka2012} as unordered.
+\cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
+As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered.
 
@@
 \includegraphics[width=15cm]{canids/stratRanges/1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
-\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, showing two main clades, one containing the extant taxa and another constituted entirely of extinct fossil species.}
+\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, \alexei{showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}}
 \end{figure}
@@
 
-In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model.
-We found that although a strict morphological clock does a surprisingly good job of estimated fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
+In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.
+We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
 We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
-This may suggest that the larger reference set of fossils in the canid data set (122 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates.
+This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates.
+However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.
 
 There are a diverse array of potential applications for this methodology. 
-The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found or a complete lack of provenance data (e.g.
+The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, or because the fossil was not associated with a sediment of known age \cite{Berger2015}, or because of a complete lack of provenance data (e.g.
 a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}).
@@
 
-Future work may investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
+\alexei{One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the sediment the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic tuffs above or below the fossiliferous horizon and uncertainty about the age difference between the volcanic tuff and the horizon the fossil is in).
+On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
+For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5\text{myr}$ (pers. comm. Graham Slater). 
+In fact the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges. 
+This raises the question of whether multiple specimens of a single species that span a significant time frame and/or different localities should be coded separately.
+Even if not coded as separate taxa it may be possible to use the information about the individual specimen ages explicitly in some other way.
+We leave these considerations for future work. }
+Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
 
@@
 Lloyd for insightful comments and suggestions on early drafts of this manuscript.
-We are very grateful to Graham Slater for providing stratigraphic age ranges for the Canid fossils.
+We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
 We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.

### 2016-04-16 fossilDating-revision.tex
small changes

@@
 
-\newcommand{\alexei}[1]{\textcolor{NavyBlue}{{#1}}}
+\newcommand{\gslater}[1]{\textcolor{NavyBlue}{{#1}}}
+\newcommand{\alexei}[1]{\textcolor{red}{{#1}}}
 
@@
 \includegraphics[width=15cm]{canids/stratRanges/1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
-\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, \alexei{showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}}
+\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, \gslater{showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}}
 \end{figure}
@@
 There are a diverse array of potential applications for this methodology. 
-The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, or because the fossil was not associated with a sediment of known age \cite{Berger2015}, or because of a complete lack of provenance data (e.g.
+The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, \alexei{or because the fossil was not associated with a horizon of known age e.g.
+\cite{Berger2015},} or because of a complete lack of provenance data (e.g.
 a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}).
@@
 
-\alexei{One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the sediment the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic tuffs above or below the fossiliferous horizon and uncertainty about the age difference between the volcanic tuff and the horizon the fossil is in).
+\gslater{One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
 On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
 For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5\text{myr}$ (pers. comm. Graham Slater). 
-In fact the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges. 
-This raises the question of whether multiple specimens of a single species that span a significant time frame and/or different localities should be coded separately.
-Even if not coded as separate taxa it may be possible to use the information about the individual specimen ages explicitly in some other way.
-We leave these considerations for future work. }
+In this context it is interesting to note the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges.
+This raises the question of whether multiple specimens of a single species that span a significant time frame and/or different localities should be coded as separate taxa.
+Even if not coded as separate taxa it may be possible to extend the method used here to explicitly account for multiple specimens and associated ages when a fossil species is represented by more than one fossil.
+We leave these considerations for future work.
+}
  Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
@@
 Lloyd for insightful comments and suggestions on early drafts of this manuscript.
-We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
-We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.
-AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).
+\alexei{We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
+We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.} AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).
 AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).

### 2016-04-26 response-to-reviewers.tex
response to reviewers

@@
 \newcommand{\response}[1]{\medskip{}\textcolor{blue}{{Response: #1}}\medskip{}}
-\newcommand{\respquote}[1]{\medskip{}\textcolor{blue}{``{\it #1}''}\medskip{}}
+\newcommand{\respquote}[1]{{\medskip{}\textcolor{blue}{{\it #1}}\medskip{}}}
+
+\newcommand{\Mstrict}{{Mk-1}}
+\newcommand{\Mrelaxed}{{Mk-8}}
 
@@
 The authors could either change their notation slightly (suggestion: Mk-1, Mk-8), or explicitly state that their nomenclature does not follow Lewis.
+
+\response{We added a section called Phylogenetic analyses with following text:}
+
+\respquote{Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
+As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered.
+We applied two phylogenetic models to the penguin data set: Mk-1 and Mk-8.
+Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
+Model Mk-8 partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.}
 

### 2016-04-26 fossilDating-revision.tex
response to reviewers

@@
 
-\newcommand{\gslater}[1]{\textcolor{NavyBlue}{{#1}}}
-\newcommand{\alexei}[1]{\textcolor{red}{{#1}}}
+%\newcommand{\gslater}[1]{\textcolor{NavyBlue}{{#1}}}
+%\newcommand{\alexei}[1]{\textcolor{red}{{#1}}}
 
@@
 
-\alexei{We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
+%\alexei{%
+We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
 species. 
@@
 The fossil ages ranged from 
-See \cite{gavryushkina2015bayesian} for further details of data selection.}
+See \cite{gavryushkina2015bayesian} for further details of data selection.
+%}%
 
@@
 
-\alexei{The second data set that we investigated was a morphological data matrix of 125 canid species \cite[9 extant and \ncanidfossils{} fossil;][]{Slater2015} with 122 characters (ranging from binary to k=5).
-}
+%\alexei{%
+The second data set that we investigated was a morphological data matrix of 125 canid species \cite[9 extant and \ncanidfossils{} fossil;][]{Slater2015} with 122 characters (ranging from binary to k=5).
+%}%
 The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini).
-\alexei{We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms). }
+%\alexei{%
+We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms). 
+%}%
 
@@
 \includegraphics[width=15cm]{canids/stratRanges/1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
-\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, \gslater{showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}}
+\caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, 
+%\gslater{%
+showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}
+%}%
 \end{figure}
@@
 There are a diverse array of potential applications for this methodology. 
-The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, \alexei{or because the fossil was not associated with a horizon of known age e.g.
-\cite{Berger2015},} or because of a complete lack of provenance data (e.g.
+The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, 
+%\alexei{%
+or because the fossil was not associated with a horizon of known age e.g. \cite{Berger2015},
+%}%
+or because of a complete lack of provenance data (e.g.
 a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}).
@@
 
-\gslater{One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
+%\gslater{%
+One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
 On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
@@
 We leave these considerations for future work.
-}
+%}%
 Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
@@
 
-The authors would like to thank Daniel Ksepka and Graeme T.
-Lloyd for insightful comments and suggestions on early drafts of this manuscript.
-\alexei{We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
-We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.} AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).
+The authors would like to thank Daniel Ksepka and Graeme T. Lloyd for insightful comments and suggestions on early drafts of this manuscript. %\alexei{%
+We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
+We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.
+%}%
+AJD and TS were partially funded by Marsden grant UOA1324 from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz/programme?s/funds/marsden/awards/2013-awards/}).
 AJD was funded by a Rutherford Discovery Fellowship from the Royal Society of New Zealand (\url{http://www.royalsociety.org.nz}).

### 2016-04-27 fossilDating-revision.tex
responses

@@
 Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \cite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \cite{BentonAyala2003}.
-%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \cite{tavare2002using}.
+%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and palaeontological data in primates \cite{tavare2002using}.
 % AJD: The following two sentences are inflammatory to palaeontologists
@@
 %\alexei{%
-We had stratigraphic ranges based on paleontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms). 
+We had stratigraphic ranges based on palaeontological data for all \ncanidfossils{} fossils (Graham Slater, pers. comms). 
 %}%
@@
 We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the palaeontological age range for each fossil.
-In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the palaeontological range.
+In this case $H_1$ is the hypothesis that the true fossil age is within the given palaeontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the palaeontological range.
 %, but within the broader interval of $(0,T_\text{max})$ Mya. 
@@
 
-One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $M_\mathcal{T}$ and record the fraction of sampling times  within a given  paleontological age range.
+One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $M_\mathcal{T}$ and record the fraction of sampling times  within a given palaeontological age range.
 However, such a simulation approach turns out to be very time-consuming, and the procedure below provided a much faster evaluation of $p(H_1|M_\mathcal{T})$.
@@
 Thus, essentially $p(H_1|M_\mathcal{T})=0$  which leads to a huge Bayes Factor.
-This means we always reject $H_2$, but not because we necessarily agree with the paleontological age range, but because our model has no prior weight for the paleontological age range.
+This means we always reject $H_2$, but not because we necessarily agree with the palaeontological age range, but because our model has no prior weight for the palaeontological age range.
 
-Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain very large trees with arbitrary many species close to the present, thus we have most of the sampling times close to the present.
+Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain very large trees with arbitrarily many species close to the present, thus we have most of the sampling times close to the present.
 (ii) If we draw $r$ and $s$ close to 1, then we obtain a very large per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
@@
 The approach was valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
-In general, even if stepping stone approaches are appropriate, we recommend to investigate $P(H_1|M)$ to investigate the prior assumption on the hypothesis to be tested.
+In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior assumption on the hypothesis to be tested is sensible.
 Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.

### 2016-04-28 response-to-reviewers.tex
responses
minor revisions

@@
 
+\response{We thank the reviewer for their positive comments.}
+
 I only recommend minimal revisions to the manuscript.
@@
 
+\response{Corrected.}
+
 Page 3 line 27: reference is not completely within parentheses
 
+\response{Corrected.}
+
 Page 6 line 53: T and large d, we obtain very large trees with arbitrary <--(Typo?) many species
+
+\response{Corrected to arbitrarily}
 
@@
 
+\response{Corrected.}
+
 Page 9: for the species that fell far beyond the regression line, do you have any hypotheses for these discrepancies?
@@
 
+\response{We do not have any hypotheses about that.}
+
 Figure 3: I did not see a reference for this in the text
+
+\response{Corrected.}
 
@@
 
+\response{Corrected.}
+
 Figure 8: I did not see a reference for this in the text
+
+\response{We have added the following sentence:}
+
+\respquote{Figure 8 shows a sample from the posterior distribution of the analysis of the canid data set.
+The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).
+}
 

### 2016-04-28 fossilDating-revision.tex
minor revisions

@@
 The approach was valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
-In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior assumption on the hypothesis to be tested is sensible.
+In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior on the hypothesis to be tested is sensible.
 Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.
@@
 \includegraphics[width=5in]{Figure2.pdf}
-\caption{A plot of the number of non-ambiguous morphological sites for the taxon against the precision of the phylogenetic estimate of fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
+\caption{\label{fig:precisionVsnumberCharacters} A plot of the number of non-ambiguous morphological characters for the taxon against the precision of the phylogenetic estimate of fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
 the precision is 1/variance in the marginal posterior distribution of the age).}
@@
 \begin{figure}
-\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
+\includegraphics[width=12cm]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=6in]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
+\includegraphics[width=12cm]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
@@
 This evidence, along with the previous result that \Mrelaxed{} has a higher marginal likelihood than \Mstrict{} \cite{gavryushkina2015bayesian} suggests that the relaxed model is overall a better fit to the data.
+Under both models, there is a positive correlation between the precision of the age estimate and the number of non-ambiguous characters coded for the fossil taxon (Figure \ref{fig:precisionVsnumberCharacters}).
 
@@
 Only 13 out of \ncanidfossils{} fossils (11\%) don't have the mean stratigraphic age in the 95\% credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
-The median error is $1.56$My, which is even more accurate than the penguin data set.
-This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$) but despite that the individual age estimates have about double the precision (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
-This may well be because the average age of the penguin fossils is considerably.
+The median error is $1.56$My, which is more accurate than the age estimates for the penguin data set.
+This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$) but despite that the individual age estimates are much more precise (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
+This may well be because the average age of the penguin fossils is considerably larger.
+
+Figure \ref{fig:canidTree} shows a sample from the posterior distribution of the analysis of the canid data set.
+The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).
 

### 2016-04-28 fossilDating-revision.tex
math bayesian

@@
 
+%TSrev
+Given the morphological data $D$ and a stratographic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratographic age range.
+The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the morpholocical evolution model are summarized in $\theta$.
+More formally, we sample from,
+$$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
+where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratographic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
+When we replaced the focal  palaeontological age constraints of fossil $i$  by 
+the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.
+
+The fossilized birth-death model is defined by the following parameters $\eta=(T,d,r,s)$:  the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability with which a fossil is observed $s$ (=sampling rate / (extinction rate + sampling rate)).
+  %%%
+
 Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model 
@@
 
+%TSrev
 We  will now derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$.
-In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters.
-Using the parameterization in \cite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
+In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters $\eta=(T,d,r,s)$.
+%Using the parameterization in \cite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
 We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. This probability density will allow us to directly determine $p(H_1|M_\mathcal{T})$.
+%%
 

### 2016-04-28 response-to-reviewers.tex
answering reviewer 1 detailed points

@@
 
+\response{We have now provided a general form of the posterior we are sampling under the section on ``Phylogenetic analyses'' and clarified details of the stepwise analysis of the fossils.}
+
 Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.
 
+\response{$T$ is not the age of the root, it is the upper limit on the age of the root.}
+
 Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).
+
+\response{We have added the prior information for the morphological substitution rate and log standard deviation of the relaxed clock.}
 
@@
 (?)
+
+\response{The ages of the fossil tips are indeed unknown parameters estimated by the MCMC.
+Their prior distribution emerges from the marginal distribution of the fossilized birth death prior of tip-timed trees, which can be sampled using MCMC, but which we don't have an analytical solution for.}
 

### 2016-04-28 fossilDating-revision.tex
answering reviewer 1 detailed points

@@
 %TSrev
-Given the morphological data $D$ and a stratographic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratographic age range.
-The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the morpholocical evolution model are summarized in $\theta$.
+Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range.
+The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$.
 More formally, we sample from,
 $$P[\mathcal{T},\eta, \theta | D, a] = P[D|\mathcal{T},\theta] P[\mathcal{T},a|\eta] P[\eta] P[\theta]/P[D,a],$$
-where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratographic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
+where $ P[\mathcal{T},a|\eta] =  P[\mathcal{T}|\eta] $ if each fossil age is within its stratigraphic age range specified in $a$, and $ P[\mathcal{T},a|\eta] =0$ else.
 When we replaced the focal  palaeontological age constraints of fossil $i$  by 
@@
 We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}.
-\Mstrict{} assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
+\Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites.
 Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
+Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates.
+We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
 

### 2016-04-28 response-to-reviewers.tex
responses

@@
 Provide a better description of the stepwise analysis of fossil ages.
+
+\response{We have clarified details of the stepwise analysis of the fossils in the new ``Phylogenetic analyses'' section.}
+
 My impression (but I am not totally sure) is that in one analysis, fossil ages are unknown parameters and have uniform prior distributions set according to the stratigraphic bounds.
+
+\response{No, they have prior distributions set according to the fossilized birth-death prior, conditioned on the stratigraphic bounds. This is detailed in the new ``Phylogenetic analyses'' section. }
+
 Then, in stepwise analysis, each fossil age is in turn represented not by the stratigraphic uniform distribution, but by a prior density derived from the fossilised birth-death process.
+
+\response{No, each fossil is in turn analyzed using the prior density derived from the fossilised birth-death process without conditioning on its stratigraphic bounds, while the remaining fossils are conditioned on their stratigraphic bounds.
+This is detailed in the new ``Phylogenetic analyses'' section.
+}
+
  Give some details please.
+
+ \response{We have added the opening three paragraphs in the new ``Phylogenetic analyses'' section that give the details.}
+
   I was trying to work out the general form of the posterior distribution, I was guessing something like:
@@
 
-\response{We have now provided a general form of the posterior we are sampling under the section on ``Phylogenetic analyses'' and clarified details of the stepwise analysis of the fossils.}
+\response{We have now provided a general form of the posterior we are sampling under the new section on ``Phylogenetic analyses''. }
 
@@
 
-\response{$T$ is not the age of the root, it is the upper limit on the age of the root.}
+\response{$T$ is not the age of the root, it is the origin of the fossilized birth-death process, and is older than the root.}
 
@@
 
-p.9 Results section.
-For both data sets the authors state that the relaxed clock model fits the data better, but I can't find the BF for this.
+p.9 Results section. For both data sets the authors state that the relaxed clock model fits the data better, but I can't find the BF for this.
+
+\response{We didn't provide BF. For the penguin data set the BF was provided in \cite{gavryushkina2015bayesian}.}
+
 Also, for both data sets, the authors have sections titled "Penguins conform well to a strict morphological clcok" and "Canids conform well to a morphological clock", but these headings seem to contradict the statement in lines 48-49 p.12 of discussion.
 I am also puzzled as the authors have not quantified here departure from the clock.
+
+\response{We mean that the data sets conform to a relaxed morphological clock, not a strict morphological clock.
+A relaxed morphological clock is still a morphological clock, just with greater variance.}
 
@@
 Then, when the authors discuss whether the data fit or not the strict clock, they should refer to the estimates in the table.
+
+\response{
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrr}
+  \hline
+Parameter & Penguins Mk-1 & Penguins Mk-8 & Canids Mk-1 \\ 
+  \hline
+$t_{MRCA}$ & & & \\
+morph. clock rate & & & \\
+log std. dev. rate & & & \\
+\hline
+  \end{tabular}
+\caption{Summary of key parameters for the three main analyses.} 
+\label{fossilTable1}
+\end{table}
+}
 

### 2016-04-28 fossilDating-revision.tex
morphological clocks

@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.
+
+\subsection*{Morphological clock}
+
 We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
+
+%AJD revision
+There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1949, Gingerich1981}.
+Phenotypic evolution is typically characterized as being less clock-like than molecular evolution \cite{Kimura}.
+While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, this is also true for molecular rates.
+We are not aware of a systematic comparison of variation in rates at the phenotypic and molecular levels.
+Regardless, variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock, as we used in model \Mrelaxed{}.
+In the penguin data set the log standard deviation of the relaxed morphological clock was comparable to values obtained for molecular clocks.
+%%%
+
+%http://earth.geology.yale.edu/~ajs/1993/11.1993.17Gingerich.pdf
+
 We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 

### 2016-04-28 response-to-reviewers.tex
canid log files and morphological clock intro

@@
 
+\response{We have indeed read some of this early literature and we thank the reviewer's encouragement to reference this earlier work on rates of phenotypic evolution.
+We have added a section entitled ``Rates of evolution and relaxed morphological clocks'':}
+
+\respquote{In order to use discrete morphological comparative data to estimate fossil ages, it is necessary to assume a (relaxed) morphological clock.
+There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1944,Haldane1949,Simpson1953,Gingerich1983,Gingerich1993}, going at least back to Darwin's Origin of Species \cite{Darwin1859}.
+Darwin noted that {\it ``Species of different genera and classes have not changed at the same rate''} and illustrated this point with examples of ``living fossils'' such as the Silurian mollusc {\it Lingula} \cite{Darwin1859}.
+However in the same chapter Darwin goes on to say {\it ``In members of the same class the average amount of change, during long and equal periods of time, may, perhaps, be nearly the same''}.
+Nevertheless, phenotypic evolution has more typically been characterized as not evolving in a clock-like manner, especially when compared to molecular evolution \cite{Kimura1983}.
+While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, we would argue that this is also true for molecular rates.
+We are not aware of a comprehensive and systematic comparison of variation in evolutionary rates at the phenotypic and molecular levels.
+Regardless, for the data sets that we analyze, we adopt the point of view that variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock.
+}
+
 I am troubled about the sensitivity of the BF to the prior in this case.

### 2016-04-28 fossilDating-revision.tex
neutral evolution url
canid log files and morphological clock intro

--- 2016-04-28 fossilDating-revision.tex
@@
 This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}.
+
 It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
@@
 19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.
+
+%AJD revision
+\subsection*{Rates of evolution and relaxed morphological clocks}
+
+In order to use discrete morphological comparative data to estimate fossil ages, it is necessary to assume a (relaxed) morphological clock.
+There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1944,Haldane1949,Simpson1953,Gingerich1983,Gingerich1993}, going at least back to Darwin's Origin of Species \cite{Darwin1859}.
+Darwin noted that {\it ``Species of different genera and classes have not changed at the same rate''} and illustrated this point with examples of ``living fossils'' such as the Silurian mollusc {\it Lingula} \cite{Darwin1859}.
+However in the same chapter Darwin goes on to say {\it ``In members of the same class the average amount of change, during long and equal periods of time, may, perhaps, be nearly the same''}.
+Nevertheless, phenotypic evolution has more typically been characterized as not evolving in a clock-like manner, especially when compared to molecular evolution \cite{Kimura1983}.
+While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, we would argue that this is also true for molecular rates.
+We are not aware of a comprehensive and systematic comparison of variation in evolutionary rates at the phenotypic and molecular levels.
+Regardless, for the data sets that we analyze, we adopt the point of view that variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock.
+
+%http://earth.geology.yale.edu/~ajs/1993/11.1993.17Gingerich.pdf
+%https://books.google.co.nz/books?id=olIoSumPevYC&printsec=frontcover#v=onepage&q&f=false
+
+%%%
 
@@
 Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
-Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates.
+Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates.
 We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
@@
 
-\subsection*{Penguins conform well to a strict morphological clock}
+\subsection*{Penguins conform well to a morphological clock}
 Although \Mstrict{} is a very simple model, the phylogenetic estimates of the ages of the penguin fossils were remarkably consistent with their palaeontological age ranges. 
@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.
-We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.
+
+%AJD
+We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of penguin fossil ages is improved by a model that includes a variation in rates of morphological evolution among lineages.
+However, in the penguin data set the variation in evolutionary rates was not extreme and the log standard deviation of the relaxed morphological clock was comparable to values obtained for molecular clocks.
+%%%
+
 We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 

### 2016-04-29 response-to-reviewers.tex
Addressing Mario's minor comments

@@
 \response{
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrr}
-  \hline
-Parameter & Penguins Mk-1 & Penguins Mk-8 & Canids Mk-1 \\ 
-  \hline
-$t_{MRCA}$ & & & \\
-morph. clock rate & & & \\
-log std. dev. rate & & & \\
-\hline
-  \end{tabular}
-\caption{Summary of key parameters for the three main analyses.} 
-\label{fossilTable1}
-\end{table}
+We have added a table with the posterior estimates for root age and evolutionary rates and variation.
 }
@@
 
+\response{Corrected.}
+
 I think Figures 4 and 5 should be merged into one figure.
+
+\response{We have left these as two separate figures.}
 
@@
 
+\response{Corrected. Both figures refer to the penguin analyses.}
+
 Figure 6, panels (b), (c) and (d) seem empty.
+
+\response{We believe this was a problem created by the manuscript submission process. The panels are not empty in the original PDF that was submitted.}
 
@@
 
+\response{This figure was completely re-generated as we added stratigraphic ranges to the canid analysis.}
+
 Figure 8: Please replace this with the consensus tree and provide 95\% CIs of internal node ages.
 
+\response{We are happy with a representative tree from the posterior distribution.
+It is not trivial to produce a consensus tree from a sampled-ancestor analysis, because the posterior has random dimension, due to a variable number of sampled ancestors.
+This is an open problem that must be addressed in a later publication.}
+
 I could not find references to Figures 3 and 8 in the text.
+
+\response{Corrected.}
 
@@
 
+\response{We feel that figures 2, 3, 4, 5 and 7, which together plot precision graphically in a number of ways, are more than adequate to clearly show the level of precision in the estimates.}
+
 Considering the point above (and the large CIs observed in figures 2 and 7), it is clear that estimates are not precise (note, I refer to accuracy as bias of estimates and precision as 1/variance of estimates).
@@
 Estimates do seem to be accurate, as the posterior means are close to stratigraphic mid-points, but the authors must stress, in the abstract, results and discussion, that uncertainties are large.
+
+\response{There are no statements in the revised manuscript that the estimates are precise.
+We state in the abstract that precision is low, and elsewhere we plot the precision and discuss that fact that estimates are more precise in the canid analysis than the penguin analysis.}
 
@@
 
+\response{The age of extant penguins in the subject of \cite{gavryushkina2015bayesian}, so we don't discuss it here, however, the estimate obtained is much younger than previous published estimates using node-dating, contrary to the reviewer's suggestion that morphological data tends to produce ancient estimates.}
+
 I think the insistence by the authors to use the term "morhpological clock" will cause controversy.
@@
 Otherwise I suggest removing the term.
+
+\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis, which is not much different to estimate of the same parameter for relaxed molecular clocks in, for example, influenza.}
 
@@
 Can the authors give some clear, easy to follow, recommendations for biologists that may wish to use the method in their own works?
+
+\response{Unfortunately we can't. This issue will need further attention in future work.}
+
 Another issue is that when the model is changed, the BF change.

### 2016-04-29 fossilDating-revision.tex
Addressing Mario's minor comments

@@
 \includegraphics[width=5in]{Figure2.pdf}
-\caption{\label{fig:precisionVsnumberCharacters} A plot of the number of non-ambiguous morphological characters for the taxon against the precision of the phylogenetic estimate of fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
+\caption{\label{fig:precisionVsnumberCharacters} A plot of the number of non-ambiguous morphological characters for the penguin taxon against the precision of the phylogenetic estimate of corresponding fossil age for (a) \Mstrict{} and (b) \Mrelaxed{} (i.e.
 the precision is 1/variance in the marginal posterior distribution of the age).}
@@
 \caption{\label{fig:compareM1M8}
-Comparison of \Mstrict{} against \Mrelaxed{}.
+Comparison of \Mstrict{} against \Mrelaxed{} analyses for the penguin data set.
 (a) Phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{} with $x=y$ line (blue), (b) Regression of error in phylogenetic estimate of fossil age of \Mstrict{} against \Mrelaxed{}, (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.}
@@
 
+Table \ref{estimatesTable} shows that the rate of morphological evolution in canids is faster than that estimated in the penguins, however this could be a simple reflection of the shorter geological time scale (and shorter average branch lengths) over which the rate has been estimated \cite{Gingerich1993}.
+
 \begin{figure}
@@
 
+\begin{table}[ht]
+\centering
+\begin{tabular}{c|rrr}
+  \hline
+Analysis & $t_{MRCA}$ & morph. rate  & log sd rate (S) \\ 
+ & (Myr) & ($\times 10^{-2}$/Myr) & \\
+  \hline
+Penguins Mk-1 & 61.7 [60.5, 63.8] & 1.80 [1.55, 2.07] & -  \\
+Penguins Mk-8 & 61.4 [60.5, 63.1] & 1.29 [0.76, 1.86] & 0.69 [0.41, 0.99] \\
+  \hline
+Canids Mk-1 &  36.8 [35.4, 38.5] & 2.83 [2.47, 3.19] & - \\
+\hline
+  \end{tabular}
+\caption{Summary of key parameters for the three main analyses.
+Note: the $t_{MRCA}$ is the time of the most recent common ancestor of all taxa, including both extinct and extant species.
+The 95\% HPD interval for each estimate is in square brackets.
+The morphological clock rate is given in percent change per million years.}
+\label{estimatesTable}
+\end{table}
+
 \section*{Discussion}
@@
 We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of penguin fossil ages is improved by a model that includes a variation in rates of morphological evolution among lineages.
-However, in the penguin data set the variation in evolutionary rates was not extreme and the log standard deviation of the relaxed morphological clock was comparable to values obtained for molecular clocks.
+However, in the penguin data set the variation in evolutionary rates was not too extreme and the estimated log standard deviation of the relaxed morphological clock ($S = 0.69$; refer to Table \ref{estimatesTable}) is comparable to values obtained for molecular clocks.
 %%%
@@
 We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
-We thank Mario do Reis and one anonymous reviewer for helpful comments during revision.
+We thank Mario do Reis and one anonymous reviewer for very helpful comments during revision.
 %}%

### 2016-04-29 response-to-reviewers.tex
need tanja to address Mario's last point
Tanja revision start

@@
 
-\response{We added a section called Phylogenetic analyses with following text:}
+\response{We added a section called Phylogenetic analyses, explaining the stepwise analysis of fossil age as well as the morphological models M1 and M8}
 
-\respquote{Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
+%\respquote{Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
 As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered.
-We applied two phylogenetic models to the penguin data set: Mk-1 and Mk-8.
+%We applied two phylogenetic models to the penguin data set: Mk-1 and Mk-8.
 Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites.
@@
 
- \response{We have added the opening three paragraphs in the new ``Phylogenetic analyses'' section that give the details.}
+ \response{We have added the opening two paragraphs in the new ``Phylogenetic analyses'' section that give the details.}
 
@@
 
-\response{$T$ is not the age of the root, it is the origin of the fossilized birth-death process, and is older than the root.}
+\response{$T$ is not the age of the root, it is the origin of the fossilized birth-death process, and is older than the root. It can be interpreted as the stem age of the clade.}
 
@@
 Or is this a general problem with any implementation of stepping-stones and this model?
+
+\response{TODO - thinking about this: why can we not use stepping stone?
+We could have calulated the marginal likelihood given the focal fossil is in the paleontological range, and the marginal likelihood given the fossil is outside the range, no?}
+
 I am also confused, do sampling times refer to the ages of fossil tips?
@@
 
-\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis, which is not much different to estimate of the same parameter for relaxed molecular clocks in, for example, influenza.}
+\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis (see Table 3), which is not much different to estimate of the same parameter for relaxed molecular clocks in, for example, influenza.}
 
@@
 We have added a section entitled ``Rates of evolution and relaxed morphological clocks'':}
+
+%TODO from here
 
@@
 For example, if a lineage has shown explosively fast rates of evolution, then I would expect the Bayesian phylogenetic estimates of the fossil age to contradict the stratigraphic range if a strict clock is imposed.
+
+\response{*** TANJA - can you try to address this one?}
 

### 2016-04-29 fossilDating-revision.tex
Tanja revision start

@@
 %TSrev
-Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range.
+Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range (i.e.
+the phylogenetic estimate of fossil age).
 The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$.

### 2016-04-30 fossilDating-revision.tex

--- 2016-04-29 fossilDating-revision.tex
@@
 
-%TS: I suggest to remove the following heading
-%\subsection*{Phylogenetic tip-dating}
-
 The phylogenetic estimation of the age of a taxon based on its molecular sequence has been previously described \cite{drummond2002computational,shapiro2011bayesian} and applied to both ancient subfossil remains and rapidly evolving viral taxa.
@@
 We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters.
-This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}.
+This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian} and \cite{zhang2016}.
 
-It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
-This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa.
-A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
-We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
-The method is applied to two rich and well-characterized morphological data sets: (i) 
-%TS how many extant penguins?
-19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.
-
+%TSrev I moved this section up
 %AJD revision
-\subsection*{Rates of evolution and relaxed morphological clocks}
-
+%\subsection*{Rates of evolution and relaxed morphological clocks}
 In order to use discrete morphological comparative data to estimate fossil ages, it is necessary to assume a (relaxed) morphological clock.
@@
 
+Our approach is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available.
+This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa.
+A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
+We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
+The method is applied to two rich and well-characterized morphological data sets: (i) 
+%TS how many extant penguins?
+19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.
+
 %http://earth.geology.yale.edu/~ajs/1993/11.1993.17Gingerich.pdf
@@
 \section*{Methods}
-
-Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach \cite[implemented in BEAST2;][]{Beast2} for phylogenetic estimation of time-trees that employs both fossil and molecular sequence data as equal partners under the rule of probability.
-We extend their work further by investigating the consistency between the phylogenetic estimate of the age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
+%TSrev
+Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach implemented in BEAST2 \citep{Beast2} for phylogenetic estimation of time-trees that employs both morphological data from fossils and extant data and molecular sequence data as equal partners under the rule of probability for estimating a time-tree.
+An equivalent method \citep{zhang2016} is introduced within MrBayes \cite{ronquist2012mrbayes}.
 The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.
 
-A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree. 
-Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. 
+We extend the approach in \cite{gavryushkina2015bayesian}  further by investigating the consistency between the phylogenetic estimate of the age of a fossil and the corresponding fossil age range determined by geological and biostratigraphic evidence.
+%Previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree \cite{gavryushkina2015bayesian,zhang2016}. 
+%Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. 
 This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree.
@@
 The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
-The fossil ages ranged from 
+The fossil ages ranged from TODO.
+As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
 See \cite{gavryushkina2015bayesian} for further details of data selection.
@@
 For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by 
-the fossilized birth-death process prior as described below.  
-For the choices of morphological substitution model and tree prior we followed \cite{gavryushkina2015bayesian} and considered two of her modelling combinations. 
-The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \cite{Lewis2001}.
-We also employed \Mrelaxed{} from \cite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \cite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}.
+the fossilized birth-death process prior, and thus we obtained a phylogenetic estimate of the fossil's age.
 
@@
 
-We applied the \Mstrict{} model to this data and assumed a uniform prior on each fossil's age within their respective stratigraphic ranges.
+As with the penguin data set we  performed an analysis for each of the \ncanidfossils{} canid fossils in turn. 
  Unlike the original study \cite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the fossils.
-As with the penguin data set we also performed an analysis for each of the \ncanidfossils{} canid fossils in turn.
-We used the same priors on macroevolutionary parameters as employed by \cite{gavryushkina2015bayesian} in their \Mstrict{} model except with an upper limit on the origin of $T_\text{max} = 80$.
-To phylogenetically estimate the age of the focal fossil we sampled its age under the fossilized birth-death prior in the range $(0,T_\text{max})$ Mya.
-%TS: which priors on d, r, s, T did we use now?
 
@@
 %TSrev
-Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees connecting the fossils, with each fossil having a specified age within its stratigraphic age range.
+Given the morphological data $D$ and a stratigraphic age range for each fossil $a = ((l_1,u_1),(l_2, u_2),\ldots,(l_n,u_n))$ (with $l_i$ being the lower age bound for fossil $i$, and $u_i$ being the upper age bound for fossil $i$), we sample phylogenetic trees with the fossils being tips or sampled ancestors, and each fossil having a specified age  in the phylogenetic tree within its stratigraphic age range.
 The parameters of the fossilized birth-death model are summarized in $\eta$, and the parameters  of the model for morphological character evolution are summarized in $\theta$.
@@
 When we replaced the focal  palaeontological age constraints of fossil $i$  by 
-the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$.
+the fossilized birth-death process prior, we simply set $l_i=0$ and $u_i=T$ (where 0 is present time and $T$ is the total height of the tree) to estimate the phylogenetic age of the focal fossil.
 
-The fossilized birth-death model is defined by the following parameters $\eta=(T,d,r,s)$:  the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability with which a fossil is observed $s$ (=sampling rate / (extinction rate + sampling rate)).
+The fossilized birth-death model is defined by the following parameters $\eta=(T,d,r,s)$:  the time of the start of the process $T$ prior to present time 0, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability with which a fossil is observed $s$ (=sampling rate / (extinction rate + sampling rate)).
   %%%
@@
 \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states.
-As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered.
 
-We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}.
-\Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites.
-Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
-Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates.
+%TSrev
+We applied two phylogenetic models to the penguin data set, \Mstrict{} and \Mrelaxed{}, and we applied \Mstrict{} to the canid dataset.
+\Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites  \cite{Lewis2001}.
+Model \Mrelaxed{} \citep{gavryushkina2015bayesian} partitioned the alignment into  partitions (six for the penguins), with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
+Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates, and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}.
+%TODO how was this additional parameter set?
 We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
+%TSref TODO Alexei please check if the following sentence is correct.
+For the parameters of the fossilized birth-death model tree prior, 
+we specify the prior below in the section ``Computing the phylogenetic evidence for an age range''.
+%we  also followed \cite{gavryushkina2015bayesian}, except with an upper limit on the origin of $T= 80$ for the canids.
 
@@
 In this case $H_1$ is the hypothesis that the true fossil age is within the given palaeontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the palaeontological range.
+A BF $>>$ 1 indicates strong support for $H_1$, given the model $M$ is appropriate for the considered data.
+
 %, but within the broader interval of $(0,T_\text{max})$ Mya. 
@@
 %TSrev
-We  will now derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$.
-In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters $\eta=(T,d,r,s)$.
+We   derive some analytic results for evaluating $p(H_1|M_\mathcal{T})$. 
+The model $M_\mathcal{T}$ is the fossilized birth-death process with priors on its parameters $\eta=(T,d,r,s)$. 
 %Using the parameterization in \cite{gavryushkina2015bayesian}, the parameters are the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
@@
 We want to note that the stepping stone sampling approach \cite{Xie2011}
-%TS which paper to cite here?
- to calculate Bayes factors would not have been possible in our case:  In stepping stone sampling the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
+ to calculate Bayes factors would not have been directly applicable in our case:  In stepping stone, sampling the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
 However, using a birth-death model, the sampling times are part of the data.
-The approach was valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
+The approach is valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
+It is not clear if the stepping stone approach can be directly applied for models with number of tips being part of the data.
 In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior on the hypothesis to be tested is sensible.
@@
 The median error is $1.56$My, which is more accurate than the age estimates for the penguin data set.
-This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$) but despite that the individual age estimates are much more precise (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
+This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$), nevertheless the individual age estimates are much more precise (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
 This may well be because the average age of the penguin fossils is considerably larger.
@@
 The Bayesian phylogenetic estimate of fossil age (median and 95\% credible interval of marginal posterior) for each of the \ncanidfossils{} canid fossils plotted against their stratigraphic age ranges, under a strict morphological clock model \Mstrict{}.
-The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. Blue line shows the $x=y$. 
+%The Bayesian estimates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval. 
+The palaeontological age estimates are represented by the mid-point of the range and the upper and lower limits.
+The Bayesian esti- mates are represented by the median of the marginal posterior distribution and the upper and lower limits of the 95\% HPD interval.
+Blue line shows the $x=y$. 
 If the vertical line doesn't cross $x=y$, then the mean of the stratigraphic age range is not in the credible interval of the phylogenetic estimate of fossil age.
@@
 In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.
+In all cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
 
@@
 However, in the penguin data set the variation in evolutionary rates was not too extreme and the estimated log standard deviation of the relaxed morphological clock ($S = 0.69$; refer to Table \ref{estimatesTable}) is comparable to values obtained for molecular clocks.
+The median error in age estimates for the two data sets investigated were 2 My and 1.6 My respectively, using either a very simple or more complex models of discrete morphological change. 
 %%%
@@
 However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.
+%TODO: well, we could simply calculate the relative error and compare!?
+
+%\gslater{%
+One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
+On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
+For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5\text{myr}$ (pers. comm. Graham Slater). 
+In this context it is interesting to note the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges.
+This raises the question of whether multiple specimens of a single species that span a significant time frame and/or different localities should be coded as separate taxa as input for the fossilized birth-death method.
+Even if not coded as separate taxa it may be possible to extend the method used here to explicitly account for multiple specimens and associated ages when a fossil species is represented by more than one fossil.
+We leave these considerations for future work.
+%}%
+Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
 
@@
 It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 
-The median error in age estimates for the two data sets investigated were 2 My and 1.6 My respectively, using either a very simple or more complex models of discrete morphological change. 
-In all cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
-
-%\gslater{%
-One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
-On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
-For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5\text{myr}$ (pers. comm. Graham Slater). 
-In this context it is interesting to note the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges.
-This raises the question of whether multiple specimens of a single species that span a significant time frame and/or different localities should be coded as separate taxa.
-Even if not coded as separate taxa it may be possible to extend the method used here to explicitly account for multiple specimens and associated ages when a fossil species is represented by more than one fossil.
-We leave these considerations for future work.
-%}%
-Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
-
-In the mean time we anticipate that this approach will help to promote the application of a consistent probabilistic framework to consider both molecular and fossil evidence.
+Overall, we anticipate that this approach will help to promote the application of a consistent probabilistic framework to consider both molecular and fossil evidence.
 Our results are encouraging in suggesting that the statistical models presented are adequate for inference of phylogenetic time-trees from morphological fossil data.

### 2016-04-30 response-to-reviewers.tex

@@
 
-\response{TODO - thinking about this: why can we not use stepping stone?
-We could have calulated the marginal likelihood given the focal fossil is in the paleontological range, and the marginal likelihood given the fossil is outside the range, no?}
+\response{One can calculate BF either in the way described here or via stepping stone approaches.
+Stepping stone approaches have been used for the coalescent, where the number of tips in a tree is fixed.
+Here, the number of trees is treated as part of the data, and while a stepping stone approach may be still appropriate, we did not test it for this scenario.
+This point is highlighted in ``We want to note that the stepping stone sampling approach \cite{Xie2011}
+ to calculate Bayes factors would not have been directly applicable in our case:  In stepping stone sampling the $D_\mathcal{T}$ is treated as part of the model, not part of the data.
+However, using a birth-death model, the sampling times are part of the data.
+The approach is valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
+It is not clear if the stepping stone approach can be directly applied for models with number of tips being part of the data.
+In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior on the hypothesis to be tested is sensible.
+Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.
+'' }
 
@@
 \response{We have indeed read some of this early literature and we thank the reviewer's encouragement to reference this earlier work on rates of phenotypic evolution.
-We have added a section entitled ``Rates of evolution and relaxed morphological clocks'':}
-
-%TODO from here
+We have added a section in the introduction.}
 
@@
 
-\response{*** TANJA - can you try to address this one?}
+\response{We added: A BF $>>$ 1 indicates strong support for $H_1$, given the model $M$ is appropriate for the considered data.}
 
@@
 I'm not suggesting more analyses, but perhaps further discussion of these issues.
+
+\response{We added a paragraph on the appropriateness of a morphological clock: paragraph 1 in ``Rates of evolution and relaxed morphological clocks
+''}
 

### 2016-04-30 fossilDating-revision.tex

@@
 An equivalent method \citep{zhang2016} is introduced within MrBayes \cite{ronquist2012mrbayes}.
-The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.
+The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{gavr2014,Gavr2013}.
 
@@
 The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
-The fossil ages ranged from TODO.
+The fossil ages ranged from 4.5 to 61.3 Myr.
 As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
@@
 %TSref TODO Alexei please check if the following sentence is correct.
-For the parameters of the fossilized birth-death model tree prior, 
-we specify the prior below in the section ``Computing the phylogenetic evidence for an age range''.
-%we  also followed \cite{gavryushkina2015bayesian}, except with an upper limit on the origin of $T= 80$ for the canids.
+For the penguin analyses, the parameters of the fossilized birth-death model tree prior  were specified as described in the section ``Computing the phylogenetic evidence for an age range''. 
+For the canid analysis we used a more standard parametrisation: origin $T$ with Uniform(0,120) prior, diversification rate $d$ with LogNormal(-3.5, 1.5), turnover and sampling proportion with Uniform(0,1) priors.
 

### 2016-05-01 response-to-reviewers.tex
final todos and proof

@@
 \response{There are no statements in the revised manuscript that the estimates are precise.
-We state in the abstract that precision is low, and elsewhere we plot the precision and discuss that fact that estimates are more precise in the canid analysis than the penguin analysis.}
+We quantify the relative precision in the abstract, and elsewhere we plot the precision and discuss that fact that estimates are relatively more precise in the canid analysis than the penguin analysis.}
 
@@
 
-\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis (see Table 3), which is not much different to estimate of the same parameter for relaxed molecular clocks in, for example, influenza.}
+\response{The estimated log sd of the evolutionary rate is 0.69 in the penguin Mk-8 analysis (see Table 3), which is not much different to estimates of the same parameter for relaxed molecular clocks in, for example, influenza.}
 

### 2016-05-01 fossilDating-revision.tex
final todos and proof

@@
 In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa.
-We exploit this statistical framework to investigate the internal consistency of these models by producing phylogenetic estimates of the age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
-We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age, although the credible intervals were often wide, reflecting low precision. 
+We exploit this statistical framework to investigate the internal consistency of these models by producing phylogenetic estimates of the age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species (penguins and canids).
+We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age and a median relative error in the two data sets of 5.7\% and 13.2\% respectively.
+The median relative standard error (RSD) was 9.2\% and 7.2\% suggesting good precision, although with some outliers.
 In fact in the two data sets we analyze the phylogenetic estimates of fossil age is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
-The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
-We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', 
+The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is an adequate fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
+We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, further testing of the ``morphological clock'',
 and for uncovering potential model misspecification and/or data errors when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
@@
 We refer to this as the phylogenetic estimate of the fossil's age. 
-In phylogenetically estimating the age of each of the fossils in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil.
+In phylogenetically estimating the age of each of the fossils in turn, two questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, and (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil?
 These two questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \cite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of \ncanidfossils{} canid fossils and their extant relatives \cite{Slater2015}.
@@
 The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
-The fossil ages ranged from 4.5 to 61.3 Myr.
+The fossil age intervals had median values ranging from 5.55 to 61.05 Myr.
 As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
@@
 
-The fossilized birth-death model is defined by the following parameters $\eta=(T,d,r,s)$:  the time of the start of the process $T$ prior to present time 0, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability with which a fossil is observed $s$ (=sampling rate / (extinction rate + sampling rate)).
+The fossilized birth-death model is defined by the following parameters $\eta=(T,d,r,s)$:  the time of the start of the process $T$ prior to present time 0, the net diversification rate ($d$ = speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability with which a fossil is observed $s$ (=sampling rate / (extinction rate + sampling rate)).
   %%%
@@
 Model \Mrelaxed{} \citep{gavryushkina2015bayesian} partitioned the alignment into  partitions (six for the penguins), with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.
-Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates, and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}.
-%TODO how was this additional parameter set?
-We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
-%TSref TODO Alexei please check if the following sentence is correct.
+Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates, and an additional parameter $\alpha$ governing the shape for gamma-distributed rate variation across sites \cite{yang:1994ma}.
+The prior distribution for $\alpha$ was uniform in the interval $(0,10)$.
+
+We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal($M=-5.5, S=2$) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396, \beta=0.3819$) prior on $S$ for the relaxed clock analyses.
 For the penguin analyses, the parameters of the fossilized birth-death model tree prior  were specified as described in the section ``Computing the phylogenetic evidence for an age range''. 
-For the canid analysis we used a more standard parametrisation: origin $T$ with Uniform(0,120) prior, diversification rate $d$ with LogNormal(-3.5, 1.5), turnover and sampling proportion with Uniform(0,1) priors.
+For the canid analysis we used the standard parametrisation $\eta=(T,d,r,s)$, with the following priors: uniform prior from in the interval $(0, 120)$ million years for origin $T$, LogNormal($M=-3.5, S=1.5$) prior for diversification rate $d$, unit uniform prior $(0,1)$ for turnover $r$ and sampling proportion $s$.
 
@@
 One way to determine $p(H_1|M_\mathcal{T})$ would be to simulate trees under the model $M_\mathcal{T}$ and record the fraction of sampling times  within a given palaeontological age range.
-However, such a simulation approach turns out to be very time-consuming, and the procedure below provided a much faster evaluation of $p(H_1|M_\mathcal{T})$.
+However, such a simulation approach turns out to be very time-consuming, and the procedure below provides a much faster evaluation of $p(H_1|M_\mathcal{T})$.
 
@@
 $$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
-with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
-The equation above calculates the required probability: The left term conditions on survival of the process ($1-p_0(T;d,r)$).
+with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probability of a single lineage producing $i$ surviving lineages at time $t$.
+The equation above calculates the required probability and the left term in that probability conditions on survival of the process ($1-p_0(T;d,r)$).
 Then we calculate the probability to have $k$ lineages at time $t$ before the present $( p_k(T-t;d,r))$, multiply by the sampling rate $k \psi$, and weight by the probability that at least one lineage of the $k$ lineages survives to the present $ (1-p_0(t;d,r)^k)$.
@@
 
-Inspection of our prior identifies two problems: (i) If we draw a large   $T$ and large $d$, we obtain very large trees with arbitrarily many species close to the present, thus we have most of the sampling times close to the present.
-(ii) If we draw $r$ and $s$ close to 1, then we obtain a very large per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
+Inspection of our prior identifies two problems: (i) If we draw a large $T$ and large $d$, we obtain very large trees with arbitrarily many species close to the present, thus we have most of the sampling times close to the present, (ii) If we draw $r$ and $s$ close to 1, then we obtain a very large per-lineage  sampling rate $\psi=\frac{s}{1-s} \frac{rd}{1-r}$.
 Thus these parameter combinations govern the probability density curve and cause again most prior weight to be close to the present.
 
-We thus assumed new prior distributions.
-The net diversification rate $d: lognorm(-3.5,0.5)$ was chosen with a  smaller standard deviation which avoids  too much weight on very fast growing trees.
-The turnover $r: Unif(0,1)$ was set as before.
+We therefore assumed new prior distributions.
+The net diversification rate $d: LogNormal($M=-3.5,S=0.5$)$ was chosen with a smaller standard deviation which avoids too much weight on very fast growing trees.
+The turnover $r: Uniform(0,1)$ was set as before.
 
-For $s$ we assume an implicit prior: we assume $lognorm(-2,1)$ for $\psi$, and $$s=\psi/(\mu+\psi)$$ (with extinction rate $\mu = r d /(1-r)$). This avoids very high sampling rates.
+For $s$ we assume an implicit prior: we assume $LogNormal(-2,1)$ for $\psi$, and $$s=\psi/(\mu+\psi)$$ (with extinction rate $\mu = r d /(1-r)$). This avoids very high sampling rates.
 
@@
 However, using a birth-death model, the sampling times are part of the data.
-The approach is valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data).
+The approach is valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model assumptions) rather than being modelled (and thus are a realisation of the model which means they are data).
 It is not clear if the stepping stone approach can be directly applied for models with number of tips being part of the data.
@@
 Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic estimates of fossil age. The points in this plot have $R^2 = 0.903$. 
-The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
+The median error (where the {\it error} is the difference between the phylogenetic median and the geological median) is 1.96 Myr.
+The median relative error (where the {\it relative error} is the {\it error} divided by the geological median) was 5.7\% and the median relative standard deviation (RSD; defined as the standard deviation of the marginal posterior divided by the posterior estimate) was of 9.2\%.
+A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.
 
@@
 $\text{log BF} > 3.0$).
-Likewise, if we consider only the posterior probability that the fossil is in the geological age range then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range.
-These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.007, 0.001 and 0.001 respectively.
+Likewise, if we consider only the posterior probability that the fossil is in the geological age range, then three of the 36 fossils has a posterior probability $< 0.05$, suggesting low posterior support for the phylogenetic estimate of fossil age being within the palaeontological age range.
+These three fossils were {\em Madrynornis mirandus}, {\em Paraptenodytes antarcticus} and {\em Sphenicus muizoni} with posterior probabilities that the phylogenetic estimate of fossil age is in the palaeontological range of 0.007, 0.001 and 0.001, respectively.
 All other fossils have posterior probabilities of $> 0.05$ of their age being in the palaeontological range. 
 It is worth noting that the absolute discrepancy in the ages are still quite moderate for the three fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.3Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
+mirandus}: 6.3 Myr vs 10 Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
 antarcticus}: 29.9 vs 22, {\em S.
@@
 Again the absolute discrepancy in the ages are quite moderate for the five fossils with low posterior probabilities, with {\em M.
-mirandus}: 6.7Myr vs 10Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
+mirandus}: 6.7 Myr vs 10 Myr (phylogenetic estimate of fossil age versus palaeontological age), {\em P.
 antarcticus}: 28.0 vs 22, {\em P.
@@
 A summary of all the individual estimates are tabulated in Table \ref{fossilTable8}. 
-The individual marginal posterior distributions of phylogenetic estimates of fossil age under \Mrelaxed{} and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
+The individual marginal posterior distributions of phylogenetic estimates of fossil age under \Mrelaxed{}, and the corresponding geological range are shown in Figures \ref{hist8_younger} and \ref{hist8_older}.
 
@@
 \begin{figure}
-\includegraphics[width=12cm]{8_penguins_eNprior/8_fossilDatingHist_younger.pdf}
+\includegraphics[width=12cm]{8_fossilDatingHist_younger.pdf}
 \caption{\label{hist8_younger}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils younger than 30 Myr using \Mrelaxed{}.
@@
 \begin{figure}
-\includegraphics[width=12cm]{8_penguins_eNprior/8_fossilDatingHist_older.pdf}
+\includegraphics[width=12cm]{8_fossilDatingHist_older.pdf}
 \caption{\label{hist8_older}Marginal posterior density plots for the phylogenetic estimate of fossil age of each of the 18 penguin fossils older than 30 Myr using \Mrelaxed{}.
@@
 
-The canid data set shows remarkably good consistency between stratigraphic age ranges and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
+The canid data set shows remarkable consistency between stratigraphic age ranges and phylogenetic estimates of fossil ages, even with the simple strict morphological clock model (\Mstrict{}).
 The $R^2 = 0.897$ between the phylogenetic and stratigraphic ages ranges (see Figure \ref{fig:canidMorphVsGeo}). 
 Only 13 out of \ncanidfossils{} fossils (11\%) don't have the mean stratigraphic age in the 95\% credible interval of the phylogenetic estimate of fossil age and there are no extreme outliers. 
-The median error is $1.56$My, which is more accurate than the age estimates for the penguin data set.
-This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$), nevertheless the individual age estimates are much more precise (median HPD range = 4.2 My for canids as opposed to 9.6My for penguins).
-This may well be because the average age of the penguin fossils is considerably larger.
+The median error is $1.56$ My, which in absolute terms is more accurate than the age estimates for the penguin data set.
+However the median relative error was 13.2\%, more than twice that for the penguin fossils.
+
+This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$), nevertheless the individual age estimates are more precise (median HPD range = 4.2 My for canids as opposed to 9.6 My for penguins).
+However this is mainly due to the fact that the average age of the penguin fossils is considerably larger and the median relative precision (i.e.
+RSD) was 7.2\%, only slightly better than the value for the penguin fossils of 9.2\%.
 
 Figure \ref{fig:canidTree} shows a sample from the posterior distribution of the analysis of the canid data set.
-The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).
+The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}) The x-axis is in units of Mya.
 
@@
   \hline
-Penguins Mk-1 & 61.7 [60.5, 63.8] & 1.80 [1.55, 2.07] & -  \\
-Penguins Mk-8 & 61.4 [60.5, 63.1] & 1.29 [0.76, 1.86] & 0.69 [0.41, 0.99] \\
+Penguins Mk-1 & 61.7 [60.5, 63.8] & 1.79 [1.53, 2.05] & -  \\
+Penguins Mk-8 & 61.4 [60.5, 63.3] & 1.29 [0.77, 1.90] & 0.69 [0.40, 0.99] \\
   \hline
@@
 
-In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.
+In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set to provide an accurate age of the fossil based on a phylogenetic model.
 In all cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.
@@
 
-We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
-This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates.
-However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.
-%TODO: well, we could simply calculate the relative error and compare!?
+In absolute terms the fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
+One might think that the larger reference set of fossils in the canid data set (115 versus 35) makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates.
+However, since the average age of the canid fossils is considerably younger than that for the penguin fossils, a more appropriate comparison uses relative error and relative precision. 
+By these measures the penguin data set actually provides the more accurate estimates, whereas relative precision is overall slightly better for the canid data set.
+Future work is needed to investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy and precision of the phylogenetic estimate of a fossil's age.
 
 %\gslater{%
-One difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
+Another difference between the two data sets analyzed here is that the penguin fossils were largely single specimens, or at least single localities, so that the age range specified for the fossil represents uncertainty in the geological age of the horizon the fossil was associated with (for example, uncertainty in radiometric dates from the volcanic layers above or below the fossil-carrying horizon and uncertainty about the age difference between the volcanic layers and the horizon the fossil is in).
 On the other hand, most of the canid species were assigned stratigraphic age ranges based on multiple specimens from multiple localities spanning a substantial time range. 
-For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5\text{myr}$ (pers. comm. Graham Slater). 
+For example, there are thousands of specimens of {\it Hesperocyon gregarious} from multiple sites in North America spanning $>5$ Myr (pers. comm. Graham J. Slater). 
 In this context it is interesting to note the canids that fall off the x=y line in Figure 7 are mostly (but not exclusively) taxa represented by singletons and therefore those with relatively short stratigraphic ranges.
@@
 %}%
-Future work may also investigate in a more systematic fashion how the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age.
 
-There are a diverse array of potential applications for this methodology. 
+There are diverse potential applications for this methodology. 
 The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found, 
 %\alexei{%
-or because the fossil was not associated with a horizon of known age e.g. \cite{Berger2015},
+or because the fossil was not associated with a horizon of known age, e.g. \cite{Berger2015},
 %}%
@@
 The authors would like to thank Daniel Ksepka and Graeme T. Lloyd for insightful comments and suggestions on early drafts of this manuscript. %\alexei{%
-We are also very grateful to Graham Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
-We thank Mario do Reis and one anonymous reviewer for very helpful comments during revision.
+We are also very grateful to Graham J.
+Slater for comments on the phylogenetic analysis of canids and for providing stratigraphic age ranges for the canid fossils.
+We thank Mario dos Reis and one anonymous reviewer for very helpful comments during revision.
 %}%

### 2016-05-01 1_penguins_eNprior/1_fossilDating.r

@@ -17,7 +17,10 @@ df1 <- createSummaryTable(result1$names, height1, fossilTable1, seq)
 caption <- "Summary of results for 36 fossil penguins under Model 1. {\\em post} is the posterior probability that the phylogenetic age is within the palaeontological age range. {\\em BF} is the bayes factor in support of the palaeontological age. {\\em phylo age} is the phylogenetic estimate of the age, along with the upper and lower of the corresponding 95\\% HPD credible interval. {\\em error} is the difference in millions of years between the phylogenetic point estimate of the fossil's age and the mean of it's palaeontological age range. {\\em ESS} is the estimated effective sample size for the phylogenetic age estimate."
 label <- "fossilTable1"
 
-writeSummaryTable("1_summaryTable.tex",caption, label, df1)
+# if uncommenting line below then add line "\scriptsize" before \begin{tabular} 
+# in the resulting tex file
+#writeSummaryTable("../1_summaryTable.tex",caption, label, df1)
+
 
 ind <- 1:nrow(df1)
 ind <- ind[df1$post < 0.05]
@@ -29,6 +32,6 @@ createPrecisionVsKnownCharactersFigure("1_precisionVsKnownCharacters.pdf", df1, 
 
 # Need to split younger and older ones
 
-createHistPlotFigures("1_fossilDatingHist_younger", result1$names[df1$geo < 32], height1[df1$geo < 32], fossilTable1[df1$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
+createHistPlotFigures("../1_fossilDatingHist_younger", result1$names[df1$geo < 32], height1[df1$geo < 32], fossilTable1[df1$geo < 32,], combinedPlots=TRUE, xmin=0, xmax=45)
 
-createHistPlotFigures("1_fossilDatingHist_older", result1$names[df1$geo > 32], height1[df1$geo > 32], fossilTable1[df1$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)
+createHistPlotFigures("../1_fossilDatingHist_older", result1$names[df1$geo > 32], height1[df1$geo > 32], fossilTable1[df1$geo > 32,], combinedPlots=TRUE, xmin=10, xmax=70)

### 2016-05-01 fossilDating-revision.tex
proofing

@@
 
-This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$), nevertheless the individual age estimates are more precise (median HPD range = 4.2 My for canids as opposed to 9.6 My for penguins).
+This data set contains half as many morphological characters as does the penguin data set ($122$ versus $245$), nevertheless the individual age estimates are much more precise in absolute terms (median HPD range = 4.2 Myr for canids as opposed to 9.6 Myr for penguins).
 However this is mainly due to the fact that the average age of the penguin fossils is considerably larger and the median relative precision (i.e.
@@
 Figure \ref{fig:canidTree} shows a sample from the posterior distribution of the analysis of the canid data set.
-The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}) The x-axis is in units of Mya.
+The tree has three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).
 
@@
 %\gslater{%
-showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}).}
+showing three main clades, one clade with extant representatives and two extinct clades ({\it Hesperocyoninae} and {\it Borophaginae}). The x-axis is in units of Mya. }
 %}%

### 2016-05-01 fossilDating-revision.tex
final changes, ready to resubmit

@@
 Recent advances have allowed for both morphological fossil evidence and molecular sequences to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability.
-In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa.
+In particular the fossilized birth-death tree prior and the Lewis-Mk model of discrete morphological evolution allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa.
 We exploit this statistical framework to investigate the internal consistency of these models by producing phylogenetic estimates of the age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species (penguins and canids).
-We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age and a median relative error in the two data sets of 5.7\% and 13.2\% respectively.
-The median relative standard error (RSD) was 9.2\% and 7.2\% suggesting good precision, although with some outliers.
+We find that the estimation accuracy of fossil ages is generally high with credible intervals seldom excluding the true age and median relative error in the two data sets of 5.7\% and 13.2\% respectively.
+The median relative standard error (RSD) was 9.2\% and 7.2\% respectively, suggesting good precision, although with some outliers.
 In fact in the two data sets we analyze the phylogenetic estimates of fossil age is on average  $< 2$ My from the midpoint age of the geological strata from which it was excavated. 
@@
 For the penguin analyses, the parameters of the fossilized birth-death model tree prior  were specified as described in the section ``Computing the phylogenetic evidence for an age range''. 
-For the canid analysis we used the standard parametrisation $\eta=(T,d,r,s)$, with the following priors: uniform prior from in the interval $(0, 120)$ million years for origin $T$, LogNormal($M=-3.5, S=1.5$) prior for diversification rate $d$, unit uniform prior $(0,1)$ for turnover $r$ and sampling proportion $s$.
+Since we did not perform Bayes factor (BF) analyses for the canid dataset, we used the standard parametrisation $\eta=(T,d,r,s)$, with the following priors: uniform prior from in the interval $(0, 120)$ million years for origin $T$, LogNormal($M=-3.5, S=1.5$) prior for diversification rate $d$, unit uniform prior $(0,1)$ for turnover $r$ and sampling proportion $s$.
 
@@
 The median error (where the {\it error} is the difference between the phylogenetic median and the geological median) is 1.96 Myr.
-The median relative error (where the {\it relative error} is the {\it error} divided by the geological median) was 5.7\% and the median relative standard deviation (RSD; defined as the standard deviation of the marginal posterior divided by the posterior estimate) was of 9.2\%.
+The median relative error (where the {\it relative error} is the {\it error} divided by the geological median) was 5.7\% and the median relative standard deviation (RSD; defined as the standard deviation of the marginal posterior divided by the posterior median estimate) was 9.2\%.
 A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

### 2016-05-03 fossilDating-revision.tex
added CCE affiliation

@@
 
-\noindent {\normalsize \sc  Alexei J.\ Drummond$^{1,2}$ and Tanja Stadler$^{2,3}$}\\
+\noindent {\normalsize \sc  Alexei J.\ Drummond$^{1,2,3}$ and Tanja Stadler$^{3,4}$}\\
 \noindent {\small \it 
-$^1$Department of Computer Science, University of Auckland, Auckland, 1010, New Zealand;\\
-$^2$Department of Biosystems Science \& Engineering, Eidgen\"{o}ssische Technische Hochschule Z\"{u}rich, 4058 Basel, Switzerland;\\
-$^3$Swiss Institute of Bioinformatics (SIB), Switzerland.}
+$^1$Centre for Computational Evolution, University of Auckland, Auckland, New Zealand;\\
+$^2$Department of Computer Science, University of Auckland, Auckland, 1010, New Zealand;\\
+$^3$Department of Biosystems Science \& Engineering, Eidgen\"{o}ssische Technische Hochschule Z\"{u}rich, 4058 Basel, Switzerland;\\
+$^4$Swiss Institute of Bioinformatics (SIB), Switzerland.}
 \end{center}
 \medskip
-\noindent{\bf Corresponding author:} Alexei J. Drummond, Department of Computer Science, University of 
-Auckland, 38 Princes Street, Auckland, 1010, New Zealand; E-mail: alexei@cs.auckland.ac.nz\\
+\noindent{\bf Corresponding author:} Alexei J. Drummond, Centre for Computational Evolution, University of 
+Auckland, Auckland, New Zealand; E-mail: alexei@cs.auckland.ac.nz\\
 

### 2016-05-03 canids/stratRanges/canids-stratRanges.r
moving figure files to top directory for ArXiv submission

@@ -1,7 +1,7 @@
 require(coda)
 source("~/Git/fossilDating/fossilDatingFunctions.r")
 
-setwd("~/Git/fossilDating/canids/stratRanges")
+setwd("~/Git/fossilDating")
 fossilTable_c <- read.table("~/Git/fossilDating/canids/stratRanges/canidFossilTable_stratRanges.txt", sep="\t", header=T)
 regexp = "\\-[0-9]+\\.log"
 seq <- read.table("~/Git/fossilDating/canids/stratRanges/canids-sorted-sequences.txt", header=T, sep="\t")

### 2016-05-03 fossilDating-revision.tex
update figure paths
removed one word from abstract to fit in ArXiv abstract limit :)

@@
 The high level of internal consistency found in our analyses suggests that the Bayesian statistical model employed is an adequate fit for both the geological and morphological data, and provides evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa.
-We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, further testing of the ``morphological clock'',
+We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing of the ``morphological clock'', 
 and for uncovering potential model misspecification and/or data errors when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
@@
 \begin{figure}
-\includegraphics{canids/stratRanges/1c_phyloAgeVsGeoAge.pdf}
+\includegraphics{1c_phyloAgeVsGeoAge.pdf}
 \caption{\label{fig:canidMorphVsGeo}
@@
 \begin{figure}
-\includegraphics[width=15cm]{canids/stratRanges/1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
+\includegraphics[width=15cm]{1_canids_ranges_3_subset-1460502871266-tree10001.pdf}
 \caption{\label{fig:canidTree} A sample from the posterior distribution of an analysis of the canid data set, 
